---
title: "AEME Inputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AEME Inputs}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(AEME)

path <- tempdir()
aeme_dir <- system.file("extdata/lake/", package = "AEME")
aeme <- yaml_to_aeme(path = aeme_dir, file = "aeme.yaml")

data("key_naming", package = "AEME")

met_vars <- key_naming |> 
  dplyr::filter(grepl("MET", var_aeme)) |> 
  dplyr::select(name_text, var_aeme, units) |> 
  dplyr::rename(Name = name_text, Variable = var_aeme, Units = units) |> 
  dplyr::mutate(
    Required = dplyr::case_when(
      Variable %in% c("MET_radswd", "MET_tmpair", "MET_wndspd", "MET_pprain") ~ "Yes",
      Variable %in% c("MET_ppsnow") ~ "Defaults to 0 if not provided",
      Variable %in% c("MET_prsttn") ~ "Can be derived from mean sea level pressure",
      Variable %in% c("MET_prmslp") ~ "Can be derived from station pressure",
      Variable %in% c("MET_tmpdew") ~ "Can be derived from air temperature and relative humidity",
      Variable %in% c("MET_humrel") ~ "Can be derived from air temperature and dew point temperature",
      Variable %in% c("MET_prvapr") ~ "Can be derived from air temperature and relative humidity",
      Variable %in% c("MET_wnduvu", "MET_wnduvv") ~ "Can be derived from wind speed and direction",
      Variable %in% c("MET_wndspd", "MET_wndir") ~ "Can be derived from u an v wind vectors",
      Variable %in% c("MET_cldcvr") ~ "Can be derived from date, air temperature, relative humiity, shortwave radiation, latitude, longitude and elevation",
      Variable %in% c("MET_radlwd") ~ "Can be derived from cloud cover, air temperature and relative humidity",
      .default = "Can be derived" 
    )
  ) |> 
  dplyr::arrange(dplyr::desc(Required))

```

## Summary

Inputs for AEME follow standardised conventions that are compatible with most
lake ecosystem modelling frameworks. All variable names are defined in the
`key_naming` data frame, which maps human-readable names to AEME variable codes
and their units.

## Lake

The `lake` list object provides physical and geographic metadata about the
waterbody. Required fields are:

| Field | Type | Description |
|-------|------|-------------|
| `name` | character | Name of the lake or reservoir |
| `id` | character / numeric | Unique identifier |
| `latitude` | numeric | Latitude (degrees North) |
| `longitude` | numeric | Longitude (degrees East) |
| `elevation` | numeric | Surface elevation above sea level (m) |
| `depth` | numeric | Maximum water depth (m) |
| `area` | numeric | Surface area at maximum depth (m²) |

The `depth` and `area` fields are used to generate a simple cone-shaped
hypsograph if none is provided in the `input` slot. For more accurate
simulations it is strongly recommended to supply a measured hypsograph.

## Meteorological Data

Meteorological data requirements for lake modelling are similar to those used in
most lake ecosystem models. The following meteorological variables are 
**required** for AEME: 

- Air temperature
- Shortwave radiation
- Wind speed
- Rain
- Either relative humidity or dew point temperature
- Either mean sea level pressure or station pressure

The following meteorological variables can be **derived** from the required
variables:

- Dew point temperature
- Relative humidity
- Precipitation vapor pressure
- Wind u and v components
- Cloud cover
- Longwave radiation

```{r echo=FALSE}
DT::datatable(met_vars, rownames = FALSE)
```


## Inflows and Outflows

### Inflows

Inflows supply water, heat, and (if the biogeochemistry module is active)
nutrients and sediment to the lake. Inflows are specified as a **named list**
of data frames passed to `add_inflows()`. The list name is used as the stream
identifier in each model.

Each inflow data frame must contain:

| Column | Units | Description |
|--------|-------|-------------|
| `Date` | `YYYY-MM-DD` | Date of the observation |
| `HYD_flow` | m³ day⁻¹ | Volumetric flow rate |
| `HYD_temp` | °C | Inflow water temperature (defaults to air temperature if absent) |
| `CHM_salt` | PSU | Salinity (0 for freshwater) |

Additional biogeochemical variables (e.g. `CHM_oxy`, `PHS_frp`, `NIT_amm`)
can be included if the ecosystem model is being used. All variable names must
match those in the `key_naming` data frame.

```{r inflow-example, eval=FALSE}
# Example: two inflows
inf_data <- list(
  stream_north = data.frame(
    Date     = seq(as.Date("2020-01-01"), as.Date("2021-12-31"), by = "day"),
    HYD_flow = 5000,
    HYD_temp = 12,
    CHM_salt = 0
  ),
  stream_south = data.frame(
    Date     = seq(as.Date("2020-01-01"), as.Date("2021-12-31"), by = "day"),
    HYD_flow = 3000,
    HYD_temp = 10,
    CHM_salt = 0
  )
)

aeme <- add_inflows(aeme, data = inf_data)
```

### Outflows

Outflows drain water from the lake or reservoir. Like inflows, they are
specified as a **named list** of data frames passed to `add_outflows()`.

Each outflow data frame must contain:

| Column | Units | Description |
|--------|-------|-------------|
| `Date` | `YYYY-MM-DD` | Date |
| `HYD_flow` | m³ day⁻¹ | Volumetric flow rate |

#### Outlet elevation

The `elevation` argument of `add_outflows()` specifies the depth at which each
outlet withdraws water:

- Set `elevation = -1` (or `elevation = list(outflow = -1)` for a named list)
  for a **surface outlet** – the outlet always withdraws from the current water
  surface.
- Set `elevation` to the actual elevation **above sea level (m)** for a
  **fixed-depth outlet** (e.g. a dam penstock or selective withdrawal). The
  value must fall within the elevation range of the hypsograph.

```{r outflow-single, eval=FALSE}
# Single surface outlet
outf_single <- list(
  outlet = data.frame(
    Date     = seq(as.Date("2020-01-01"), as.Date("2021-12-31"), by = "day"),
    HYD_flow = 8000
  )
)

aeme <- add_outflows(
  aeme,
  data      = outf_single,
  elevation = list(outlet = -1)  # surface outlet
)
```

#### Multiple outlets at different levels

Reservoirs commonly have several outlets at different depths. Provide one entry
per outlet in both `data` and `elevation`:

```{r outflow-multiple, eval=FALSE}
# Reservoir with a surface spillway and a deep penstock
outf_multi <- list(
  spillway = data.frame(
    Date     = seq(as.Date("2020-01-01"), as.Date("2021-12-31"), by = "day"),
    HYD_flow = 2000  # only active during high water events
  ),
  penstock = data.frame(
    Date     = seq(as.Date("2020-01-01"), as.Date("2021-12-31"), by = "day"),
    HYD_flow = 8000  # daily regulated release
  )
)

# Reservoir with full supply level at 150 m a.s.l. and max depth 25 m
aeme <- add_outflows(
  aeme,
  data = outf_multi,
  elevation = list(
    spillway = -1,   # surface overflow
    penstock = 130   # 20 m below the full supply level
  )
)
```

For a worked example of a reservoir with two outlets, see
`vignette("reservoir-aeme")`.

## Parameters

Being able to specify the parameters of the model is crucial for the
simulation. Parameters can be stored in a data frame within the `aeme` object.
The parameters in this data frame are used to update the model configuration
files and/or update the meteorological data (e.g. scaling wind factor) and/or
update the inflow/outflow data (e.g. scaling inflow factor). 

```{r echo=TRUE, eval=FALSE}
utils::data("aeme_parameters")
aeme_parameters
```

```{r echo=FALSE}
DT::datatable(aeme_parameters, rownames = FALSE)
```

```{r}
parameters(aeme) <- aeme_parameters
```



