---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
unlink("45819_wainamu/", recursive = TRUE, force = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Setting up the Aquatic Ecosystem Model Ensemble is straight forward and can be
done in a few steps. This vignette will guide you through the process of setting
up the model for a lake in New Zealand.

```{r setup}
library(AEME)
```


# Lake data

The first step is to define the lake data. This includes the location of the
lake, the depth and area of the lake, and the elevation of the lake.


```{r lake-location}
# Define the location of the lake
lat <- -36.8898
lon <- 174.46898
```

```{r view-in leaflet}
# View the location of the lake in a map
library(leaflet)
leaflet() |> 
  addTiles() |> 
  addMarkers(lng = lon, lat = lat, popup = "Wainamu Lake")

```

## Depth and area data

The depth and area of the lake are required for the model. These can be
obtained from a variety of sources, including regional councils, the New Zealand
Hydrological Database, or from the lake owner.


```{r}
# Set depth & area
depth <- 13.07 # Depth of the lake in metres
area <- 152343 # Area of the lake in m2
```


## Elevation data

Elevation data can be acquired for New Zealand from the digital elevation model
hosted on the LINZ Data Service. There is a wrapper function for this in the
`aemetools` package. This requires an API key from LINZ.

You can easily create on the LINZ website: https://data.linz.govt.nz/ or use the
function within the `aemetools` package to create one.

```{r get-linz-api-key, eval = FALSE}
aemetools::create_linz_key()
```

Then adding it to your `.Renviron` file.

```{r add-linz-api-key, eval = FALSE}
# Add the LINZ API key to your .Renviron file
aemetools::add_linz_key(key = "your_key_here")
```

The `get_dem_value` function will return the elevation of the lake in metres
above sea level.

```{r get-elevation}
# Get the elevation of the lake
elevation <- aemetools::get_dem_value(lat = lat, lon = lon)
elevation # in metres above sea level

```

We will now create a list of the lake data. This will be used to construct the
AEME object.

```{r create-lake-list}
# Define lake list
lake = list(
    name = "Wainamu",
    id = 45819,
    latitude = lat,
    longitude = lon,
    elevation = elevation,
    depth = depth,
    area = area
  )
```

# Time data

The time data is required for the model. This includes the start and stop times
for the model run. 

```{r time-data}
# Define start and stop times
start <- "2020-08-01 00:00:00"
stop <- "2021-06-30 00:00:00"

time <- list(
    start = start,
    stop = stop
  )
```



# Input data

## Meteorological data

### Download ERA5 data

We will use the `aemetools` package to download the ERA5 meteorological data for
the location of our lake. This only works for locations in New Zealand.

```{r get-era5-met}
# Get ERA5 meteorological data
met <- aemetools::get_era5_point(lat = lat, lon = lon, years = 2020:2021,
                                 parallel = TRUE)

```

View the summary of the meteorological data. The units have been converted to 
more common units used in aquatic ecosystem modelling.

```{r met-summary}
# Summary of meteorological data
summary(met)
```

The depth of this lake is 13.07 m, the area is 152343 m2, and the light 
extinction coefficient (Kw) is 1.31 m-1.


```{r}
# Set Kw
Kw <- 1.31 # Light extinction coefficient in m-1
```


## Hypsograph data

If you have hypsograph data for the lake, you can use it as input for the model.
This is a critical input for the model, as it defines the relationship between
the lake area and the lake elevation.

However, if you do not have hypsograph data, the model will use a simple
cone-shaped hypsograph based on the lake depth and area.

```{r generate-hypsorgaph}
# Generate a simple hypsograph
hypsograph <- data.frame(area = c(0, 152343), 
                         elev = c(elevation - depth, elevation))
hypsograph
```

```{r plot-hypsograph}
# Plot the hypsograph
plot(hypsograph$area, hypsograph$elev, type = "l", xlab = "Area (m2)", 
     ylab = "Elevation (m)")
```

As you can see, the hypsograph is a simple cone shape. Ideally, you would have
more detailed hypsograph data for your lake, but this will work for this 
example.

```{r make-aeme-list}
# Define input list
input = list(
    init_depth = depth,
    hypsograph = hypsograph,
    meteo = met,
    use_lw = TRUE,
    Kw = Kw
  )
```

# Construct the AEME object

The `aeme_constructor` function will take the input data and construct the AEME
object. The minimum inputs are the `lake`, `time`, and `input` data.


```{r construct-aeme-object}
# Construct AEME object
aeme <- aeme_constructor(lake = lake, 
                         time = time,
                         input = input)

```

## View AEME object

The AEME object can be inspected by printing it to the console. This will show
the inputs that have been use to construct the object along with default values
for inputs not provided.

```{r view-aeme-object}
aeme
```

In the configuration section of the output, under "Physical" and 
"Biogeochemical" for each model are labelle "Absent". This is because the 
model configurations have not been built. This is done in the next step.

# Building the AEME ensemble

## Model controls

The model controls are the settings for the AEME ensemble. These are
read in from a CSV file. The default CSV file is stored within the package
and can be accessed using the `get_model_controls` function. It has the 
argument  `use_bgc` which is a logical value to indicate whether to simulate the
default biogeochemical variables with the hydrodynamic variables or just the
hydrodynamic variables.

The model controls has the following columns:

- `var_aeme`: The variable name in the AEME object
- `simulate`: Whether to simulate the variable
- `inf_default`: The default inflow value
- `initial_wc`: The initial water column value
- `initial_sed`: The initial sediment value

```{r get-model-controls}
# Get model controls
model_controls <- get_model_controls()
model_controls
```

## Build the ensemble

The `build_ensemble` function will take the AEME object and the model controls
and build the ensemble. The `model` argument is a character vector of the models
to include in the ensemble. The models available are `dy_cd`, `glm_aed`, and
`gotm_wet`.

```{r select-models}
# Select models
model <- c("dy_cd", "glm_aed", "gotm_wet")
```

```{r build-ensemble}
# Build ensemble
aeme <- build_ensemble(aeme = aeme, model = model,  
                       model_controls = model_controls, use_bgc = TRUE)

print(aeme)
```

By default, the `build_ensemble` function will build the
file configuration for each model. This will create the necessary files for
each model to run. The files are also stored in the `aeme` object in the
`configuration` slot with a list for hydrodynamic and ecosystem model 
configurations. 

```{r view-files}
# View the files
cfg <- configuration(aeme)
names(cfg[["glm_aed"]])
```

All of the information and data needed to run an ensemble of models is now
contained within the `aeme` object. This allows for easy storage of all the data
and also for easy sharing of the data with others. Sharing the `aeme` object
with others allows them to run the ensemble of models without needing to
reconstruct the object. 

```{r run-ensemble}
# Run the ensemble
aeme <- run_aeme(aeme = aeme, model = model, model_controls = model_controls)
```

## View the output

The output from the model run is stored in the `output` slot of the `aeme` object.
This is a list with a list for each model. The list contains the output data
from the model run. 

```{r view-output}
# View the output
plot_output(aeme = aeme, model = model)
```


## Saving the AEME object

Saving the `aeme` object to a file can be done using the `saveRDS` function. 
This will save the object to a file with the `.rds`.

```{r save-aeme}
# Save the AEME object
saveRDS(aeme, "aeme.rds")
```
