---
title: "Guide to AEME"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guide to AEME}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(AEME)
```

# Introduction

We will load in an `Aeme` object and demonstrate how to access data and input 
variables into the object.

We will use lake Wainamu as an example.

```{r load-aeme}
rds_file <- system.file("extdata/lernzmp/LID45819.rds", package = "AEME")
aeme <- readRDS(rds_file)
class(aeme)

```

The `Aeme` object is an S4 R object that contains all the data and input variables
required to run the AEME model.

We can view all the slots within this object using the `slotNames` function.

```{r slot-names}
slotNames(aeme)
```

## Access data

To access the data from each of the 'slots' within the `Aeme` object, we can use
functions of the same name as the slot to retrieve the data.

For example, to access the data from the `lake` slot, we can use the `lake()` function.

```{r lake}

lke <- lake(aeme)
lke

```

Each of the slots is a list containing the data required to run the AEME model.

```{r slot-class}
class(lke)
```

## Update data

If you wish to update the data within the `Aeme` object, you must:

1. Extract the data from the slot using the associated function
2. Update the data
3. Assign the updated data back to the slot using the associated function

For example, to update the lake name, you can do the following:

```{r update-lake}
lke <- lake(aeme) # Extract the data
lke$name <- "TEST NAME" # Update the data
lake(aeme) <- lke # Assign the updated data back to the slot
aeme
```

The `Aeme` object now contains the updated lake name.

## Update the input data

To update the input data, you can follow the same process as updating the data.

## Lake
