---
title: "Installing Model Binaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing Model Binaries}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(AEME)
```

# Introduction

Since AEME v0.4.0, the package does not bundle model executables with the R package -- GLM-AED,
GOTM-WET, DYRESM-CAEDYM and Simstrat-AED2 are downloaded separately, once,
after installing AEME itself. This vignette covers:

1. Which models are available on which platform
2. Installing everything with one call, or one model at a time
3. Checking what's installed, and picking a specific version
4. Where AEME looks for an installed model, if you need to know

If you already have models installed (or are working through
`vignette("intro-aeme")` in an environment where this has been done for you),
skip ahead to `vignette("intro-aeme")`.

# Platform availability

Not every model builds on every operating system:

| Model | Windows | macOS | Linux |
|-------|---------|-------|-------|
| DYRESM-CAEDYM | Yes | No | No |
| GLM-AED | Yes | Yes | Yes |
| GOTM-WET | Yes | No | No |
| Simstrat-AED2 | Yes | Yes | Yes |

`install_models()` installs whatever is available for the platform it's run
on; asking for a model that doesn't build there raises an error naming what
is available instead of failing silently.

# Installing everything

The simplest path -- one call installs the latest release of every model
available for your platform:

```{r install-all, eval=FALSE}
library(AEME)
install_models()
```

This is safe to re-run: by default, a model already installed at the
resolved "latest" version is left untouched (`force = FALSE`).

# Installing one model at a time

Each model also has its own installer, for when you only need one:

```{r install-one, eval=FALSE}
install_glm_aed()
install_gotm_wet()
install_dy_cd()
install_simstrat_aed2()
```

`install_models()` accepts a `model` argument if you'd rather filter the
all-in-one call instead of calling installers individually:

```{r install-subset, eval=FALSE}
install_models(model = c("glm_aed", "simstrat_aed2"))
```

# Picking a specific version

By default, every installer resolves to the latest available release. To see
what's available -- for example, to pin a version for reproducibility, or to
match a model version a collaborator is using -- list versions before
installing:

```{r list-versions, eval=FALSE}
list_glm_versions()
list_gotm_wet_versions()
list_dy_cd_versions()
list_simstrat_aed2_versions()
list_simstrat_aed_versions()
```
```{r list-versions-glm, eval=TRUE}
list_glm_versions()
```


```{r install-pinned, eval=FALSE}
install_glm_aed(version = "4.0.0")
```

# Checking what's installed

```{r model-version, eval=FALSE}
get_model_version(model = "glm_aed")
```

`list_models()` returns every model code AEME knows about, independent of
what's actually installed -- useful as the input to functions like
`install_models(model = ...)` that expect the short code (`"glm_aed"`) rather
than the display name (`"GLM-AED"`):

```{r list-models}
library(AEME)
list_models()
```

# Where installed models live

Installers place binaries under a per-user AEME data directory (not inside
the R package library, so a package reinstall doesn't wipe them). You
shouldn't normally need to know the exact path -- `run_aeme()` and the
model-specific `run_*()` wrappers find it automatically -- but if a build
step reports a missing executable, `get_model_version()` and the
`*_exe_path()` family (`glm_exe_path()`, `gotm_wet_exe_path()`,
`dy_cd_exe_path()`, `simstrat_aed2_exe_path()`) resolve exactly which binary,
if any, AEME will use.

```{r check-exe-path}
exe_path <- glm_exe_path()
exe_path
```

# Which model should I pick?

Well, it depends on what you want to do. The model-specific vignettes cover the 
differences in detail, but in brief:

- GLM-AED is the most widely used and supported, and is the default in AEME. It 
is suitable for a wide range of lakes and reservoirs, and has a large user community.

- Simstrat-AED2/AED is a model that is also suitable for a wide range of lakes, 
it is particularly well-suited for deep, stratified lakes. It is less widely 
used than GLM-AED, but has a strong user community and is actively maintained. 
The Simstrat-AED model is an experimental version of the model, coupled with 
the recently (Augusu 2026) released AED API. It has only been developed within 
the Limnotrack branch, but may be of interest to researchers looking to explore 
new features and capabilities. 

- GOTM-WET is a more specialized model, suitable for certain types of lakes and 
reservoirs. It is less widely used than GLM-AED, but may be more appropriate for
certain applications.

- DYRESM-CAEDYM is a one-dimensional model that is suitable for certain types of
lakes and reservoirs. However, it is quite old and is not actively maintained, 
so it is not recommended for new projects.

# Next steps

With a model installed, `vignette("intro-aeme")` walks through the `Aeme`
object model and builds your first ensemble member.
