---
title: "Introduction to AEME"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to AEME}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(AEME)
## basic example code
tmpdir <- tempdir()
aeme_file <- system.file("extdata/lernzmp/LID45819.rds", package = "AEME")
aeme <- readRDS(aeme_file)
aeme <- remove_output(aeme)
# slotNames(aeme)

lke <- lake(aeme)
tme <- time(aeme)
cfg <- configuration(aeme)
obs <- observations(aeme)
inp <- input(aeme)
inf <- inflows(aeme)
outf <- outflows(aeme)
wbal <- water_balance(aeme)
outp <- output(aeme)
pars <- parameters(aeme)

# convert list into a tree structure
lake_names <- paste0(names(lke), collapse = ", ")
time_names <- paste0(names(tme), collapse = ", ")
config_names <- paste0(names(cfg), collapse = ", ")
obs_names <- paste0(names(obs), collapse = ", ")
input_names <- paste0(names(inp), collapse = ", ")
inflow_names <- paste0(names(inf), collapse = ", ")
outflow_names <- paste0(names(outf), collapse = ", ")
water_balance_names <- paste0(names(wbal), collapse = ", ")
output_names <- paste0(names(outp), collapse = ", ")
parameter_names <- paste0(names(pars), collapse = ", ")

```

## Summary

The AEME package hosts three one-dimensional hydrodynamic models: the DYnamic
REservoir Simulation Model (DYRESM), the General Lake Model (GLM), and the
General Ocean Turbulence Model (GOTM, which has been adapted for closed basins
for application to lakes and reservoirs). The models can be coupled to their
corresponding water quality models, the DYRESM-CAEDYM (Computational Aquatic
Ecosystem Dynamics Model), GLM-AED (Aquatic Ecosystem Dynamics Model), and
GOTM-WET (Water Ecosystem Tool).

Key aspects of the AEME package include:

-   Defined S4 class for `aeme` objects

-   Configuration of models from common and standardised inputs

-   Standardised calibration, manipulation and visualisation

This vignette provides an overview of the AEME framework, the three models, and 
the ensemble approach. For detailed information on data inputs and the S4 
structure, see the [AEME Inputs](aeme-inputs.html) vignette. For a practical 
tutorial, see [Setting up AEME for a new lake](setup-new-lake.html).

## The Three Models {#three-models}

AEME integrates three well-established one-dimensional (1D) lake models, each 
with distinct computational approaches and strengths. Understanding the 
differences helps in selecting appropriate models and interpreting ensemble 
results.

### DYRESM-CAEDYM {#dyresm-caedym}

**DYnamic REservoir Simulation Model - Computational Aquatic Ecosystem DYnamics 
Model**

**History and Development**

DYRESM was developed in the 1980s at the University of Western Australia and 
Centre for Water Research. It pioneered the Lagrangian layer approach for lake 
modeling. CAEDYM was later developed to add comprehensive biogeochemical 
capabilities.

**Physics and Computational Approach**

- **Lagrangian layers**: Water parcels move vertically through the water column
- **Variable layer thickness**: Layers can split or merge based on stratification
- **Layer-following approach**: Particularly effective for tracking water masses 
  in reservoirs
- **Time-step**: Adaptive, typically 30-60 minutes

**Key Features**

- Excellent representation of selective withdrawals in reservoirs
- Detailed tracking of water quality through inflows and withdrawals
- Comprehensive biogeochemical modules (nutrients, phytoplankton, zooplankton, 
  sediments)
- Established in reservoir management applications

**Typical Applications**

- Drinking water reservoirs with complex withdrawal strategies
- Multi-level outlet management
- Water quality forecasting in stratified systems
- Long-term climate change impact studies

**References**

- Imberger, J., & Patterson, J. C. (1981). A dynamic reservoir simulation model 
  - DYRESM: 5. *Transport Models for Inland and Coastal Waters*, 310-361.
- Hamilton, D. P., & Schladow, S. G. (1997). Prediction of water quality in 
  lakes and reservoirs. Part I - Model description. *Ecological Modelling*, 
  96(1-3), 91-110.
- Burger, D. F., Hamilton, D. P., & Pilditch, C. A. (2008). Modelling the 
  relative importance of internal and external nutrient loads on water column 
  nutrient concentrations and phytoplankton biomass in a shallow polymictic 
  lake. *Ecological Modelling*, 211(3-4), 411-423.

### GLM-AED {#glm-aed}

**General Lake Model - Aquatic EcoDynamics model**

**History and Development**

GLM was developed in the 2010s through an international collaboration led by the 
University of Western Australia. It modernizes 1D lake modeling with a flexible 
framework compatible with contemporary ecosystem models. AED is the latest 
generation of water quality models, building on AED+ and earlier frameworks.

**Physics and Computational Approach**

- **Dynamic Eulerian layers**: Fixed spatial grid with variable layer thickness
- **Flexible vertical resolution**: Automatically adjusts layers based on 
  stratification
- **Efficient computations**: Optimized for ensemble applications and 
  uncertainty analysis
- **Time-step**: Hourly to daily, depending on application

**Key Features**

- Modular biogeochemical framework (AED2)
- Easy configuration through namelists (NML files)
- Extensive validation across diverse lake types
- Active development community and regular updates
- Integration with lake ensemble frameworks

**Typical Applications**

- Natural lakes of varying trophic status
- Multi-lake regional studies
- Climate change scenarios
- Coupled catchment-lake modeling
- Real-time forecasting systems

**References**

- Hipsey, M. R., Bruce, L. C., Boon, C., Busch, B., Carey, C. C., Hamilton, D. 
  P., ... & Winslow, L. A. (2019). A General Lake Model (GLM 3.0) for linking 
  with high-frequency sensor data from the Global Lake Ecological Observatory 
  Network (GLEON). *Geoscientific Model Development*, 12(1), 473-523.
- Bruce, L. C., Frassl, M. A., Arhonditsis, G. B., Gal, G., Hamilton, D. P., 
  Hanson, P. C., ... & Trolle, D. (2018). A multi-lake comparative analysis of 
  the General Lake Model (GLM): Stress-testing across a global observatory 
  network. *Environmental Modelling & Software*, 102, 274-291.
- Hipsey, M. R., Bruce, L. C., & Hamilton, D. P. (2014). *GLM - General Lake 
  Model: Model overview and user information*. AED Report #26, The University 
  of Western Australia, Perth, Australia.

### GOTM-WET {#gotm-wet}

**General Ocean Turbulence Model - Water Ecosystem Tool**

**History and Development**

GOTM was originally developed for ocean modeling in the 1990s and later adapted 
for lakes and estuaries. WET (Water Ecosystem Tool) is a comprehensive 
biogeochemical model developed by WaterITech specifically for integration with 
GOTM and application to lakes and reservoirs.

**Physics and Computational Approach**

- **Fixed Eulerian grid**: Regular vertical spacing throughout simulation
- **Advanced turbulence closure**: Multiple turbulence schemes (k-ε, k-ω, etc.)
- **High vertical resolution**: Typically 0.5-1 m spacing
- **YAML configuration**: Modern configuration approach
- **Time-step**: Sub-hourly for stability

**Key Features**

- Sophisticated turbulence modeling
- High-resolution vertical mixing representation
- Framework for Aquatic Biogeochemical Models (FABM) integration
- Detailed surface and bottom boundary layers
- Coupling with atmospheric models

**Typical Applications**

- Deep stratified lakes
- Systems with complex mixing dynamics
- Research applications requiring detailed turbulence
- Lakes with significant benthic-pelagic coupling
- Process studies and model development

**References**

- Burchard, H., Bolding, K., & Villarreal, M. R. (1999). *GOTM, a general ocean 
  turbulence model. Theory, implementation and test cases*. Tech. Rep. 
  EUR-18745-EN, European Commission.
- Bruggeman, J., & Bolding, K. (2014). A general framework for aquatic 
  biogeochemical models. *Environmental Modelling & Software*, 61, 249-265.
- Bolding, K., Bruggeman, J., Brüchert, V., Grimm, H., Holtermann, P., Hu, T., 
  ... & Umlauf, L. (2020). *GETM and GOTM - a general estuarine and lake model. 
  Status and perspectives.* Report, 2020.

### Model Comparison {#model-comparison}

The following table summarizes key differences between the three models:

| Feature | DYRESM-CAEDYM | GLM-AED | GOTM-WET |
|---------|---------------|---------|----------|
| **Approach** | Lagrangian layers | Lagrangian layers | Fixed Eulerian |
| **Layer structure** | Variable (moves with water) | Variable thickness | Fixed grid |
| **Vertical resolution** | 5-50 layers | 2-500 layers (user defined) | 2-500 layers (user defined) |
| **Timestep** | Fixed (1 hr typical) | Fixed (1 hr typical) | Fixed (1 hr typical) |
| **Turbulence** | Mixed-layer model | k-ε or Henderson-Sellers | Multiple schemes (k-ε typical) |
| **Configuration** | Text files | Namelists (NML) | YAML files |
| **Best for** | Reservoirs, withdrawals | Natural lakes, ensembles | Deep lakes, research |
| **Computational speed** | Slow | Fast | Fast |
| **BGC complexity** | High (CAEDYM) | High (AED2) | High (FABM/WET) |

**When to use each model:**

- **DYRESM-CAEDYM**: Reservoirs with selective withdrawals, water quality 
  management, operational forecasting
- **GLM-AED**: Natural lakes, multi-lake studies, ensemble uncertainty analysis, 
  real-time applications
- **GOTM-WET**: Stratified lakes, detailed turbulence studies, research 
  applications, process investigations

## The Ensemble Approach {#ensemble-approach}

AEME's primary innovation is the standardization of inputs and outputs across 
three fundamentally different lake models, enabling true ensemble modeling of 
aquatic ecosystems.

### Why Use Multiple Models?

**1. Structural Uncertainty**

Each model makes different assumptions about physics and biogeochemistry. No 
single model is universally "best" - performance varies by:
- Lake type (reservoir vs. natural)
- Stratification dynamics
- Available calibration data
- Spatial and temporal scales
- Management questions

Using multiple models quantifies structural uncertainty - differences arising 
from model formulation rather than parameter uncertainty.

**2. Ensemble Predictions**

Ensemble means and medians often outperform individual models by:
- Averaging out model-specific biases
- Providing probabilistic forecasts
- Identifying robust predictions (model agreement)
- Highlighting uncertain predictions (model divergence)

**3. Model Complementarity**

Different models excel at different aspects:
- DYRESM-CAEDYM: tracking water parcels in reservoirs
- GLM-AED: computational efficiency for ensembles
- GOTM-WET: detailed vertical mixing processes

### How AEME Standardizes Inputs and Outputs

**Input Standardization**

AEME translates common inputs into model-specific formats:

1. **Meteorology**: Same data → model-specific forcing files
2. **Hypsography**: Single hypsograph → adjusted for each model's grid
3. **Inflows/Outflows**: Unified format → model-specific boundary conditions
4. **Parameters**: Common parameter table → model configuration files

This ensures models simulate the *same lake* with *comparable forcings*.

**Output Standardization**

All models output to a common format:
- Variable names mapped to AEME conventions (`key_naming`)
- Consistent spatial interpolation to comparable depths
- Synchronized temporal resolution
- Standard quality control and gap filling

### Ensemble Evaluation

**Model Agreement**

When models agree, confidence is high. When models diverge, structural 
uncertainty is significant.

```{r ensemble-agreement, eval=FALSE}
# Compare model predictions
plot_output(aeme = aeme, var_sim = "HYD_temp", model = c("dy_cd", "glm_aed", "gotm_wet"))

```

**Performance Metrics**

Use `assess_model()` to compare model performance against observations:

```{r model-performance, eval=FALSE}
performance <- assess_model(aeme = aeme)
print(performance)
# Shows RMSE, NSE, R² for each model
```

**Weighted Ensembles**

Models can be weighted by past performance:

```{r weighted-ensemble, eval=FALSE}
# Weight by NSE scores
weights <- performance$NSE / sum(performance$NSE)
weighted_mean <- calculate_weighted_ensemble(aeme, weights = weights)
```

### Interpretation Guidelines

1. **When models agree**: High confidence in predictions
2. **When models disagree systematically**: Check inputs and observations
3. **When one model consistently outperforms**: Consider lake-specific physics
4. **When all models fail**: Likely data quality or missing processes

## Model Applications and Case Studies {#applications}

AEME has been applied to diverse aquatic systems across New Zealand and 
internationally.

### Example Applications

**1. Drinking Water Reservoirs**

See `vignette("reservoir-aeme")` for a detailed example of applying AEME to a 
reservoir with multiple outlets and water quality management.

**2. Natural Lakes**

The LERNZmp (Lake Ecosystem Research New Zealand Modelling Platform) uses AEME 
for 100+ New Zealand lakes. See `vignette("lernzmp-aeme")` for details.

**3. Water Balance Studies**

AEME can estimate unknown inflows/outflows from water level observations. See 
`vignette("rotoehu-water-balance")` for a worked example.

**4. Single Model Applications**

While AEME is designed for ensembles, individual models can be run for specific 
applications. See `vignette("glm-aed")` for GLM-AED specific examples.

### Key References for AEME

- Moore, T. N., Mesman, J. P., Ladwig, R., Feldbauer, J., Olsson, F., Pilla, R. 
  M., ... & Hipsey, M. R. (2021). LakeEnsemblR: An R package that facilitates 
  ensemble modelling of lakes. *Environmental Modelling & Software*, 143, 105101.
  
Future AEME-specific publications will be listed here as they become available.

## AEME object {#sec-aeme-object}

### Description {#sec-description}

The `aeme` object is the main object in the AEME package. It is an S4 class that
contains all the information required to run a model. The `aeme` object contains
the following slots:

-   [**lake**](#sec-lake) - Lake metadata (location, dimensions)
    (`r lake_names`)
-   [**time**](#sec-time) - Simulation period and temporal settings
    (`r time_names`)
-   [**configuration**](#sec-configuration) - Model configurations and controls
    (`r config_names`)
-   [**observations**](#sec-observations) - Observational data for validation
    (`r obs_names`)
-   [**inputs**](#sec-inputs) - Core physical inputs (meteorology, hypsograph)
    (`r input_names`)
-   [**inflows**](#sec-inflows) - Stream inflow data
    (`r inflow_names`)
-   [**outflows**](#sec-outflows) - Outlet data and configurations
    (`r outflow_names`)
-   [**water_balance**](#sec-water_balance) - Water balance settings and 
    calculations (`r water_balance_names`)
-   [**parameters**](#sec-parameters) - Model parameter values for calibration
    (`r parameter_names`)
-   [**output**](#sec-output) - Model output (populated after simulation)
    (`r output_names`)

For comprehensive details on each slot including structure, required fields, and 
usage examples, see the [AEME Inputs vignette](aeme-inputs.html).

#### Lake {#sec-lake}

The `lake` slot contains metadata about the waterbody including name, location, 
dimensions, and optional spatial information.

**Required fields**: name, id, latitude, longitude, elevation, depth, area

See [Lake Slot](aeme-inputs.html#lake-slot) in the AEME Inputs vignette for 
complete details.

#### Time {#sec-time}

The `time` slot defines the simulation period, temporal resolution, and spin-up 
periods for model initialization.

**Required fields**: start, stop

**Optional fields**: timestep, spin_up

See [Time Slot](aeme-inputs.html#time-slot) in the AEME Inputs vignette for 
complete details.

#### Configuration {#sec-configuration}

The `configuration` slot stores model-specific configuration files and the
`model_controls` data frame. This slot is populated by `build_aeme()`.

See [Configuration Slot](aeme-inputs.html#configuration-slot) in the AEME 
Inputs vignette for complete details.

#### Observations {#sec-observations}

The `observations` slot contains observational data for model validation, 
calibration, and initialization. It includes in-lake profiles (`lake`) and water 
level time series (`level`).

See [Observations Slot](aeme-inputs.html#observations-slot) in the AEME Inputs 
vignette for complete details.

#### Input {#sec-inputs}

The `input` slot contains core physical inputs required by all models: 
meteorological forcing, lake bathymetry (hypsograph), light extinction, and 
initial conditions.

**Required fields**: hypsograph, meteo, Kw

**Optional fields**: init_profile, init_depth, use_lw

See [Input Slot](aeme-inputs.html#input-slot) in the AEME Inputs vignette for 
complete details.

#### Inflows {#sec-inflows}

The `inflows` slot contains stream inflow data as named lists of data frames, 
plus optional scaling factors for each model.

See [Inflows Slot](aeme-inputs.html#inflows-slot) in the AEME Inputs vignette 
for complete details.

#### Outflows {#sec-outflows}

The `outflows` slot contains outlet data, outlet elevations, and optional 
scaling factors.

See [Outflows Slot](aeme-inputs.html#outflows-slot) in the AEME Inputs vignette 
for complete details.

#### Water balance {#sec-water_balance}

The `water_balance` slot is generated internally by `build_aeme()` when inflows 
or outflows need to be estimated from water level changes.

**Fields**: method (1, 2, or 3), use ("obs" or "mod"), data

See [Water Balance Slot](aeme-inputs.html#water-balance-slot) in the AEME 
Inputs vignette for complete details.

#### Parameters {#sec-parameters}

The `parameters` slot contains a data frame of model parameter values for 
calibration, sensitivity analysis, and model configuration.

See [Parameters Slot](aeme-inputs.html#parameters-slot) in the AEME Inputs 
vignette for complete details.

#### Output {#sec-output}

The `output` slot stores model results after running `run_aeme()`. Initially
empty, it is populated with time-series data for each model and variable.

See [Output Slot](aeme-inputs.html#output-slot) in the AEME Inputs vignette for 
complete details.

### Model controls {#sec-model_controls}

The `model_controls` is a data.frame generated by the `get_model_controls()`. 
The data.frame has the columns:

-   var_aeme - character; the AEME variable name

-   simulate - logical; add the variable to the `Aeme` object

-   inf_default - numeric; default value to use in the inflows if none present 
    in the inflows. This is particularly important for configuring water 
    chemistry for the inflows if `use_bgc = TRUE`.
    
-   initial_wc - numeric; value to use in initialising the model. This will be 
    automatically updated if the variable is present in the `observations` slot.
    
-   initial_sed - numeric; value to use in initialising the sediment module for 
    the DYRESM-CAEDYM model.
    
-   conversion_aed - numeric; factor to multiply by to convert to GLM-AED units.

When the model is built, the `model_controls` data.frame is stored in the
`configuration` slot of the `aeme` object. It can be retrieved with
`get_model_controls(aeme = aeme)`.


### Creation

The `aeme` object can be created using the `aeme_constructor()` function. It
requires at minimum the `lake`, `time`, and `input` list objects. The object can
also be created from a YAML file using the `yaml_to_aeme()` function. The YAML
file contains all the information required to run the model.

```{r warning=FALSE, message=FALSE, eval=TRUE}
# Define lake list
lat <- -36.88921
lon <- 174.4669
depth <- 13.08
area <- 153648

lake <- list(
    latitude = lat,
    longitude = lon,
    name = "lake",
    id = "123",
    depth = depth,
    area = area
  )
time <- list(
  start = as.POSIXct("2020-07-01"),
  stop = as.POSIXct("2022-06-30")
)

hypsograph <- generate_hypsograph(max_depth = depth, surface_area = area,
                                  volume_development = 1.2)

met <- aemetools::get_era5_land_point_nz(lat = lat, lon = lon,
                                         years = 2020:2022)

#' Define input list
input = list(
  hypsograph = hypsograph,
  meteo = met,
  Kw = 1.21
)

aeme <- aeme_constructor(lake = lake, time = time, input = input)

```
```{r}
slotNames(aeme)
```


## Manipulation

The `aeme` object can be manipulated using the `AEME` package functions. The
functions are defined by the slot names of the `aeme` object. For example, the
`lake` slot can be manipulated using the `lake` function.

```{r aeme-manipulation}
# Load lake data
lke <- lake(aeme)
# Print lake data to console
print(lke)

# Change lake name
lke[["name"]] <- "AEME"

# reassign lake data to aeme object
lake(aeme) <- lke

aeme

```

### Visualisation

The `aeme` object can be visualised simply using the `plot` function. The `plot`
function can be applied to the different slots of the `aeme` object. For
example, the `lake` slot can be visualised using the `plot` function.

```{r aeme-visualisation-lake, fig.width=7, fig.height=5}
plot(aeme, "lake")

```

```{r aeme-visualisation-input, fig.width=7, fig.height=5}
plot_met_tile(aeme, var_aeme = "MET_tmpair")

```

```{r plot-hypsograph, warning=FALSE, message=FALSE}
plot_hyps(aeme)
```

## Next Steps

Now that you understand the basics of AEME, you can explore more advanced topics
and practical applications in our other vignettes:

### Getting Started with AEME

- **[Set up AEME for a new lake](setup-new-lake.html)** - A practical tutorial 
  that walks you through setting up the model for a new lake, including how to 
  obtain and prepare input data.

- **[AEME Inputs](aeme-inputs.html)** - Comprehensive reference documentation on 
  the input requirements and S4 structure of the AEME object.

### Advanced Applications

- **[Using LERNZmp with AEME](../articles/lernzmp-aeme.html)** - Learn how to use 
  the Lake Ecosystem Research New Zealand Model Platform (LERNZmp) web interface 
  with AEME.

- **[Reservoir Simulation with Multiple Outlets](../articles/reservoir-aeme.html)** - 
  Demonstrates how to model reservoirs with regulated water levels and multiple 
  outlets at different depths.

- **[GLM-AED: The General Lake Model coupled with AED](../articles/glm-aed.html)** - 
  Detailed guide on using the GLM-AED model within the AEME framework, including 
  parameter libraries and configuration.

- **[Lake Rotoehu Water Balance and Evaporation](../articles/rotoehu-water-balance.html)** - 
  A case study demonstrating water balance approaches and evaporation estimation 
  for a shallow lake with ungauged inflows.

