---
title: "Introduction to AEME"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to AEME}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(AEME)
## basic example code
tmpdir <- tempdir()
aeme_file <- system.file("extdata/lernzmp/LID45819.rds", package = "AEME")
aeme <- readRDS(aeme_file)
aeme <- remove_output(aeme)
# slotNames(aeme)

lke <- lake(aeme)
tme <- time(aeme)
cfg <- configuration(aeme)
obs <- observations(aeme)
inp <- input(aeme)
inf <- inflows(aeme)
outf <- outflows(aeme)
wbal <- water_balance(aeme)
outp <- output(aeme)
pars <- parameters(aeme)

# convert list into a tree structure
lake_names <- paste0(names(lke), collapse = ", ")
time_names <- paste0(names(tme), collapse = ", ")
config_names <- paste0(names(cfg), collapse = ", ")
obs_names <- paste0(names(obs), collapse = ", ")
input_names <- paste0(names(inp), collapse = ", ")
inflow_names <- paste0(names(inf), collapse = ", ")
outflow_names <- paste0(names(outf), collapse = ", ")
water_balance_names <- paste0(names(wbal), collapse = ", ")
output_names <- paste0(names(outp), collapse = ", ")
parameter_names <- paste0(names(pars), collapse = ", ")

```

## Summary

The AEME package hosts three one-dimensional hydrodynamic models: the DYnamic
REservoir Simulation Model (DYRESM), the General Lake Model (GLM), and the
General Ocean Turbulence Model (GOTM, which has been adapted for closed basins
for application to lakes and reservoirs). The models can be coupled to their
corresponding water quality models, the DYRESM-CAEDYM (Computational Aquatic
Ecosystem Dynamics Model), GLM-AED2 (Aquatic Ecosystem Dynamics Model 2), and
GOTM-WET (Water Ecosystem Tool).

Key aspects of the AEME package include:

-   Defined S4 class for `aeme` objects

-   Configuration of models from common and standardised inputs

-   Standardised calibration, manipulation and visualisation

## AEME object {#sec-aeme-object}

### Description {#sec-description}

The `aeme` object is the main object in the AEME package. It is an S4 class that
contains all the information required to run a model. The `aeme` object contains
the following slots:

-   [**lake**](#sec-lake) - a list object containing information about the lake
    (`r lake_names`)
-   [**time**](#sec-time) - a list object containing information about the time
    (`r time_names`)
-   [**configuration**](#sec-configuration) - a list object containing
    information about the configuration (`r config_names`)
-   [**observations**](#sec-observations) - a list object containing information
    about the observations (`r obs_names`)
-   [**inputs**](#sec-inputs) - a list object containing information about the
    inputs (`r input_names`)
-   [**inflows**](#sec-inflows) - a list object containing information about the
    inflows (`r inflow_names`)
-   [**outflows**](#sec-outflows) - a list object containing information about
    the outflows (`r outflow_names`)
-   [**water_balance**](#sec-water_balance) - a list object containing
    information about the water balance configuration (`r water_balance_names`)
-   [**parameters**](#sec-parameters) - a data.frame describing the parameters
    to be input with column names (`r parameter_names`)
-   [**output**](#sec-output) - a list object containing information about the
    outputs (`r output_names`)

#### Lake {#sec-lake}

The `lake` slot contains information about the lake. It is a list object that
contains the following objects:

-   name - Name of the lake (character).

-   id - Lake ID (character or numeric).

-   **latitude** - Latitude of the lake (numeric). If not provided, the latitude
    will be calculated from the centroid of the shape.

-   **longitude** - Longitude of the lake (numeric). If not provided, the
    longitude will be calculated from the centroid of the shape.

-   **elevation** - Elevation of the lake (numeric).

-   shape - Shape of the lake (sf object). The shape of the lake can be
    represented as a polygon. The centroid of the polygon will be used to
    calculate the latitude and longitude of the lake.

-   **depth** - Max depth of the lake (m) (numeric). Depth and area are used to
    generate a simple hypsographic curve if none is provided in the `inputs`
    slot.

-   **area** - Surface area of the lake (m2) (numeric)

Objects in **bold** are required for building and running the model.

#### Time {#sec-time}

The `time` slot contains information about the time. It is a list object that
contains the following objects:

-   **start** - Start date of the simulation (character). The start date must be
    in the format `YYYY-MM-DD HH:MM:SS`.

-   **end** - End date of the simulation (character). The end date must be in
    the format `YYYY-MM-DD HH:MM:SS`.

-   **timestep** - Timestep of the simulation (numeric). The timestep must be in
    seconds.

-   **spin_up** - List of spin up periods of the simulation for each model
    (numeric). The spin up period must be in days. The spin up period is the
    period of time that the model is run before the simulation period. The spin
    up period is used to initialise the model which is then discarded when
    examining the simulation period.

#### Configuration {#sec-configuration}

The `configuration` slot is a list that contains each of the model
configurations. This includes the configurations files for the hydrodynamic
components as well as the aquatic ecosystem model components:

| Model | Hyrodynamic | Ecosystem |
|--------------------|---------------------------|---------------------------------|
| DYRESM-CAEDYM | *.cfg* file and *.par* file | *.con*, *caedym3p1.bio, caedym3p1.chm* and *caedym3p1.sed* files |
| GLM-AED | *glm3.nml* file | *aed2.nml*, *phytos.nml*, *zoops.nml* files |
| GOTM-WET | *gotm.yaml* and *output.yaml* files | *fabm.yaml* file |

: Files for hydrodynamic and ecosystem models.

When `build_aeme()` is ran, the `model_controls` data.frame that is passed to 
the function as an argument is added to the `configuration` slot.

For more details on the `model_controls`, see it's section [below](#sec-model_controls).

#### Observations {#sec-observations}

The `observations` slot is a list that contains observations of in-lake (`lake`)
variables (e.g. water temperature, chlorophyll-a, dissolved oxygen etc.) and
water level (`level`). The observations are used to assess model performance
using the function \code{assess_model()} and also to calibrate the model using
the [aemetools](https://github.com/limnotrack/aemetools) package.

The `lake` observations are stored in a data frame with the following columns:

-   **Date** - Date of the observation (character). The date must be in the
    format `YYYY-MM-DD HH:MM:SS`.

-   **depth** - Depth of the observation (m) (numeric). The depth must be in
    metres.

-   **var** - Variable name of the observation (character). The variable names
    an input preparation are designed in the [AEME inputs
    article](aeme-inputs.html).

-   **value** - Value of the observation (numeric).

The `level` observations are stored in a data frame with the following columns:

-   **Date** - Date of the observation (character). The date must be in the
    format `YYYY-MM-DD HH:MM:SS`.

-   **value** - Value of the observation (numeric).

#### Input {#sec-inputs}

The `input` slot is a list that contains the following objects:

-   init_profile - profile to initialise the lake simulation (data.frame). It
    has the columns: "depth", "temperature" and "salt". If this is not provided,
    it is automatically generated using the values in the
    [`model_controls`](#sec-model_controls).

-   init_depth - depth of the lake when initialising the model (vector), If not
    provided, will assume the depth from the `hypsograph`.

-   **hypsograph** - the lake hypsograph. This is a data.frame with the columns
    "elev", "depth" and "area". If you do not have this data, you can generate
    one using the `generate_hypsograph()` function.

-   **meteo** - the meteorological data. A data.frame which at a minimum must
    contain the columns date ("Date"), air temperature ("MET_tmpair), wind speed
    ("MET_wndspd") and shortwave radiation ("MET_radswd"). See [AEME
    Inputs](https://limnotrack.github.io/AEME/articles/aeme-inputs.html#meteorological-data)

-   use_lw - Logical switch to use longwave radiation. Defaults to TRUE.

-   **Kw** - the light extinction coefficient ($m^{-1}$).

#### Inflows {#sec-inflows}

The inflows slot is a list that contains the following objects:

-   data - named list of data.frames which contain the columns date ("Date"),
    flow ("HYD_flow"; $m^3 day^{-1}$). Temperature can also be provided
    ("HYD_temp"), however if not provided then it will be estimated using air
    temperature. The name for the list will be used as the stream identifier in
    each model. If `method` in the water_balance section is set to`3`, then
    "wbal" will be added to the data, this contains a separate inflow for each
    model, estimated using the different evaporation functions inside each
    model.

-   factor - list of scaling factors to be applied to the inflows. Named
    according to each model.

If no inflows are present then this slot will be empty.

#### Outflows {#sec-outflows}

The outflows slot is a list that contains the following objects:

-   data - named list of data.frames which contain the columns date ("Date"),
    flow ("HYD_flow"; $m^3 day^{-1}$). If `method` in the water_balance section
    is set to `2` or `3`, then "wbal" will be added to the data, this contains a
    separate outflow for each model, estimated using the different evaporation
    functions inside each model.

-   factor - list of scaling factors to be applied to the outflows. Named
    according to each model. However, this can also be passed as a parameter in
    the model_parameters section and calibrated there. This is the preferred
    method for applying scaling factors.

If no outflows are present then this slot will be empty.

#### Water balance {#sec-water_balance}

The water balance slot is generated internally when the `build_aeme()` function
is called and the `wb_method` is set to `2` or `3`. The slot contains:

-   use: define which lake level to use. Can either be observed ("obs") or
    modelled ("mod") lake level. Default = "obs".

-   method: This can be `1` (no inflows or outflows) or `2` (outflows
    calculated) or `3` (inflows and outflows calculated). The default is `2` .

-   data: list of two data.frames. "wbal" which contains the diagnostics for
    estimating evaporation and water balance for each model and "model" which
    contains the modelled lake water level if `use` is "mod".

#### Parameters {#sec-parameters}

The parameters slot contains a data.frame of parameters which are used when
building the model. These will update the default model parameters or scale the
meteorological or scale the inflows and outflows for each model.

The columns for the parameters data.frame are:

-   model - Either "dy_cd", "glm_aed" and "gotm_wet".

-   file - Either the name of the file e.g. "glm3.nml" for model specific files
    or "met" for meteorological variables or "inf" for inflow or "wdr" for
    outflows. (Outflows were initiallly referred to as withdrawals, hence the
    "wdr" notation, this will probably be updated to reflect the current
    outflows slot soon...).

-   name - Name of the parameter. If the name of the parameter is nested in a
    nml/yaml file, then the whole hierarchy needs to be provide with each level
    separated by a "/" e.g. "light/Kw" for Kw in GLM-AED.

-   value - Value of the parameter.

-   min - Minimum range of the parameter. This is used in the
    `aemetools::calib_aeme()` and `aemetools::sa_aeme()` function.

-   max - Maximum range of the parameter. This is used in the
    `aemetools::calib_aeme()` and `aemetools::sa_aeme()` function.

-   module - Module that the parameter is in. Not necessary, but is helpful for
    identifying which parameters are in which module for GLM-AED and GOTM-WET.

-   group - Phytoplankton group. This is only used for GOTM-WET phytoplankton
    parameters.

There is a function `get_aeme_parameters()` which allows you to select
parameters based on model and module. See `?get_aeme_parameters()` for more
details.

#### Output {#sec-output}

The output slot is a list that is updated when `run_aeme()` is executed. Each
time it is executed, variable outputs designated in the `model_controls` will be
added to the `Aeme` object.

### Model controls {#sec-model_controls}

The `model_controls` is a data.frame generated by the `get_model_controls()`. 
The data.frame has the columns:

-   var_aeme - character; the AEME variable name

-   simulate - logical; add the variable to the `Aeme` object

-   inf_default - numeric; default value to use in the inflows if none present 
    in the inflows. This is particularly important for configuring water 
    chemistry for the inflows if `use_bgc = TRUE`.
    
-   initial_wc - numeric; value to use in initialising the model. This will be 
    automatically updated if the variable is present in the `observations` slot.
    
-   initial_sed - numeric; value to use in intialising the sediment module for 
    the DYRESM-CAEDYM model.
    
-   conversion_aed - numeric; factor to multiply by to convert to GLM-AED units.

When the model is built the mode


### Creation

The `aeme` object can be created using the `aeme_constructor()` function. It 
requires the basic inputs to 
from a YAML file using the `yaml_to_aeme`
function. The YAML file contains all the information required to run the model.

```{r warning=FALSE, message=FALSE, eval=FALSE}
# Define lake list
lat <- -36.88921
lon <- 174.4669
depth <- 13.08
area <- 153648

lake <- list(
    latitude = lat,
    longitude = lon,
    # elevation = elevation,
    depth = depth,
    area = area
  )
time <- list(
  start = as.POSIXct("2020-07-01"),
  stop = as.POSIXct("2022-06-30")
)

hypsograph <- generate_hypsograph(max_depth = depth, surface_area = area,
                                  volume_development = 1.2)

met <- aemetools::get_era5_land_point_nz(lat = lat, lon = lon,
                                         years = 2020:2022)

#' Define input list
input = list(
  hypsograph = hypsograph,
  meteo = met,
  Kw = 1.21
)

aeme <- aeme_constructor(lake = lake, time = time, input = input)

```
```{r}
slotNames(aeme)
```


## Manipulation

The `aeme` object can be manipulated using the `AEME` package functions. The
functions are defined by the slot names of the `aeme` object. For example, the
`lake` slot can be manipulated using the `lake` function.

```{r aeme-manipulation}
# Load lake data
lke <- lake(aeme)
# Print lake data to console
print(lke)

# Change lake name
lke[["name"]] <- "AEME"

# reassign lake data to aeme object
lake(aeme) <- lke

aeme

```

### Visualisation

The `aeme` object can be visualised simply using the `plot` function. The `plot`
function can be applied to the different slots of the `aeme` object. For
example, the `lake` slot can be visualised using the `plot` function.

```{r aeme-visualisation-lake, fig.width=7, fig.height=5}
plot(aeme, "lake")

```

```{r aeme-visualisation-input, fig.width=7, fig.height=5}
plot(aeme, "input")

```

## Build Model Ensemble

```{r}
model_controls <- get_model_controls()
inf_factor = c("dy_cd" = 1, "glm_aed" = 1, "gotm_wet" = 1)
outf_factor = c("dy_cd" = 1, "glm_aed" = 1, "gotm_wet" = 1)
model <- c("dy_cd", "glm_aed", "gotm_wet")
path <- "aeme"
aeme <- build_aeme(path = path, aeme = aeme, model = model,
                            model_controls = model_controls, inf_factor = inf_factor,
                            ext_elev = 5, use_bgc = TRUE)
aeme
```

```{r}
cfg <- configuration(aeme)
```
