---
title: "Set up AEME for a new lake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Set up AEME for a new lake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
unlink("45819_wainamu/", recursive = TRUE, force = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Setting up the Aquatic Ecosystem Model Ensemble is straightforward and can be
done in a few steps. This vignette will guide you through the process of setting
up the model for a lake in New Zealand.

This is a **practical tutorial** focused on workflow. For detailed information 
on the AEME models and ensemble approach, see [Introduction to 
AEME](intro-aeme.html). For comprehensive reference on inputs and the S4 
structure, see [AEME Inputs](aeme-inputs.html).

```{r setup}
library(AEME)
library(sf) # For spatial data
library(tmap) # For mapping
tmap_mode("view") # Set tmap mode to interactive view model
tmap_options(basemap.server = "OpenStreetMap") # Set the basemap to OpenStreetMap

```


# Lake data

The first step is to define the lake data. This includes the location of the
lake, the depth and area of the lake, and the elevation of the lake. See [Lake 
Slot](aeme-inputs.html#lake-slot) for complete field requirements.


```{r lake-location}
# Define the location of the lake
lat <- -36.8898
lon <- 174.46898
```

```{r view-in leaflet}
# View the location of the lake in a map

coords <- data.frame(lat = lat, lon = lon) |> 
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

tm_shape(coords) +
  tm_bubbles() +
  tm_view(set_view = c(lon, lat, 16))

```

## Get lake shapefile

If you have a shapefile for the lake, you can use the `sf` package to read it 
in. However, if you do not have a shapefile you can download most lake 
shapefiles from [OpenStreetMap](https://www.openstreetmap.org/) using the [`osmdata`](https://docs.ropensci.org/osmdata/) package. 

```{r get-lake-shapefile, eval=FALSE}
library(osmdata)

# Get lake shapefile
osm_data <- opq(bbox = "New Zealand") |> 
  add_osm_feature(key = "name", value = "Lake Wainamu") |> 
  osmdata_sf()

# Extract lake polygon
lake <- osm_data$osm_multipolygons |> 
  st_make_valid()
lake
```
```{r read-lake-shapefile, echo=FALSE}
lake <- readRDS(system.file("extdata/lake_shapefile.rds", package = "AEME"))
```

Visualise the lake shapefile on a map.

```{r view-lake-shapefile}
# View the lake shapefile
tm_shape(lake) +
  tm_borders(lwd = 2) 
```

As you can see it matches the location of the lake we defined earlier and that 
it perfectly matches the outline from the OpenStreetMap polygon.

## Depth and area data

The depth and area of the lake are required for the model. These can be
obtained from a variety of sources. For this example, we will use the known 
depth of the lake (13.07 m), however for the area we will calculate this from
the shapefile.

```{r set-depth}
# Set depth & area
depth <- 13.07 # Depth of the lake in metres
```

```{r calc-area}
# Calculate the area of the lake in m2
area <- st_area(lake) |> 
  units::set_units("m^2") |> 
  as.numeric()
area

```

## Elevation data

Elevation data can be acquired for New Zealand from the digital elevation model
hosted on the LINZ Data Service. There is a wrapper function for this in the
`aemetools` package. This requires an API key from LINZ.

You can easily create a key on the LINZ website: https://data.linz.govt.nz/ or use the
function within the `aemetools` package to create one.

```{r get-linz-api-key, eval = FALSE}
aemetools::create_linz_key()
```

Then adding it to your `.Renviron` file.

```{r add-linz-api-key, eval = FALSE}
# Add the LINZ API key to your .Renviron file
aemetools::add_linz_key(key = "your_key_here")
```

The `get_dem_value` function will return the elevation of the lake in metres
above sea level.

```{r get-elevation}
# Get the elevation of the lake
key <- Sys.getenv("LINZ_KEY")
elevation <- aemetools::get_dem_value(lat = lat, lon = lon, key = key)
elevation # in metres above sea level

```
```{r, set-elevation-if-na, include=FALSE}
if (length(elevation) > 1) {
if (!elevation[["success"]]) {
  elevation <- 29
}
}

```

```{r}
elevation
```


We will now create a list of the lake data. This will be used to construct the
AEME object.

```{r create-lake-list}
# Define lake list
lake <- list(
    name = "Wainamu",
    id = 45819,
    latitude = lat,
    longitude = lon,
    elevation = elevation,
    depth = depth,
    area = area
  )
```

# Time data

The time data defines the simulation period. See [Time 
Slot](aeme-inputs.html#time-slot) for details on date formats and spin-up 
configuration.

```{r time-data}
# Define start and stop times
start <- "2020-08-01 00:00:00"
stop <- "2021-06-30 00:00:00"

time <- list(
    start = start,
    stop = stop
  )
```



# Input data

The AEME `input` slot requires meteorological data, a hypsograph, and the light 
extinction coefficient. See [Input Slot](aeme-inputs.html#input-slot) for 
detailed requirements.

## Meteorological data

### Download ERA5 data

We will use the `aemetools` package to download the ERA5 meteorological data for
the location of our lake. This works for all locations around the world. 
However, its date range is only from 1900-2021.

```{r get-era5-met}
# Get ERA5 meteorological data
met <- aemetools::get_era5_isimip_point(lat = lat, lon = lon, years = 2020:2021)

```

View the summary of the meteorological data. The units have been converted to 
more common units used in aquatic ecosystem modelling.

```{r met-summary}
# Summary of meteorological data
summary(met)
```

For details on meteorological variables and derived fields, see [Meteorological 
Data](aeme-inputs.html#meteorological-data).

The depth of this lake is 13.07 m, the area is 152343 m2, and the light 
extinction coefficient (Kw) is 1.31 m-1.


```{r}
# Set Kw
Kw <- 1.31 # Light extinction coefficient in m-1
```


## Hypsograph data

If you have hypsograph data for the lake, you can use it as input for the model.
This is a critical input for the model, as it defines the relationship between
the lake area and the lake elevation.

However, if you do not have hypsograph data, the model will use a simple
cone-shaped hypsograph based on the lake depth and area. This is not ideal, but
it will work for this example.

For complete details on hypsograph requirements and generation, see [Input 
Slot - Hypsograph](aeme-inputs.html#input-slot).

Required column names for the hypsograph data are `area`, `elev`, and `depth`.

```{r generate-hypsorgaph}
# Generate a simple hypsograph
hypsograph_simple <- data.frame(area = c(area, 0), 
                         elev = c(elevation, elevation - depth),
                         depth = c(0, -depth))
hypsograph_simple
```

```{r plot-hypsograph}
# Plot the hypsograph
library(ggplot2)

ggplot(hypsograph_simple, aes(x = area, y = elev)) +
  geom_line() +
  geom_point() +
  xlab("Area (m2)") +
  ylab("Elevation (m)") +
  theme_bw()

```

As you can see, the hypsograph is a simple cone shape. Ideally, you would have
more detailed hypsograph data for your lake.

If you have information regarding the maximum depth of the lake, the surface
area and an estimate of volume development, you can generate a hypsograph using
the `generate_hypsograph` function. The `volume_development` parameter is a
scaling factor for the volume development of the lake. Values below 1.5 are 
lakes with a concave hypsograph, values above 1.5 are lakes with a
convex hypsograph, and values of 1.5 are lakes with a linear hypsograph. 

For Wainamu Lake, we will use a volume development of 1.62 which was calculated 
from a bathymetry survey of the lake. You can view this on the 
[LERNZmp platform](https://limnotrack.shinyapps.io/LERNZmp/).

```{r generate-hypsograph}

# Generate a hypsograph
hypsograph <- generate_hypsograph(max_depth = depth, surface_area = area,
                                  volume_development = 0.5, elev = elevation,
                                  ext_elev = 1)

ggplot(hypsograph, aes(x = area, y = elev)) +
  geom_line() +
  geom_point() +
  geom_line(data = hypsograph_simple, aes(x = area, y = elev), 
            linetype = "dashed") +
  xlab("Area (m2)") +
  ylab("Elevation (m)") +
  theme_bw()

```




```{r make-aeme-list}
# Define input list
input <- list(
    init_depth = depth,
    hypsograph = hypsograph,
    meteo = met,
    use_lw = TRUE,
    Kw = Kw
  )
```

# Construct the AEME object

The `aeme_constructor` function will take the input data and construct the AEME
object. The minimum inputs are the `lake`, `time`, and `input` data.


```{r construct-aeme-object}
# Construct AEME object
aeme <- aeme_constructor(lake = lake, 
                         time = time,
                         input = input)

```

## View AEME object

The AEME object can be inspected by printing it to the console. This will show
the inputs that have been used to construct the object along with default values
for inputs not provided.

```{r view-aeme-object}
aeme
```

In the configuration section of the output, under "Physical" and 
"Biogeochemical" for each model are labelled "Absent". This is because the 
model configurations have not been built. This is done in the next step.

# Building the AEME ensemble

## Model controls

The model controls define which variables to simulate. For details on the 
structure and customization, see [Configuration 
Slot](aeme-inputs.html#configuration-slot).

The model controls has the following columns:

- `var_aeme`: The variable name in the AEME object
- `simulate`: Whether to simulate the variable
- `inf_default`: The default inflow value
- `initial_wc`: The initial water column value
- `initial_sed`: The initial sediment value

```{r get-model-controls}
# Get model controls
model_controls <- get_model_controls()
model_controls
```

## Build the ensemble

The `build_aeme` function creates model configurations from the AEME object. For 
background on the three models, see [The Three Models](intro-aeme.html#three-models).

The `model` argument is a character vector of the models
to include in the ensemble. The models available are `dy_cd`, `glm_aed`, and
`gotm_wet`.

```{r select-models}
# Select models
model <- c("dy_cd", "glm_aed", "gotm_wet")
```

```{r build-ensemble}

# Path for model directory
path <- "aeme"

# Build ensemble
aeme <- build_aeme(aeme = aeme, model = model, model_controls = model_controls, 
                   path = path)

print(aeme)
```

By default, the `build_aeme` function will build the
file configuration for each model. This will create the necessary files for
each model to run. The files are also stored in the `aeme` object in the
`configuration` slot with a list for hydrodynamic and ecosystem model 
configurations. 

```{r view-files}
# View the files
cfg <- configuration(aeme)
names(cfg[["glm_aed"]])
```

All of the information and data needed to run an ensemble of models is now
contained within the `aeme` object. This allows for easy storage of all the data
and also for easy sharing of the data with others. Sharing the `aeme` object
with others allows them to run the ensemble of models without needing to
reconstruct the object. 

```{r run-ensemble}
# Run the ensemble
aeme <- run_aeme(aeme = aeme)
```

## View the output

The output from the model run is stored in the `output` slot of the `aeme` object.
This is a list with a list for each model. The list contains the output data
from the model run. 

```{r view-output}
# View the output
plot_output(aeme = aeme)
```


## Saving the AEME object

Saving the `aeme` object to a file can be done using the `saveRDS` function. 
This will save the object to a file with the `.rds`.

```{r save-aeme}
# Save the AEME object
saveRDS(aeme, "aeme.rds")
```

## Using AEME in pipes

The AEME functions can be used in pipes to make the workflow more efficient. For
example, the `build_aeme` function can be used in a pipe to build the ensemble and
then the `run_aeme` function can be used in the same pipe to run the ensemble.

This allows for a more streamlined workflow and reduces the amount of code needed to
build and run the ensemble. This is especially useful when building and running
multiple ensembles for different lakes or different scenarios.

This approach has been used with the [{targets}](https://books.ropensci.org/targets/) 
package to build and run multiple ensembles for different lakes and scenarios in
a reproducible workflow.

```{r pipe-workflow, eval = FALSE}
# Build and run the ensemble in a pipe
aeme <- aeme |> 
  add_obs(lake = lake_obs, level = level_obs) |> 
  add_hypsograph(hypsograph = hypsograph) |>
  add_met(met = met) |> 
  add_inflows(data = inf_data) |> 
  add_outflows(data = outf_data) |>
  add_param(param = param) |> 
  build_aeme(model = model, use_bgc = TRUE, path = path) |> 
  run_aeme()
```

# Troubleshooting {#troubleshooting}

Common issues and solutions when setting up AEME:

## Model Build Errors

**Error: "Hypsograph elevation range does not include initial depth"**

- **Solution**: Ensure your hypsograph spans from below the lake bottom to above 
  the maximum expected water level. Use `ext_elev` parameter in 
  `generate_hypsograph()` or add rows to your hypsograph.

**Error: "Missing required meteorological variables"**

- **Solution**: Check your meteorological data has the required columns: `Date`, 
  `MET_tmpair`, `MET_wndspd`, `MET_radswd`. See [Meteorological 
  Data](aeme-inputs.html#meteorological-data) for details.

**Error: "Outflow elevation outside hypsograph range"**

- **Solution**: Verify outlet elevations fall within your hypsograph elevation 
  range. Use `-1` for surface outlets.

## Model Run Errors

**Error: "Model failed to run"**

- **Check**: Timestep may be too large for stability
- **Check**: Initial conditions may be unrealistic  
- **Check**: Meteorological data has no gaps or unrealistic values
- **Solution**: Run with `verbose = TRUE` to see detailed error messages

**Warning: "Simulated water level outside hypsograph range"**

- **Solution**: Extend hypsograph range using `ext_elev` parameter or check 
  water balance inputs

## Data Quality Issues

**Unrealistic model output**

- **Check**: Light extinction coefficient (Kw) is appropriate for lake type
- **Check**: Hypsograph is realistic (not just a simple cone)
- **Check**: Meteorological data units are correct
- **Check**: Inflow/outflow magnitudes are reasonable

**Models diverge significantly**

- **Expected**: Some divergence is normal due to different physics (see 
  [Ensemble Approach](intro-aeme.html#ensemble-approach))
- **Check**: Ensure all models use the same inputs
- **Consider**: Different models may be appropriate for different lake types

## Getting Help

If you encounter issues not covered here:

1. Check the [AEME GitHub Issues](https://github.com/limnotrack/AEME/issues)
2. Review the comprehensive [AEME Inputs](aeme-inputs.html) reference
3. Open a new issue with a reproducible example

# Next Steps {#next-steps}

Now that you have a working AEME ensemble, consider these extensions:

## Model Calibration and Sensitivity Analysis

Use the [aemetools](https://github.com/limnotrack/aemetools) package for:

- **Automated calibration**: Optimize parameters against observations
- **Sensitivity analysis**: Identify influential parameters
- **Uncertainty quantification**: Ensemble-based uncertainty estimates

```{r aemetools-example, eval=FALSE}
library(aemetools)

# Calibrate GLM-AED
calib_result <- calib_aeme(
  aeme = aeme,
  param = param,
  model = "glm_aed",
  vars_sim = c("HYD_temp", "CHM_oxy")
)

# Sensitivity analysis
sa_result <- sa_aeme(
  aeme = aeme,
  param = param,
  model = "glm_aed",
  method = "sobol"
)
```

See the [aemetools documentation](https://limnotrack.github.io/aemetools/) for 
details.

## Advanced Applications

Explore additional AEME features:

### Water Balance Estimation

Estimate unknown inflows/outflows from water level observations:

```{r water-balance-example, eval=FALSE}
aeme <- build_aeme(
  aeme = aeme,
  model = model,
  model_controls = model_controls,
  path = path,
  wb_method = 2,  # Estimate outflows
  use_wbal = "obs"
)
```

See `vignette("rotoehu-water-balance")` for a complete example.

### Reservoir Applications

For reservoirs with multiple outlets and selective withdrawal:

```{r reservoir-example, eval=FALSE}
# Add multiple outlets at different depths
aeme <- add_outflows(
  aeme,
  data = list(
    spillway = spillway_data,
    penstock = penstock_data
  ),
  elevation = list(
    spillway = -1,      # Surface
    penstock = 130      # Deep outlet
  )
)
```

See `vignette("reservoir-aeme")` for details.

### Scenario Analysis

Test management scenarios by modifying inputs:

```{r scenario-example, eval=FALSE}
# Scenario 1: Increased nutrient loading
aeme_scenario1 <- aeme
inf_data_enriched <- inf_data
inf_data_enriched$stream$PHS_frp <- inf_data$stream$PHS_frp * 2
aeme_scenario1 <- add_inflows(aeme_scenario1, data = inf_data_enriched)

# Scenario 2: Climate change (+2°C)
met_future <- met
met_future$MET_tmpair <- met_future$MET_tmpair + 2
aeme_scenario2 <- aeme
input(aeme_scenario2)$meteo <- met_future

# Run scenarios
aeme_scenario1 <- run_aeme(aeme_scenario1)
aeme_scenario2 <- run_aeme(aeme_scenario2)

# Compare results
plot_output(aeme_scenario1, var_sim = "PHY_tchla")
plot_output(aeme_scenario2, var_sim = "HYD_temp")
```

## Related Vignettes

- [Introduction to AEME](intro-aeme.html) - Models, ensemble approach, and 
  theory
- [AEME Inputs](aeme-inputs.html) - Comprehensive reference for all inputs
- [Reservoir Example](articles/reservoir-aeme.html) - Multi-outlet reservoir
- [Water Balance](articles/rotoehu-water-balance.html) - Estimating unknown 
  flows
- [LERNZmp Application](articles/lernzmp-aeme.html) - Multi-lake ensemble 
  modeling

## Additional Resources

- [AEME GitHub](https://github.com/limnotrack/AEME)
- [aemetools package](https://github.com/limnotrack/aemetools)
- [GLM-AED documentation](https://aquatic.science.uwa.edu.au/research/models/GLM/)
- [GOTM documentation](https://gotm.net/)
- [LERNZmp platform](https://limnotrack.shinyapps.io/LERNZmp/)
