---
title: "GLM-AED: The General Lake Model coupled with AED"
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
asciicast::init_knitr_engine(
  startup = quote({
    library(AEME)
    set.seed(1)
  }),
  echo = TRUE,
  echo_input = FALSE
)
options(AEME.inform = FALSE)
library(AEME)
```

<a href="#parameter-library" class="skip-btn">⬇ Skip to Parameter Library</a>

<style>
.skip-btn {
  display: inline-block;
  margin: 8px 0 16px 0;
  padding: 8px 16px;
  background-color: #2c7bb6;
  color: white !important;
  border-radius: 6px;
  text-decoration: none;
  font-size: 14px;
  font-weight: bold;
  transition: background-color 0.2s;
}
.skip-btn:hover {
  background-color: #1a5a8a;
  text-decoration: none;
}
</style>

## Introduction

### The General Lake Model (GLM)

The **General Lake Model (GLM)** is a one-dimensional (1-D), variable-layer
hydrodynamic model for lakes, reservoirs, and estuaries. Originally described
by Hipsey *et al.* (2019), GLM simulates the vertical profiles of temperature,
salinity, and density driven by surface heat exchange, shortwave radiation,
wind mixing, inflows, and outflows. Its key architectural features include:

- **Variable-thickness layer scheme** — layers merge and split dynamically,
  giving high resolution where it is needed (near the thermocline) and coarser
  resolution elsewhere.
- **Flexible inflow/outflow** — supports multiple rivers, surface spillways, and
  subsurface outlets at arbitrary elevations.
- **Rich surface heat-flux options** — sensible heat, latent heat, longwave, and
  shortwave radiation computed from standard meteorological inputs.
- **Sediment heat exchange** — allows depth-varying bed temperatures to drive
  conductive heat flux into the water column.
  
There is more detailed GLM documentation available from the [GLM website](https://aquatic.science.uwa.edu.au/research/models/GLM/),

Source code: :github: <https://github.com/AquaticEcoDynamics/GLM/tree/master>

Binaries: :github: <https://github.com/AquaticEcoDynamics/Binaries>

#### References

Hipsey, M.R., Bruce, L.C., Boon, C., Busch, B., Carey, C.C., Hamilton, D.P.,
Hanson, P.C., Read, J.S., de Sousa, E., Weber, M. & Winslow, L.A. (2019).
A General Lake Model (GLM 3.0) for linking with high-frequency sensor data
from the Global Lake Ecological Observatory Network (GLEON).
*Geoscientific Model Development*, 12, 473–523.
<https://doi.org/10.5194/gmd-12-473-2019>

---

### The Aquatic Ecosystem Dynamics (AED) library

The **Aquatic Ecodynamics (AED)** library (Hipsey *et al.*, 2013) is a modular
biogeochemical modelling framework designed to be coupled with hydrodynamic
models such as GLM. Each biogeochemical process is implemented as a
self-contained *module* that can be switched on or off independently, making it
easy to build simulations that range from simple oxygen tracking through to
full nutrient–phytoplankton–zooplankton food-web dynamics.

There is a detailed AED manual available from the [AED website](https://aquaticecodynamics.github.io/aed-science/),
but here we provide a brief overview of the key modules relevant for GLM-AED.

#### AED modules available in AEME

| Module | AED name | Description |
|--------|----------|-------------|
| Sediment flux | `aed_sedflux` | Sediment–water interface exchange of O~2~, nutrients, and silica. Supports constant, constant-2D (depth-zone-specific), and dynamic flux models. |
| Oxygen | `aed_oxygen` | Dissolved-oxygen dynamics including reaeration, sediment oxygen demand (SOD), and photosynthesis/respiration coupling. |
| Silica | `aed_silica` | Reactive silica cycling — important for diatom growth. |
| Nitrogen | `aed_nitrogen` | Full dissolved inorganic nitrogen cycle: ammonium (NH~4~^+^), nitrite (NO~2~^−^), nitrate (NO~3~^−^), N~2~O; nitrification, denitrification, N-fixation. |
| Phosphorus | `aed_phosphorus` | Dissolved reactive phosphorus (DRP / FRP) including redox-sensitive sediment release. |
| Organic matter | `aed_organic_matter` | Particulate and dissolved organic carbon (POC/DOC) and organic nitrogen and phosphorus pools; mineralisation, decomposition. |
| Phytoplankton | `aed_phytoplankton` | Multi-group phytoplankton dynamics: growth, respiration, nutrient uptake, light limitation, settling, mortality. Default groups: cyanobacteria, green algae, diatoms. |
| Zooplankton | `aed_zooplankton` | Zooplankton grazing and dynamics. |
| Macrophytes | `aed_macrophyte` | Submerged and emergent macrophyte dynamics. |
| Totals | `aed_totals` | Diagnostic aggregates: TN, TP, TOC, chlorophyll-*a*. |

: AED biogeochemical modules available in AEME.

### GLM-AED Parameter Library {#parameter-library}

The `glm_aed_parameter_library` dataset provides a comprehensive list of all 
parameters used in the GLM-AED configuration, including their default values, 
units, and typical ranges. This library serves as a reference for users to 
understand the parameters governing the model behaviour and to guide 
parameterisation for specific applications. It also includes metadata such as 
the associated AED module and a brief description of each parameter's role in 
the model and a web source link for further information.

```{r glm-aed-lib, echo=FALSE}
# Prep the data
df <- glm_aed_parameter_library |>
  dplyr::mutate(
    # Make source URLs into clickable hyperlinks
    source = ifelse(
      !is.na(source) & grepl("^http", source),
      paste0('<a href="', source, '" target="_blank">&#x1F517;</a>'),
      source
    ),
    # Make symbol italic if not NA
    symbol = ifelse(!is.na(symbol), paste0("<i>", symbol, "</i>"), NA)
  )

DT::datatable(
  df,
  escape = FALSE,  # allows HTML rendering in cells
  rownames = FALSE,
  colnames = c(
    "Module", "Group", "Parameter", "Label",
    "Symbol", "Description", "Units", "Default", "Source"
  ),
  filter = "top",  # per-column filters
  extensions = c("Buttons", "FixedHeader"),
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel", "pdf"),
    pageLength = 10,
    lengthMenu = c(10, 20, 50, 100),
    fixedHeader = TRUE,
    scrollX = TRUE,
    autoWidth = TRUE,
    columnDefs = list(
      list(width = "200px", targets = 5),   # Description column wider
      list(width = "80px",  targets = c(0, 1, 6, 7)),  # narrow cols
      list(className = "dt-center", targets = c(4, 6, 7, 8))  # center symbol/units/default/source
    )
  )
) |>
  DT::formatStyle(
    "module",
    fontWeight = "bold"
  ) |>
  DT::formatStyle(
    columns = 1:9,
    fontSize = "13px"
  ) |>
  DT::formatStyle(
    "default",
    color = "#2c7bb6",
    fontWeight = "bold"
  )

```

### AED Phytoplankton Group Parameters

```{r aed-phyto-lib, echo=FALSE}

phyto_file <- system.file(
  "extdata/glm_aed/aed_phyto_pars.csv",
  package = "AEME"
)
aed_phyto_params <- readr::read_csv(phyto_file, show_col_types = FALSE)

param_desc <- aed_phyto_pars |>
  dplyr::select(parameter_name, description) |> 
  dplyr::distinct(parameter_name, .keep_all = TRUE)

df_wide <- aed_phyto_params |> 
  dplyr::left_join(param_desc, by = c("p_name" = "parameter_name"))

# Get group column names (everything except p_name and description)
group_cols <- setdiff(names(df_wide), c("p_name", "description"))
n_groups   <- length(group_cols)

DT::datatable(
  df_wide,
  escape    = FALSE,
  rownames  = FALSE,
  filter    = "top",
  extensions = c("Buttons", "FixedHeader", "FixedColumns"),
  options   = list(
    dom        = "Bfrtip",
    buttons    = c("colvis", "csv", "excel"),
    pageLength = 25,
    lengthMenu = c(10, 25, 50),
    fixedHeader = TRUE,
    scrollX    = TRUE,
    scrollY    = "600px",
    fixedColumns = list(leftN = 1, rightN = 1),
    autoWidth  = FALSE,
    columnDefs = list(
      list(width = "120px", targets = 0),
      list(width = "80px",  targets = seq_len(n_groups)),
      list(width = "300px", targets = n_groups + 1),
      list(className = "dt-center", targets = seq_len(n_groups))
    )
  )
) |>
  DT::formatStyle(
    "p_name",
    fontWeight = "bold",
    fontFamily = "monospace"
  ) |>
  DT::formatStyle(
    "description",
    color    = "grey40",
    fontSize = "12px"
  ) |>
  DT::formatRound(
    columns = group_cols,
    digits  = 3
  ) |>
  DT::formatStyle(
    columns   = 1:(n_groups + 2),
    fontSize  = "13px"
  )

```

---

## Getting started

```{r setup}
library(AEME)
library(ggplot2)
library(dplyr)
```

We will use the example lake dataset bundled with the AEME package throughout
this vignette.

```{r load-aeme}
aeme_file <- system.file("extdata/aeme.rds", package = "AEME")
aeme     <- readRDS(aeme_file)
aeme
```

```{asciicast show-aeme}
#| echo: FALSE
aeme_file <- system.file("extdata/aeme.rds", package = "AEME")
aeme     <- readRDS(aeme_file)
aeme
```

---

## Building a GLM-AED simulation

### Model controls

`get_model_controls()` returns a data frame that governs which biogeochemical
variables are simulated and how they are initialised. Pass `use_bgc = TRUE` to
enable the full suite of water-quality variables.

```{r model-controls}
model_controls <- get_model_controls(use_bgc = TRUE)
head(model_controls)
```

You can narrow the set of simulated variables using `set_vars_sim()`:

```{r set-vars}
vars_sim <- c(
  "HYD_strat",   # stratification flag
  "HYD_temp",    # water temperature
  "HYD_thmcln",  # thermocline depth
  "CHM_oxy",     # dissolved oxygen
  "CHM_oxycln",  # oxycline depth
  "NIT_amm",     # ammonium
  "NIT_nit",     # nitrate
  "NIT_tn",      # total nitrogen
  "PHS_frp",     # filterable reactive phosphorus
  "PHS_tp",      # total phosphorus
  "PHY_tchla"    # total chlorophyll-a
)

model_controls <- set_vars_sim(
  model_controls = model_controls,
  vars_sim       = vars_sim
)
```

### Build the model

We will build the model in a directory called `aeme`,

```{r config-path}
path <- "aeme"
```


`build_aeme()` translates the AEME object into all configuration files required
by GLM-AED. Setting `use_bgc = TRUE` writes the `aed/aed.nml` file and
supporting CSV parameter files for phytoplankton, zooplankton, and macrophytes.

```{r build-aeme}
model <- "glm_aed"

aeme <- build_aeme(
  aeme           = aeme,
  model          = model,
  path           = path,
  model_controls = model_controls,
  ext_elev       = 3,
  use_bgc        = TRUE
)
```

The configuration files are stored in the `configuration` slot of the `aeme`
object. For GLM-AED the slot contains the parsed `glm3.nml` and `aed/aed.nml`:

```{r inspect-config}
cfg <- configuration(aeme)
names(cfg[["glm_aed"]])
```

---

## GLM-AED specific features

### Selecting AED biogeochemical modules

By default, all AED modules are enabled. `set_glm_aed_models()` lets you choose
a subset — for example, running hydrodynamics coupled with oxygen and nutrient
cycling *only*, without phytoplankton or zooplankton:

```{r set-models}
# Rebuild first so we have a clean state
aeme <- build_aeme(
  aeme           = aeme,
  model          = model,
  model_controls = model_controls,
  path           = path,
  ext_elev       = 5,
  use_bgc        = TRUE
)

# Switch on all available modules (the default)
aeme <- set_glm_aed_models(
  aeme       = aeme,
  path       = path,
  aed_models = c(
    "aed_sedflux",
    "aed_oxygen",
    "aed_silica",
    "aed_nitrogen",
    "aed_phosphorus",
    "aed_organic_matter",
    "aed_phytoplankton",
    "aed_zooplankton",
    "aed_macrophyte",
    "aed_totals"
  )
)
```

To run a minimal simulation with only oxygen and nutrient cycling:

```{r set-models-minimal, eval=FALSE}
aeme <- set_glm_aed_models(
  aeme       = aeme,
  path       = path,
  aed_models = c(
    "aed_sedflux",
    "aed_oxygen",
    "aed_nitrogen",
    "aed_phosphorus"
  )
)
```

---

### AED module dependency hierarchy

AED modules are not independent of one another — most reference state
variables that are only computed if another module is switched on. For
example, `aed_nitrogen` reduces ammonium using `OXY_oxy` as its oxidant, so it
requires `aed_oxygen` to also be active; if it is not, GLM aborts at runtime
with an "Undefined variable" error rather than silently ignoring the missing
link.

`build_aeme()` resolves this automatically: when you set `simulate = TRUE`
for a biogeochemical variable via `model_controls`, `initialise_aed()`
determines which module that variable belongs to, then walks the dependency
graph below to a fixed point, adding every module that dependency chain
requires — you never need to enable a module's prerequisites by hand. The
one place this *isn't* automatic is `set_glm_aed_models()` (used above),
which sets the active-module list exactly as given, so it is up to you to
include the full chain if you call it directly.

The dependency graph, verified directly against the target-variable links
configured in AEME's bundled `aed.nml` template:

```
aed_sedflux
  └─ aed_oxygen           (sediment oxygen demand)
       ├─ aed_silica          (silica reactant = OXY_oxy)
       ├─ aed_nitrogen        (nitrification reactant = OXY_oxy; also needs aed_sedflux directly)
       ├─ aed_phosphorus      (P release reactant = OXY_oxy; also needs aed_sedflux directly)
       └─ aed_organic_matter  (mineralisation reactant = OXY_oxy)
            ├─ needs aed_nitrogen        (DON mineralisation product)
            ├─ needs aed_phosphorus      (DOP mineralisation product = PHS_frp)
            ├─ aed_zooplankton      (excretion/mortality targets)
            └─ aed_phytoplankton    (uptake/excretion/mortality targets;
                                      needs aed_oxygen, aed_nitrogen,
                                      aed_phosphorus, aed_silica, and
                                      aed_organic_matter directly)
                 └─ aed_totals      (TN/TP/TOC totals aggregate state
                                      variables owned by aed_nitrogen,
                                      aed_phosphorus, aed_organic_matter,
                                      and aed_phytoplankton)

aed_macrophyte  (no enforced dependencies — no target-variable links in the
                 bundled template)
```

| Module | Required dependencies | Why |
|--------|-----------------------|-----|
| `aed_sedflux` | — | Sediment flux is the base of the chain; nothing else must be active first. |
| `aed_oxygen` | `aed_sedflux` | Sediment oxygen demand flux comes from `aed_sedflux`. |
| `aed_silica` | `aed_oxygen` | Silica reactant variable is `OXY_oxy`. |
| `aed_nitrogen` | `aed_oxygen`, `aed_sedflux` | Nitrification reactant is `OXY_oxy`; sediment NH~4~/NO~3~ fluxes come from `aed_sedflux`. |
| `aed_phosphorus` | `aed_oxygen`, `aed_sedflux` | Redox-sensitive P release reactant is `OXY_oxy`; sediment FRP flux comes from `aed_sedflux`. |
| `aed_organic_matter` | `aed_oxygen`, `aed_nitrogen`, `aed_phosphorus` | Mineralisation reactant is `OXY_oxy`; DON/DOP mineralisation products feed `aed_nitrogen`/`aed_phosphorus`. |
| `aed_phytoplankton` | `aed_oxygen`, `aed_nitrogen`, `aed_phosphorus`, `aed_silica`, `aed_organic_matter` | Growth/uptake/excretion/mortality target variables span all five modules (e.g. diatom silica uptake, nutrient uptake, DOM excretion). |
| `aed_zooplankton` | `aed_organic_matter` | Excretion and mortality products are organic-matter state variables. |
| `aed_totals` | `aed_nitrogen`, `aed_phosphorus`, `aed_organic_matter`, `aed_phytoplankton` | Its `TN_vars`/`TP_vars`/`TOC_vars` lists aggregate state variables owned by these four modules (e.g. `NIT_nit`, `PHS_frp`, `OGM_doc`, `PHY_green`). |
| `aed_macrophyte` | — | No target-variable links to other modules in the bundled template. |

: AED module dependency requirements enforced by `initialise_aed()`.

For example, requesting only `NIT_tn` (total nitrogen) in `model_controls`
looks like a single-variable request, but because it is a diagnostic total
computed by `aed_totals`, and `aed_totals` in turn depends on nitrogen,
phosphorus, organic matter, and phytoplankton, `build_aeme()` activates the
following module set automatically:

```{r nit-tn-example}
mc <- get_model_controls(use_bgc = TRUE) |>
  set_vars_sim(vars_sim = "NIT_tn")

aeme_nit_tn <- build_aeme(
  aeme           = aeme,
  model          = model,
  path           = path,
  model_controls = mc,
  ext_elev       = 5,
  use_bgc        = TRUE
)

config_files <- get_model_config_files(aeme)


nml <- read_nml(config_files$glm_aed["aed"])
get_nml_value(nml, "models")
```

---

### Sediment zones

One of the most important GLM-AED features for water-quality simulation is
**depth-varying sediment parameters**. GLM divides the lake bed into *sediment
zones*, each with its own temperature regime and (when using `aed_sedflux` in
`Constant2d` mode) its own sediment–water interface fluxes of oxygen and
nutrients.

Deeper zones typically accumulate more organic matter, experience longer periods
of anoxia, and therefore have higher sediment oxygen demand (SOD) and nutrient
release than shallow littoral zones.

#### Estimating zone boundaries from the hypsograph

`estimate_sed_zones()` automatically partitions the hypsograph into an
appropriate number of zones by detecting natural breakpoints in the
depth–area relationship:

```{r sed-zones}
hypsograph   <- get_hypsograph(aeme)
zone_heights <- estimate_sed_zones(hypsograph)
zone_heights
```

The returned vector gives the *upper height* of each zone measured from the lake
bed (metres). The last value equals the maximum lake depth.

#### Generating GLM sediment parameters

`glm_sed_params()` builds a parameters data frame (compatible with the AEME
`parameters` slot) for the GLM sediment module:

```{r glm-sed-params}
sed_params <- glm_sed_params(
  n_zones            = length(zone_heights),
  zone_heights       = zone_heights,
  sed_temp_mean      = c(14, 16, 18)[seq_along(zone_heights)],
  sed_temp_amplitude = c(6, 4, 2)[seq_along(zone_heights)],
  sed_temp_peak_doy  = rep(30, length(zone_heights))
)
sed_params
```

Add these parameters to the `aeme` object so that they are applied during the
next `build_aeme()` call:

```{r add-sed-params}
parameters(aeme) <- sed_params

# Rebuild to apply the new sediment zone parameters
aeme <- build_aeme(
  aeme           = aeme,
  model          = model,
  model_controls = model_controls,
  path           = path,
  ext_elev       = 5,
  use_bgc        = TRUE
)
```

#### Inspecting sediment zones in the built model

After building, you can retrieve the number of zones and their parameters:

```{r get-sed-zones}
n_zones   <- get_glm_sed_zones(aeme = aeme, path = path)
sed_pars  <- get_glm_sed_params(aeme = aeme, path = path)
cat("Number of sediment zones:", n_zones, "\n")
sed_pars
```

#### Estimating depth-varying sediment fluxes

`estimate_zone_fluxes()` scales literature-baseline sediment fluxes to each
zone according to its mean depth and bed-area fraction (Tier 1). When observed
near-bed concentrations are available in the `aeme` object, an optional Tier 2
adjustment refines the inter-zone ratios using summer near-bottom data:

```{r zone-fluxes}
fluxes <- estimate_zone_fluxes(
  aeme      = aeme,
  path      = path,
  baseline  = c(
    fsed_oxy = -25,   # mmol O2/m2/d  (negative = into sediment)
    fsed_amm =   2,   # mmol N/m2/d
    fsed_nit =   0.2, # mmol N/m2/d
    fsed_frp =   0.05 # mmol P/m2/d
  ),
  verbose   = TRUE
)
```

The zone summary shows each zone's depth range, bed area, and estimated fluxes:

```{r zone-summary}
fluxes$zone_summary
```

#### Applying sediment fluxes to the AED configuration

`set_aed_sed_const2d()` writes the zone-specific fluxes directly into the
`aed/aed.nml` file and updates the `aeme` object:

```{r set-sed-const2d}
aeme <- set_aed_sed_const2d(
  aeme     = aeme,
  path     = path,
  baseline = c(
    fsed_oxy = -25,
    fsed_amm =  2,
    fsed_nit =  0.2,
    fsed_frp =  0.05
  )
)
```

After calling `set_aed_sed_const2d()`, you can confirm the written parameters:

```{r confirm-sed-const2d}
get_aed_sed_const2d_param(aeme = aeme, path = path) |>
  dplyr::select(name, value, index) |>
  head(20)
```

---

### Visualising the model configuration

`plot_glm_config()` generates an interactive HTML visualisation of the complete
GLM-AED setup, including the hypsograph, sediment zones, inflow/outflow
positions, active AED modules, and key parameter values. If called inside an
RStudio session, the output opens in the Viewer pane; otherwise it is saved to
a temporary HTML file.

```{r plot-config, eval=FALSE}
config_html <- plot_glm_config(aeme = aeme)
```
This html widget provides a comprehensive overview of the model configuration, 
making it easy to verify that the setup matches your intentions and to identify 
any potential issues before running the model.

You can view the HTML widget directly in RStudio's Viewer pane or open it in 
your web browser.

```{r render-html, echo=FALSE}
plot_glm_config(aeme = aeme, return_widget = TRUE)
```

---

## Running GLM-AED

```{r run-aeme}
aeme <- run_aeme(aeme = aeme)
```

---

## Visualising model output

### Temperature and stratification

`plot_output()` produces a filled contour plot for depth-varying variables
(e.g. temperature) and a time-series line plot for scalar variables:

```{r plot-temp, fig.cap="Simulated water temperature (°C) at each model layer over time."}
plot_output(aeme = aeme, var_sim = "HYD_temp")
```

### Water quality variables

Any variable that was listed in `vars_sim` and is present in the output can be
plotted the same way:

```{r plot-oxy, fig.cap="Simulated dissolved oxygen (mmol O2 m⁻³) profiles."}
plot_output(aeme = aeme, var_sim = "CHM_oxy")
```

```{r plot-tchla, fig.cap="Simulated total chlorophyll-a (µg L⁻¹) time series."}
plot_output(aeme = aeme, var_sim = "PHY_tchla")
```

---

## Assessing model performance

When observations are stored in the `aeme` object, `assess_model()` computes a
suite of skill metrics (RMSE, NSE, bias, Pearson *r*, etc.) for each simulated
variable:

```{r assess-model, eval=TRUE}
skill <- assess_model(aeme = aeme, model = model)
skill
```

---

## GLM-AED diagnostics

### Comprehensive diagnostic report

`run_glm_aed_diagnostics()` reads the model output and produces a structured
diagnostic report — a summary table and a set of grouped plots — that helps you
quickly identify unrealistic values or mass-balance issues:

```{r run-diagnostics, fig.width=9, fig.height=8}
diag <- run_glm_aed_diagnostics(
  aeme        = aeme,
  plot        = TRUE,
  print_table = TRUE
)
```

The function returns a list with three components:

- `$summary` — a data frame with min, median, mean, max, and a `flag` column
  (`"ok"` or a warning string) for each variable.
- `$plots` — a named list of `patchwork` plot objects grouped by biogeochemical
  element (`O`, `N`, `P`, `Phy`, `Sed`).
- `$data` — the tidy data frame used to produce the plots.

You can filter the diagnostics to a specific element or type:

```{r run-diagnostics-N, eval=FALSE}
# Nitrogen state variables only
diag_N <- run_glm_aed_diagnostics(
  aeme        = aeme,
  groups      = "N",
  plot        = TRUE,
  print_table = FALSE
)

# Process-rate variables only
diag_proc <- run_glm_aed_diagnostics(
  aeme        = aeme,
  groups      = "process",
  plot        = FALSE,
  print_table = TRUE
)
```

### Oxygen diagnostic page

`plot_glm_diagnostics()` provides a focused four-page diagnostic panel
specifically designed to debug anomalous dissolved oxygen behaviour — a common
challenge when coupling hydrodynamics with biogeochemistry:

```{r plot-diag, fig.width=9, fig.height=8}
pages <- plot_glm_diagnostics(aeme = aeme)

# Page 1: oxygen state and key physical drivers
print(pages$oxy)
```

```{r plot-diag-mixing, fig.width=9, fig.height=8}
# Page 2: mixing and physical structure
print(pages$physical)
```

```{r plot-diag-bod, fig.width=9, fig.height=8}
# Page 3: biological oxygen demand
print(pages$bod)
```

```{r plot-diag-sed, fig.width=9, fig.height=8}
# Page 4: sediment–water interface fluxes
print(pages$sediment)
```

---

## Working with the GLM configuration directly

The raw `glm3.nml` file and `aed/aed.nml` file can be read, modified, and
written using the NML helpers bundled with AEME:

```{r nml-helpers, eval=FALSE}
# Retrieve the parsed configuration
cfg <- read_model_config(model = "glm_aed",
                         lake_dir = get_lake_dir(aeme))

# Access GLM hydrodynamic section
glm_nml <- cfg$hydrodynamic
glm_nml$morphometry$lake_name

# Access AED biogeochemistry section
aed_nml <- cfg$bgc$aed
aed_nml$aed_nitrogen$rnitrif   # nitrification rate
```

### Retrieving parameters by module

`get_aeme_parameters()` provides a convenient way to query the AEME parameter
library for all GLM-AED parameters belonging to a particular module:

```{r get-params, eval=TRUE}
# All GLM-AED light parameters
light_params <- get_aeme_parameters(model = "glm_aed",
                                    module = "light")
light_params[, c("name", "value", "min", "max")]
```

```{r get-params-n, eval=TRUE}
# AED nitrogen module parameters
n_params <- get_aeme_parameters(model = "glm_aed",
                                module = "nitrogen")
n_params[, c("name", "value", "min", "max")]
```

---

## Summary

This vignette demonstrated the key GLM-AED-specific features available in AEME:

1. **Model description** — GLM provides variable-layer 1-D hydrodynamics; AED
   supplies modular biogeochemistry ranging from simple oxygen dynamics through
   to full nutrient–phytoplankton food-web simulation.

2. **Module selection** — `set_glm_aed_models()` lets you enable or disable
   individual AED modules to match the complexity and data requirements of your
   application.

3. **Sediment zones** — The `estimate_sed_zones()` → `glm_sed_params()` →
   `set_aed_sed_const2d()` workflow automatically derives depth-varying sediment
   parameters from the lake hypsograph and optional near-bed observations.

4. **Configuration visualisation** — `plot_glm_config()` provides an interactive
   overview of the model setup.

5. **Diagnostics** — `run_glm_aed_diagnostics()` and `plot_glm_diagnostics()`
   offer structured, automated checks for common biogeochemical issues
   (unrealistic concentrations, oxygen anomalies, excessive or negligible fluxes).

6. **Parameter access** — `get_aeme_parameters()` makes it easy to explore and
   modify individual model parameters for manual tuning or automated calibration
   (see the [aemetools](https://github.com/limnotrack/aemetools) package).
