---
title: "Hourly vs daily meteorological forcing"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)

# The worked comparison at the end needs model binaries. Detect what is
# available up front and only evaluate the run chunks accordingly, so the
# article still renders on a machine with no models. GLM is the anchor (the
# detailed flux/water-balance breakdown is GLM-only); GOTM and Simstrat are
# added to the cross-model comparison when present.
.ready <- function(f) tryCatch(nzchar(suppressMessages(f(quiet = TRUE))),
                               error = function(e) FALSE)
glm_ready      <- .ready(AEME::install_glm_aed)
gotm_ready     <- .ready(AEME::install_gotm_wet)
simstrat_ready <- .ready(AEME::install_simstrat_aed2)

models_ready <- c("glm_aed", "gotm_wet", "simstrat_aed2")[
  c(glm_ready, gotm_ready, simstrat_ready)
]
```

# Introduction

AEME accepts meteorological forcing at any regular time step. The two common
choices are **daily** (one row per day, the historical default) and **hourly**
(sub-daily, e.g. ERA5 at native resolution). This article explains what changes
when you switch from daily to hourly forcing, and — importantly — how that
change is expressed differently by each of the three hydrodynamic models
(GLM-AED, GOTM-WET, Simstrat-AED2).

The single most important thing to understand:

> **AEME never temporally disaggregates forcing.** It does not invent
> sub-daily structure from daily inputs. Sub-daily *output* therefore requires
> sub-daily *input*. What daily-vs-hourly changes is (a) how much real
> sub-daily variability the model sees, and (b) how each model fills the gaps
> between the timestamps you supplied.

To temporally disaggregate forcing, use the [metscale](https://limnotrack.com/metscale/) package, which can generate sub-daily forcing from daily inputs.

A second point that is easy to miss:

> The meteorological time step is **not** the model integration time step. Even
> a "daily" AEME run integrates internally at `time(aeme)$time_step`
> (3600 s by default) for GLM and GOTM, and at 300 s for Simstrat. Daily
> forcing does not make the physics daily — it makes the *forcing* daily.

# What the daily-vs-hourly switch controls

The switch is detected automatically by `is_subdaily()`, which looks at the
median spacing between successive timestamps in the meteo `Date` column. A
`Date` vector, or a `POSIXct` vector that is entirely at midnight, is treated as
daily; anything with a median spacing below 24 h is sub-daily.

```{r add-met, eval = FALSE}
library(AEME)

# daily_met  <- read.csv(
#   system.file("extdata/lake/data/meteo.csv", package = "AEME")
# )
hourly_met <- readr::read_csv(
  system.file("extdata/lake/data/meteo_era5_hr.csv.gz", package = "AEME")
)
daily_met <- collapse_met_daily(hourly_met, precip = "sum")

aeme_daily  <- add_met(aeme, met = daily_met)
aeme_hourly <- add_met(aeme, met = hourly_met)
```

To get sub-daily **output** as well, raise `output_time_step` (it defaults to
86400 s and must be `>= time_step`). It can be set as part of `set_time()`:

```{r set-time, eval = FALSE}
aeme_hourly <- set_time(
  aeme_hourly,
  start = "2021-07-01 00:00:00",
  stop  = "2022-07-01 00:00:00",
  spin_up = 35,
  output_time_step = 3600   # hourly output; omit for daily output from hourly forcing
)
```

or, if the run period is already set, with the generic convenience wrapper
`set_output_time_step()`, which accepts either seconds or a frequency string
(`"hourly"`, `"daily"`, `"weekly"`, `"6 hours"`, `"30 min"`, …) and applies to
every model in the ensemble:

```{r set-output-freq, eval = FALSE}
aeme_hourly <- set_output_time_step(aeme_hourly, "hourly")
aeme_6h     <- set_output_time_step(aeme_hourly, "6 hours")
```

## Pipeline steps that behave differently

| Stage | Daily forcing | Hourly forcing |
|-------|---------------|----------------|
| **Derived variables** (`expand_met()` / `standardise_met()`) | Vapour pressure, humidity, wind components, and **cloud cover** are derived at daily resolution. `calc_cc()` builds *temporary* internal hourly steps to integrate clear-sky radiation correctly, then averages cloud cover back to daily. | The same variables are derived directly at the native sub-daily resolution; `calc_cc()` uses the supplied timestamps as-is. `standardise_met()` also rescales `MET_pprain` / `MET_ppsnow` from a per-step accumulation to a mm/day rate (see below). |
| **Water balance** (`calc_water_balance()`) | Runs per day. | Sub-daily meteo is **averaged to a daily time step first** (`collapse_met_daily()`; precipitation is already a mm/day rate by this point, so a daily mean is the daily rate). The fitted outflow / inflow correction (`outf_wbal`) is therefore always a daily series, exactly like a normal inflow. |
| **Surface-temperature estimate** (`estimate_surface_temperature()`) | Daily energy-balance integration (`dt = 86400`). | Same — it consumes the daily-collapsed meteo from the water-balance step, not the raw hourly series. |
| **Model forcing files** | Written with a bare date (GLM) or `date <tab> 12:00:00 <tab> value` (GOTM), or an integer day number (Simstrat). | Written with a full timestamp split into date + time-of-day columns (GLM, GOTM) or a **fractional** day number (Simstrat). |
| **Output cadence** | One row per day. | `output_time_step / dt` rows per day; the output time axis keeps its time-of-day (`.collapse_output_date()` only drops it when every timestamp is midnight). |

Everything below the water-balance line in that table is where the **models
diverge**.

# Per-model behaviour

## GLM-AED

GLM has an explicit switch, `&meteorology/subdaily`, set by `build_glm()`:

```r
glm_nml$meteorology$subdaily <- (output_time_step < 86400) || is_subdaily(met$Date)
glm_nml$output$nsave         <- max(1L, round(output_time_step / dt_glm))  # dt_glm default 3600
```

* **Daily forcing, `subdaily = .false.`** (the default). GLM reads one met row
  per day and **synthesises the diurnal shortwave cycle itself** from the daily
  mean short-wave radiation and the solar geometry for the lake's
  latitude/longitude and day of year. Air temperature, wind, humidity, pressure
  and rainfall are supplied only as daily values (GLM interpolates between
  them), so they carry no weather-scale sub-daily structure — no gust fronts,
  morning fog burn-off, or afternoon wind maximum. A daily GLM run still has a
  day/night heating cycle, but it is a *modelled* cycle driven by clear-sky
  geometry scaled to the daily mean.
* **Hourly forcing, `subdaily = .true.`** GLM reads the meteo file at its own
  integration step and uses each value directly — no synthesis. The shortwave
  cycle, and every other variable, now carries the observed sub-daily
  variability. `bcs/meteo_glm.csv` is written with a full
  `YYYY-MM-DD HH:MM:SS` timestamp instead of a bare date.
* **Output.** `nsave` is the number of integration steps per output row:
  24 for daily output (`dt` 3600 s), 1 for hourly output.

**Net effect for GLM:** switching to hourly forcing adds *weather-scale*
variability on top of a diurnal cycle GLM already had, and — because GLM's bulk
turbulent-flux formulae are sensitive to the diurnal covariance of wind and the
air–water gradient — it can also shift the **seasonal mean**: in the worked
comparison below, GLM's mid-summer surface temperature and Schmidt stability
both move substantially. Expect changes in mixed-layer depth, the magnitude of
stratification / mixing events, near-surface temperature, and the surface energy
and water budgets.

## GOTM-WET

GOTM has **no daily/sub-daily mode**. It always integrates at
`time(aeme)$time_step` (3600 s by default) and **linearly interpolates every
input time series** (`method: 2` in `gotm.yaml`) from the timestamps in the
file to its integration step.

The consequence is that GOTM's behaviour depends entirely on what is *in* the
files:

* **Shortwave radiation is special-cased.** For daily forcing, `build_gotm()`
  (via `make_met_gotm(est_swr_hr = TRUE)`) calls `estimate_hourly_swr()` to
  disaggregate the daily-mean short-wave into an hourly, clear-sky-shaped curve
  written to `inputs/meteo_swr.dat`. So a daily GOTM run *does* get a realistic
  diurnal light cycle. For hourly forcing this step is skipped and the native
  short-wave series is written through unchanged.
* **All other variables** (air temperature, wind components, pressure,
  humidity/dewpoint, precipitation, cloud) are written to `inputs/meteo.dat` at
  whatever resolution you supplied. With daily forcing they are piecewise-linear
  ramps between daily values — no diurnal cycle in air temperature or wind.
  With hourly forcing they carry the real sub-daily signal.
* **File format.** `meteo.dat` / `meteo_swr.dat` / the inflow and outflow files
  are `date <tab> time <tab> value…`. For daily data the time column is the
  nominal `12:00:00`; for sub-daily data the real time-of-day is written. (A
  timestamp with an embedded space in the date column shifts every field and
  breaks GOTM's reader — this is why the split is done explicitly.)
* **Output.** `output.yaml` is left at its shipped daily setting unless
  `output_time_step < 86400`, in which case it is rewritten to
  `time_unit: second`, `time_step: <output_time_step>`.

**Net effect for GOTM:** the diurnal *light* cycle is present either way, but
with daily forcing the diurnal cycles of **air temperature and wind** are
absent (linear ramps). Hourly forcing adds day/night wind and air-temperature
swings, which drive night-time convective mixing and daytime near-surface
warming that a daily GOTM run damps out. This is typically a **larger**
day-to-day difference than GLM shows.

## Simstrat-AED2

Simstrat, like GOTM, has no daily/sub-daily mode. It integrates at
`Simulation/"Timestep s"` (300 s in the shipped template) and **linearly
interpolates `MeteoForcing.dat`** to that step. Unlike GOTM, there is **no
shortwave disaggregation** — `make_met_simstrat()` writes `met$MET_radswd`
exactly as supplied.

* **Daily forcing.** Every meteo column, *including short-wave radiation*, is a
  daily mean. Linear interpolation between daily means means the modelled lake
  sees an almost flat, ~24-h-smoothed radiation input: **the diurnal light
  cycle is essentially absent.** Daytime surface heating and night-time cooling
  are both strongly damped, so a daily Simstrat run tends to under-predict the
  amplitude of near-surface diurnal temperature variation and can mistime the
  onset/breakdown of daily stratification.
* **Hourly forcing.** The native diurnal cycles of radiation, wind and air
  temperature are all present; this is the configuration Simstrat is really
  designed for.
* **Time convention.** `MeteoForcing.dat` (and the inflow/outflow files) use a
  day number relative to `Simulation/"Reference year"`. Daily data lands on
  integer day numbers; hourly data lands on **fractional** day numbers
  (`date_to_simstrat_day()` returns fractional days for any sub-daily
  timestamp), which Simstrat handles natively.
* **Output.** `Output/Times` is set to `output_time_step / "Timestep s"` so
  that one output row corresponds to one `output_time_step` interval.

**Net effect for Simstrat:** daily forcing removes the diurnal radiation cycle
entirely (no synthesis fallback), so Simstrat has the most degraded inputs of
the three. In the worked comparison its summer stability is as
resolution-sensitive as GLM-AED's (~60 % lower under hourly forcing) once the
lake is tuned to stratify at all. If you care about sub-daily dynamics, or the
correct *amplitude* of daily surface-layer behaviour in Simstrat, supply hourly
forcing.

## Summary of per-model differences

| | GLM-AED | GOTM-WET | Simstrat-AED2 |
|---|---------|----------|---------------|
| Daily/sub-daily switch | `&meteorology/subdaily` (explicit) | none — interpolates the files | none — interpolates the files |
| Integration step (unchanged by met resolution) | `time_step` (def. 3600 s) | `time_step` (def. 3600 s) | `"Timestep s"` (def. 300 s) |
| Diurnal **shortwave** with *daily* forcing | synthesised from solar geometry | synthesised (`estimate_hourly_swr()`) | **none** (daily mean, interpolated) |
| Diurnal **air temp / wind** with *daily* forcing | interpolated between daily values (no weather-scale cycle) | linear ramp between days | linear ramp between days |
| Forcing-file timestamp | bare date / full `YYYY-MM-DD HH:MM:SS` | `date <tab> time` (`12:00:00` if daily) | integer / fractional day number |
| Output cadence control | `output/nsave` | `output.yaml` `time_step` | `Output/Times` |
| *Mechanistic* exposure to the switch (diurnal structure daily forcing removes) | low–moderate | moderate | high |

The last row ranks how much diurnal structure daily forcing strips from each
model's *inputs*. The **realised** difference for a given lake depends on the
lake's stratification regime and on which quantity you look at. In the
[worked comparison](#does-every-model-respond-the-same-way) below, summer
stratification weakens under hourly forcing in all three models by a broadly
similar amount (least in GOTM-WET, whose shortwave cycle is reconstructed either
way), but only GLM-AED's seasonal-mean *surface temperature* shifts materially.

# When to use hourly forcing

Use **hourly** forcing when:

* you need sub-daily output (diel oxygen, diel temperature, sub-daily mixing);
* diurnal wind and air-temperature cycles matter to your question — night-time
  convective mixing, daytime surface warming, diel stratification;
* you are running **Simstrat** and care about the amplitude or timing of
  surface-layer dynamics — its daily-forcing inputs are the most degraded, and
  the [worked comparison](#does-every-model-respond-the-same-way) bears this out:
  once the lake is tuned to stratify, its summer stability falls ~60 % under
  hourly forcing;
* your forcing product is natively sub-daily (ERA5, local AWS) — there is no
  reason to average it down.

**Daily** forcing is fine when:

* you only need daily or coarser output and seasonal-to-interannual dynamics;
* you are running GLM or GOTM, where the diurnal light cycle is reconstructed;
* forcing storage / run time is a constraint (hourly forcing is ~24× the rows
  and roughly proportionally slower to read and interpolate).

Regardless of which you pick, the **water balance** is computed daily — hourly
forcing is averaged to a daily time step before the outflow rating curve and
surface-temperature relaxation are fitted, so `wb_method` results are directly
comparable between a daily and an hourly run of the same lake.

Inflows and outflows do **not** have to match the meteo resolution. They are
boundary fluxes each model interpolates to its own integration step, so a daily
inflow series with an hourly meteo file and hourly output is fine — only
meteorology is held to the `output_time_step` cadence.

# Comparing the daily and hourly meteo

Before running anything, it is worth checking that the daily and hourly files
actually describe the *same* forcing at different resolutions.

## Precipitation: rate vs accumulation (handled automatically)

AEME defines `MET_pprain` and `MET_ppsnow` as a **rate in mm/day**. Every model
met-writer assumes that — `make_met_glm()` divides by 1000 to get m/day,
`make_met_simstrat()` by `1000 * 24` to get m/hr, `make_met_gotm()` by 86400.

Sub-daily reanalysis products (ERA5 `tp`, most AWS logs) instead report
precipitation as an **accumulation per timestep** — millimetres that fell
*during that hour*. Numerically the two are identical for daily data (mm in a
day = mm/day, which is why the daily path always "just worked"); they are
**not** identical sub-daily, where an accumulation of 0.4 mm in an hour is a
rate of 9.6 mm/day.

`standardise_met()` now resolves this automatically: when the meteo is
sub-daily it rescales `MET_pprain` / `MET_ppsnow` from a per-step accumulation
to the mm/day rate by `86400 / step_seconds`, and says so:

```
i Sub-daily meteo: rescaled "MET_pprain" and "MET_ppsnow" from a per-3600-s
  accumulation to a mm/day rate (x24).
```

`build_aeme()` calls `standardise_met()` for you, so nothing extra is needed.
If your sub-daily precipitation is genuinely already a mm/day *rate*, opt out
with `standardise_met(met, precip_accum = FALSE)`.

The check below shows the effect — the sum of the raw hourly values over a day
(≈ the true daily total) lines up with the daily file, while their mean is
~1/24 of it; the automatic `x24` rescale turns the per-hour values into that
daily-total rate:

```{r precip-check, message = FALSE}
library(dplyr)
library(ggplot2)

hourly_met <- readr::read_csv(
  system.file("extdata/lake/data/meteo_era5_hr.csv.gz", package = "AEME"),
  show_col_types = FALSE
) |>
  mutate(Date = as.POSIXct(Date, tz = "UTC"))

daily_met <- collapse_met_daily(hourly_met, precip = "sum")

hourly_daily <- hourly_met |>
  mutate(day = as.Date(Date)) |>
  group_by(day) |>
  summarise(rain_sum  = sum(MET_pprain),   # ≈ true daily total (mm)
            rain_mean = mean(MET_pprain),  # per-hour mean (mm/hr)
            .groups = "drop")

daily_met |>
  transmute(day = Date, rain_daily = MET_pprain) |>
  inner_join(hourly_daily, by = "day") |>
  summarise(
    mean_daily_source     = mean(rain_daily),
    mean_hourly_summed     = mean(rain_sum),
    mean_hourly_permin      = mean(rain_mean),
    ratio_summed_to_daily   = mean(rain_sum)  / mean(rain_daily),
    ratio_permin_to_daily   = mean(rain_mean) / mean(rain_daily)
  )
```

## Other variables: check the aggregation is faithful

For instantaneous variables (air temperature, wind, radiation, pressure,
humidity) the daily file should be close to the *mean* of the hourly file over
the same day. Radiation is the exception where it matters most — a daily-mean
short-wave of, say, 200 W m⁻² is the same energy as a 0–800 W m⁻² diurnal
cycle, but only the models that reconstruct the cycle (GLM, GOTM) will heat the
surface realistically from it.

```{r met-compare, message = FALSE, fig.height = 6}
vars <- c("MET_tmpair", "MET_radswd", "MET_wndspd", "MET_prsttn")

hourly_daily <- hourly_met |>
  mutate(day = as.Date(Date)) |>
  group_by(day) |>
  summarise(across(any_of(vars), mean), .groups = "drop") |>
  tidyr::pivot_longer(-day, names_to = "var", values_to = "hourly_mean")

daily_long <- daily_met |>
  rename(day = Date) |>
  tidyr::pivot_longer(any_of(vars), names_to = "var", values_to = "daily")

inner_join(daily_long, hourly_daily, by = c("day", "var")) |>
  filter(day >= as.Date("2021-07-01"), day <= as.Date("2021-09-01")) |>
  ggplot(aes(daily, hourly_mean)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60") +
  geom_point(size = 1.5, alpha = 1, colour = "#0065a9") +
  facet_wrap(~var, scales = "free") +
  labs(x = "Daily file", y = "Hourly file, averaged to daily",
       title = "Are the two meteo files consistent?") +
  theme_bw()
```

Points on the 1:1 line mean the two files agree once resolution is removed.
Systematic departures point to a unit or product difference that will show up
as a daily-vs-hourly model difference having nothing to do with resolution.
(Here the two files are different products — ERA5 hourly vs a daily
reanalysis — so some scatter is expected; the precipitation ratio above is the
signal to act on.)

# Worked comparison

### Isolating resolution from product

The obvious comparison — run the shipped daily file against the shipped hourly
file — confounds two things: the two files are **different reanalysis products**
(the consistency check above), so their differences are part product, part
resolution. To make this a clean *resolution* experiment, the daily forcing here
is the **hourly file aggregated to a daily step** (mean of every variable over
the day; precipitation summed to a daily total). Same numbers, same source — only
the time step differs. Every difference below is therefore the resolution.

The chunk builds and runs each available model (`r paste(models_ready, collapse = ", ")`)
twice — once with each forcing — over the 2021–22 stratification season, with
**3-hourly output** in both cases. Everything else is held fixed: same lake,
same period, same spin-up, same daily inflow (`FWMT`). (These chunks only
evaluate if a GLM binary is available — `r if (glm_ready) "it is" else "it is not, so the code is shown but not run"` here.)

### Making Simstrat stratify

With its shipped parameters Simstrat-AED2 keeps this lake — a small (15 ha),
hill-sheltered reservoir — almost completely mixed all summer (top-to-bottom
temperature difference well under 1 °C), so there is no stratification for the
daily→hourly switch to weaken and nothing to compare. Two hydrodynamic
parameters bring it into line with the other two models:

* **`a_seiche`** `0.0042 → 0.001` — the fraction of wind energy fed into the
  basin-scale internal seiche, which is Simstrat's main route for mixing the
  deep water. A sheltered lake with a short fetch dissipates less energy this
  way. This is the seiche knob to try first; on its own, though, it only nudges
  the stratification (surface stress still dominates the mixing here).
* **`f_wind`** `1.22 → 0.8` — a multiplier on the forcing wind speed. Gridded
  reanalysis wind is representative of open terrain and overstates the wind
  actually reaching a small lake ringed by bush-covered hills; scaling it down
  ~20 % is a standard, defensible correction and is what actually lets the lake
  stratify.

These are applied with [`set_simstrat_param()`] after `build_aeme()` and before
`run_aeme()`. This is **illustrative tuning to get a stratifying test case**, not
a calibration — the values were chosen so the daily-forced Simstrat run reaches a
summer stability comparable to GLM-AED and GOTM-WET, not fitted to observations.

```{r build-run, eval = glm_ready, message = FALSE, warning = FALSE}
library(AEME)
library(ggplot2)
library(dplyr)

aeme   <- readRDS(system.file("extdata/aeme.rds", package = "AEME"))
models <- models_ready

# the default controls already derive HYD_strat / HYD_thmcln; add the two
# extra stratification diagnostics this article uses
mc <- get_model_controls() |>
  set_vars_sim(c("HYD_schstb", "HYD_epidep"), simulate = TRUE)

hourly_met <- readr::read_csv(
  system.file("extdata/lake/data/meteo_era5_hr.csv.gz", package = "AEME"),
  show_col_types = FALSE
)

# daily forcing = the SAME data at a daily step. collapse_met_daily() averages
# every variable over the day; precip = "sum" because the raw ERA5 file reports
# rain/snow as a per-hour accumulation, not a mm/day rate.
daily_met <- collapse_met_daily(hourly_met, precip = "sum")

# With its shipped parameters Simstrat keeps this small, hill-sheltered lake
# essentially unstratified, which leaves nothing for the daily-vs-hourly switch
# to act on. Two hydrodynamic parameters are nudged so it stratifies like the
# other two models (see "Making Simstrat stratify", below):
#   a_seiche  0.0042 -> 0.001   less wind energy into the internal seiche
#   f_wind    1.22   -> 0.8     scale down the (over-exposed) reanalysis wind
simstrat_tuning <- list(`ModelParameters.a_seiche` = 0.001,
                        `ModelParameters.f_wind`   = 0.8)

run_one <- function(met, tag) {
  path <- file.path(tempdir(), paste0("hrdaily_", tag))
  a <- aeme |>
    add_met(met = met) |>
    set_time(start = "2020-08-01 00:00:00", stop = "2022-07-31 00:00:00",
             spin_up = 35) |>
    set_output_time_step("3 hours") |>
    build_aeme(path = path, model = models, model_controls = mc, ext_elev = 5)

  if ("simstrat_aed2" %in% models) {
    do.call(set_simstrat_param, c(
      list(file.path(get_lake_dir(a, path), "simstrat_aed2")), simstrat_tuning
    ))
  }
  run_aeme(a, model = models, path = path, model_controls = mc, verbose = FALSE)
}

runs <- list(daily = run_one(daily_met, "daily"),
             hourly = run_one(hourly_met, "hourly"))

# helper: pull one variable from both runs, for one or more models, into a
# long data frame
grab <- function(v, mods = models, ...) {
  bind_rows(lapply(names(runs), function(tag) {
    bind_rows(lapply(mods, function(m) {
      d <- tryCatch(
        get_var(runs[[tag]], model = m, var_sim = v, remove_spin_up = TRUE, ...),
        error = function(e) NULL
      )
      if (is.null(d) || !nrow(d)) return(NULL)
      transmute(d, Date, value, var = v, model = m, forcing = tag)
    }))
  }))
}
```

The walk-through below is **GLM-AED**, which is the only model that writes a
complete surface energy balance and internal water budget. The cross-model
comparison follows.

## Near-surface temperature

The headline result: GLM-AED's near-surface temperature under hourly forcing runs
**cooler** through summer here, and carries a diel wobble the daily run cannot
produce.

```{r surface-temp, eval = glm_ready, message = FALSE, warning = FALSE}
surf <- grab("HYD_temp", mods = models, depth = 0)

ggplot(surf, aes(Date, value, colour = forcing)) +
  geom_line(linewidth = 0.4) +
  scale_colour_manual(values = c(daily = "#0065a9", hourly = "#d73027")) +
  facet_wrap(~ model, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "Near-surface temperature (°C)", colour = "Forcing",
       title = "GLM-AED near-surface temperature",
       subtitle = "2021-10 to 2022-05, 3-hourly output") +
  theme_bw()
```

## Surface energy fluxes

GLM reports its surface energy balance as four daily-mean components — shortwave
`LKE_Qsw`, net longwave `LKE_Qlw`, latent (evaporative) `LKE_Qe` and sensible
`LKE_Qh` — with the convention **positive = heat into the lake**. (On a
sub-daily run these are written once per day and held constant across it; see
the note on GLM's `daily_*` diagnostics in `?read_glm_output`.)

```{r energy-fluxes, eval = glm_ready, message = FALSE, warning = FALSE}
flux <- bind_rows(lapply(c("LKE_Qsw", "LKE_Qlw", "LKE_Qe", "LKE_Qh"),
                         grab, mods = models)) |>
  mutate(day = as.Date(Date)) |>
  group_by(model, forcing, var, day) |>
  summarise(value = mean(value), .groups = "drop")

flux_lab <- c(LKE_Qsw = "Shortwave", LKE_Qlw = "Net longwave",
              LKE_Qe = "Latent", LKE_Qh = "Sensible")

ggplot(flux, aes(day, value, colour = forcing)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line(linewidth = 0.3) +
  scale_colour_manual(values = c(daily = "#0065a9", hourly = "#d73027")) +
  # facet_wrap(~ factor(flux_lab[var], flux_lab), scales = "free_y") +
  facet_grid(model ~ factor(flux_lab[var], flux_lab), scales = "free_y") +
  labs(x = NULL, y = expression("Daily-mean flux (W m"^-2*")"), colour = "Forcing",
       title = "Surface energy balance components") +
  theme_bw()
```

The shortwave and longwave terms barely move — GLM synthesises the shortwave
diurnal cycle from solar geometry either way, so resolving it in the forcing
adds little. The **turbulent** fluxes are where resolution bites. `LKE_Qe` and
`LKE_Qh` are bulk-aerodynamic terms, roughly proportional to
*wind speed* × (*surface* − *air*) gradient; both factors swing diurnally and,
critically, **co-vary** — the strongest winds and the largest air–water
contrasts do not occur at the same clock hour. A daily mean throws that
covariance away. Here the effect is a large redistribution between the two:

```{r flux-means, eval = glm_ready, message = FALSE, warning = FALSE}
flux_means <- flux |>
  group_by(model, forcing, var) |>
  summarise(mean_wm2 = mean(value), .groups = "drop")

# per-component means plus the net (sum of the four components)
net_means <- flux_means |>
  group_by(model, forcing) |>
  summarise(var = "Net (sum)", mean_wm2 = sum(mean_wm2), .groups = "drop")

bind_rows(mutate(flux_means, var = unname(flux_lab[var])), net_means) |>
  mutate(mean_wm2 = round(mean_wm2, 1)) |>
  tidyr::pivot_wider(names_from = forcing, values_from = mean_wm2) |>
  rename(component = var)
```

The daily-forced run gains more net heat (or loses less) over the season, which
is consistent with its warmer surface layer in the first plot.

## Water balance

AEME's **fitted** water balance (`calc_water_balance()`, the `outf_wbal`
correction) is always computed on a daily-collapsed copy of the meteo, so it is
identical between the two runs by construction. What *does* differ is the
water budget the model closes internally each step — and its one
meteorology-driven term is **evaporation**, which follows the latent heat flux
above.

```{r water-balance, eval = glm_ready, message = FALSE, warning = FALSE}
wb <- bind_rows(lapply(c("LKE_evpvol", "LKE_pcpvol", "LKE_inflow", "LKE_outflow"),
                       grab, mods = models)) |>
  mutate(day = as.Date(Date)) |>
  group_by(model, forcing, var, day) |>
  summarise(value = mean(value), .groups = "drop") |>
  arrange(day) |>
  group_by(model, forcing, var) |>
  mutate(cumulative = cumsum(value)) |>
  ungroup()

wb_lab <- c(LKE_evpvol = "Evaporation", LKE_pcpvol = "Precipitation",
            LKE_inflow = "Inflow", LKE_outflow = "Outflow")

ggplot(wb, aes(day, cumulative, colour = forcing)) +
  geom_line(linewidth = 0.4) +
  scale_colour_manual(values = c(daily = "#0065a9", hourly = "#d73027")) +
  # facet_wrap(~ factor(wb_lab[var], wb_lab), scales = "free_y") +
  facet_grid(model ~ factor(wb_lab[var], wb_lab), scales = "free_y") +
  labs(x = NULL, y = expression("Cumulative volume (m"^3*")"), colour = "Forcing",
       title = "Cumulative water-balance fluxes") +
  theme_bw()
```

Inflow is a boundary series — untouched — so its two lines lie exactly on top of
each other, and precipitation matches too (the daily forcing is the summed
hourly total). Outflow tracks inflow closely. **Evaporative loss** is markedly
lower under hourly forcing — the diurnal wind/humidity covariance again — which
leaves the hourly run slightly fuller.

```{r lake-level, eval = glm_ready, message = FALSE, warning = FALSE}
grab("LKE_lvlwtr", mods = models) |>
  ggplot(aes(Date, value, colour = forcing)) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~ model, scales = "free_y", ncol = 1) +
  scale_colour_manual(values = c(daily = "#0065a9", hourly = "#d73027")) +
  labs(x = NULL, y = "Lake surface (m)", colour = "Forcing",
       title = "Modelled lake level") +
  theme_bw()
```

## Stratification

The cooler, more variable surface forcing under hourly meteo — in particular the
**night-time** wind and convective cooling that a daily mean cannot represent —
mixes the water column harder. Schmidt stability (the work needed to mix the
lake to uniform density) is the clearest summary:

```{r schmidt, eval = glm_ready, message = FALSE, warning = FALSE}
grab("HYD_schstb", mods = models) |>
  ggplot(aes(Date, value, colour = forcing)) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~ model, scales = "free_y", ncol = 1) +
  scale_colour_manual(values = c(daily = "#0065a9", hourly = "#d73027")) +
  labs(x = NULL, y = expression("Schmidt stability (J m"^-2*")"), colour = "Forcing",
       title = "GLM-AED water-column stability") +
  theme_bw()
```

The hourly GLM run is substantially **less stable** through summer, its
thermocline sits **shallower**, and it dips in and out of a fully mixed state on
days the daily run stays stratified:

```{r strat-metrics, eval = glm_ready, message = FALSE, warning = FALSE}
strat <- bind_rows(lapply(c("HYD_thmcln", "HYD_epidep", "HYD_schstb"),
                          grab, mods = models),
                   grab("HYD_temp", mods = models, depth = 0) |>
                     mutate(var = "T_surface"),
                   grab("HYD_temp", mods = models, depth = 0.5,
                        depth_ref = "bottom") |> mutate(var = "T_bottom"))

# surface-minus-bottom temperature difference
dT <- strat |>
  filter(var %in% c("T_surface", "T_bottom")) |>
  tidyr::pivot_wider(names_from = var, values_from = value) |>
  mutate(value = T_surface - T_bottom, var = "dT_surf_bot")

bind_rows(strat, dT) |>
  filter(var %in% c("HYD_schstb", "HYD_thmcln", "dT_surf_bot")) |>
  mutate(var = recode(var,
                      HYD_schstb  = "Schmidt stability (J/m2)",
                      HYD_thmcln  = "Thermocline depth (m)",
                      dT_surf_bot = "Surface - bottom (deg C)")) |>
  group_by(model, forcing, var) |>
  summarise(summer_mean = round(mean(value[format(Date, "%m") %in%
                                             c("12", "01", "02")], na.rm = TRUE), 2),
            .groups = "drop") |>
  tidyr::pivot_wider(names_from = forcing, values_from = summer_mean)
```

A diel composite — every value binned by hour of day over mid-summer — shows
where the daily run is blind. Its stability curve is nearly flat (only the
synthesised shortwave cycle drives it); the hourly run has a real daytime build
-up and night-time erosion of stratification:

```{r diel, eval = glm_ready, message = FALSE, warning = FALSE}
grab("HYD_schstb", mods = models) |>
  filter(format(Date, "%m") %in% c("12", "01", "02")) |>
  mutate(hour = as.integer(format(Date, "%H"))) |>
  group_by(model, forcing, hour) |>
  summarise(value = mean(value), .groups = "drop") |>
  ggplot(aes(hour, value, colour = forcing)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ model, scales = "free_y", ncol = 1) +
  scale_colour_manual(values = c(daily = "#0065a9", hourly = "#d73027")) +
  scale_x_continuous(breaks = seq(0, 21, 3)) +
  labs(x = "Hour of day (UTC)", y = expression("Schmidt stability (J m"^-2*")"),
       colour = "Forcing", title = "Diel cycle of water-column stability",
       subtitle = "Mid-summer mean by hour of day") +
  theme_bw()
```

## Does every model respond the same way?

Everything above is GLM-AED. GOTM-WET and Simstrat-AED2 have no explicit
sub-daily switch — they linearly interpolate whatever is in the forcing files to
their integration step — and, as the *Per-model behaviour* section explains,
daily forcing strips the diurnal cycle from their air-temperature and wind
inputs (and, for Simstrat, from short-wave as well). Running the same
daily-vs-hourly experiment through all three (Simstrat with the stratification
tuning above) gives:

```{r multi-model, eval = glm_ready, message = FALSE, warning = FALSE}
mm <- bind_rows(
  grab("HYD_temp", depth = 0) |> mutate(metric = "Near-surface temp (°C)"),
  grab("HYD_schstb")          |> mutate(metric = "Schmidt stability (J m⁻²)")
) |>
  mutate(model = toupper(sub("_.*", "", model)))

ggplot(mm, aes(Date, value, colour = forcing)) +
  geom_line(linewidth = 0.3) +
  scale_colour_manual(values = c(daily = "#0065a9", hourly = "#d73027")) +
  facet_grid(metric ~ model, scales = "free_y", switch = "y") +
  labs(x = NULL, y = NULL, colour = "Forcing",
       title = "Daily vs hourly forcing, three models",
       subtitle = "Same lake, same aggregated-to-daily forcing") +
  theme_bw() +
  theme(strip.placement = "outside")
```

```{r multi-model-table, eval = glm_ready, message = FALSE, warning = FALSE}
dT <- bind_rows(grab("HYD_temp", depth = 0) |> mutate(pos = "s"),
                grab("HYD_temp", depth = 0.5, depth_ref = "bottom") |>
                  mutate(pos = "b")) |>
  tidyr::pivot_wider(names_from = pos, values_from = value) |>
  mutate(value = s - b, metric = "Surface - bottom (°C)")

bind_rows(
  grab("HYD_temp", depth = 0) |> mutate(metric = "Near-surface temp (°C)"),
  grab("HYD_schstb")          |> mutate(metric = "Schmidt stability (J/m²)"),
  dT
) |>
  filter(format(Date, "%m") %in% c("12", "01", "02")) |>   # mid-summer
  group_by(model, metric, forcing) |>
  summarise(v = mean(value, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = forcing, values_from = v) |>
  mutate(`hourly - daily` = round(hourly - daily, 2),
         daily = round(daily, 2), hourly = round(hourly, 2)) |>
  arrange(metric, model)
```

Two things stand out.

**Stratification weakens under hourly forcing in every model**, and by a similar
amount: mid-summer Schmidt stability falls ~55 % in GLM-AED, ~60 % in Simstrat,
and ~30 % in GOTM-WET, with the top-to-bottom temperature difference dropping in
step. The night-time wind and convective cooling that a daily mean cannot
represent mixes the surface layer down in all three. GOTM moves least because
`build_gotm()` already disaggregates the daily-mean shortwave into an hourly
clear-sky curve (`estimate_hourly_swr()`), so its daily run keeps the dominant
surface-heating cycle; only the air-temperature and wind swings are missing.

**Seasonal-mean surface temperature is a different story** — only GLM-AED's
moves (down ~3 °C). GOTM and Simstrat re-arrange heat *within* the water column
when the forcing is resolved, but hold about the same total: their surface means
shift by ~0.1 °C. GLM-AED's bulk turbulent-flux formulae respond to the diurnal
covariance of wind and the air–water gradient (the energy-flux section above),
so for GLM the switch also changes the *net* surface heat flux, not just its
vertical distribution.

So the direction is universal — hourly forcing never *increases* summer
stability — but which quantities carry the signal, and how big it is, is
model-specific.

## What this tells you

For this lake, over one stratification season, moving from daily to hourly
forcing of the *same* meteorological data:

* **left the radiative fluxes essentially unchanged** — GLM (and GOTM) already
  reconstruct the shortwave diurnal cycle, so resolving it in the input adds
  little;
* **redistributed GLM's turbulent fluxes** — resolving the diurnal covariance of
  wind and the air–water gradient changed the latent/sensible split and lowered
  net heat gain, cooling the surface layer;
* **cut GLM's modelled evaporative water loss** in step with the latent flux,
  nudging lake level up (inflow/outflow, being boundary series, did not move);
* **weakened stratification in every model** — Schmidt stability down 30–60 %, a
  shallower thermocline, and occasional full mixing on days the daily run stayed
  stratified — but changed the seasonal-mean *surface temperature* only in GLM
  (GOTM and Simstrat redistribute heat vertically without changing the total).

The direction and size of these differences are lake- and season-specific, and —
as the three-model comparison shows — model-specific. Note also that Simstrat had
two parameters adjusted just to give it a stratified water column to compare;
different tuning would change its numbers. Treat this as a worked template, not a
universal result, and always derive your daily and sub-daily forcing from the
*same* source so the comparison isolates resolution.

To keep sub-daily *output*, leave `output_time_step` below 86400 as above; set it
to `86400` (or call `set_output_time_step(aeme, "daily")`) if you only want the
sub-daily forcing but daily output.

# Gotchas

* **Forcing must span the run.** `check_time()` requires the meteo to cover
  `start - spin_up` … `stop`. Hourly ERA5 often starts at 13:00 UTC on the
  first day; make sure your `start` is within the data.
* **No meteo disaggregation.** Setting `output_time_step = 3600` with daily
  *meteo* does *not* give you meaningful hourly output — `check_time()` will
  stop you — and even where it ran, the sub-daily signal would be synthetic
  (shortwave) or a linear ramp (everything else). Sub-daily output needs
  sub-daily meteo; inflows and outflows may stay daily.
* **Precipitation is auto-rescaled.** Sub-daily `MET_pprain` / `MET_ppsnow` are
  taken to be per-step accumulations and multiplied to a mm/day rate by
  `standardise_met()`. If yours are already a rate, pass
  `standardise_met(precip_accum = FALSE)` (and note `build_aeme()` always uses
  the default).
* **Water balance is always daily.** Don't expect `water_balance(aeme)$data`
  or `outf_wbal.dat` to be hourly even when the run is — this is deliberate.
* **Cloud cover from short-wave.** If you supply hourly short-wave but the
  derived cloud-cover series looks wrong, check that short-wave is in W m⁻²
  and that night-time values are ~0 — `calc_cc()` interpolates across gaps but
  cannot recover from a unit error.
* **Run cost.** Hourly forcing is ~24× the input rows. GOTM and Simstrat pay
  this at every integration step (interpolation); GLM pays it once per step
  when `subdaily = .true.`. The three-model worked comparison below therefore
  takes noticeably longer than the GLM-only chunks — Simstrat, at a 300 s
  internal step, dominates.

# See also

* `vignette("aeme-inputs")` — the full meteo input schema and
  `standardise_met()`.
* `vignette("rotoehu-water-balance")` — how the (daily) water balance and
  surface-temperature estimate work.
* `?set_time` / `?set_output_time_step` — `time_step` and `output_time_step`.
