---
title: "Initial conditions, uncertainty and spin-up"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

<!--
DRAFT -- written to demonstrate the AEME 0.4.0 initial-condition / spin-up
toolset (`set_initial_conditions()`, `get_initial_conditions()`,
`update_init()`, `assess_spin_up()`, `plot_spin_up()`), which is not shown in
any other vignette or article. Registered under a new "Uncertainty" articles
group in `_pkgdown.yml`.

TODO before publishing:
  - Run every code chunk end-to-end on a machine with R and the GLM-AED binary
    (`AEME::install_glm_aed()`). The narrative around the printed tables and the
    recommended spin-up lengths (sections "Reading the result", "A
    biogeochemical variable needs longer") is written from the documented
    behaviour of `assess_spin_up()`, not from a confirmed run of this article --
    reconcile the prose with the actual numbers once knit, in particular the
    claim that `CHM_oxy` needs a longer spin-up than `HYD_temp` for this lake.
  - The `HYD_temp` grid stops at 180 d and the `CHM_oxy` grid at 365 d; the
    example lake ships meteo from 2019-01-01 (~577 d before `time$start`), so
    both fit. If either grid is extended, check `check_time()` still passes.
  - Consider trimming the `CHM_oxy` ensemble/grid if knit time is a problem
    (it is the larger of the two: 5 spin-up lengths x 3 members = 15 builds).
-->

# Why the initial state matters

Every lake simulation has to start somewhere. Before the first timestep, each
model needs a complete initial state: a water level, a temperature and salinity
profile, and -- with biogeochemistry switched on -- a starting value for every
water-column pool (oxygen, nutrients, phytoplankton, ...). That state is almost
always a guess. It might come from a single CTD cast, a climatological profile,
or nothing better than "10 &deg;C, top to bottom".

**Spin-up** is the standard fix: run the model for a lead-in period *before* the
window you actually care about, and discard that lead-in from the output. Given
enough lead-in, the physics forgets the initial guess -- surface heat exchange,
mixing and inflow/outflow replace the guessed state with one that is consistent
with the forcing.

How much lead-in is "enough" depends on the variable:

- the **surface mixed layer** re-equilibrates with the atmosphere in days;
- the **hypolimnion** is only ventilated when the lake mixes, so a wrong deep
  temperature can survive most of a stratified season;
- **biogeochemical pools** turn over on the timescale of their sources and sinks
  -- weeks to years.

Too short a spin-up leaves an initial-condition bias in the results that is easy
to mistake for model error. Too long a spin-up wastes forcing data and compute,
and AEME will not run it: `check_time()` requires meteorological (and inflow)
forcing to span `start - spin_up` through `stop`.

This article shows how to set the initial state with the AEME 0.4.0 functions,
how to treat the initial guess as one axis of ensemble uncertainty, and how to
use `assess_spin_up()` to choose a spin-up length that is long enough and no
longer.

```{r setup, message = FALSE, warning = FALSE}
library(AEME)
library(ggplot2)
library(patchwork)

options(AEME.inform = FALSE)
```

We use the example lake shipped with the package. Copy its configuration
directory somewhere writable and load it:

```{r build, message = FALSE, results = "hide"}
tmpdir <- tempdir()
aeme_dir <- system.file("extdata/lake/", package = "AEME")
file.copy(aeme_dir, tmpdir, recursive = TRUE)
path <- file.path(tmpdir, "lake")

aeme <- yaml_to_aeme(path = path, "aeme.yaml")
```

This lake runs from 2020-08-01 to 2021-06-30 and ships daily meteorology back to
2019-01-01, so there is roughly a year and a half of forcing lead available for
spin-up, plus a lake observation file we can seed the initial profile from.

# Setting the initial state

`set_initial_conditions()` is the single entry point for the initial state.
The generic arguments -- `depth`, `profile`, `wq` -- apply to every model;
`model_init` overrides them for one model.

```{r set-ic-explicit}
prof <- data.frame(
  depth       = c(0, 5, 10, 13),
  temperature = c(12, 11, 9, 8),
  salt        = 0
)
aeme <- set_initial_conditions(aeme, depth = 13, profile = prof)
```

`profile` is a data.frame with a `depth` column (positive-down, 0 = surface) and
a `temperature` column in &deg;C; `salt` (ppt) is optional and defaults to 0.
`depth` is the initial water level, i.e. the height of the surface above the
lowest point of the hypsograph.

`wq` sets water-column biogeochemical initials, keyed by AEME variable name --
either a single value (constant with depth) or a `depth`/`value` data.frame:

```{r set-ic-wq}
aeme <- set_initial_conditions(aeme, wq = list(CHM_oxy = 300))
```

`model_init` takes per-model overrides that are merged over the generic
specification when that model is built -- for example a GLM-AED-specific
temperature profile:

```{r set-ic-model-init}
aeme <- set_initial_conditions(
  aeme,
  model_init = list(
    glm_aed = list(profile = data.frame(depth = c(0, 15),
                                        temperature = c(12, 8)))
  )
)
```

The whole specification is stored in `configuration(aeme)$initial_conditions` as
a list with a `default` entry plus one entry per model. It survives a
`load_configuration()` (and therefore the reload that happens at the end of
`build_aeme()`). `get_initial_conditions()` reads it back -- the full spec, or
one model's *resolved* conditions with its overrides merged in:

```{r get-ic}
get_initial_conditions(aeme)                       # full specification
get_initial_conditions(aeme, model = "glm_aed")    # resolved for GLM-AED
```

## Seeding from observations

If the lake has observations, `from_obs = TRUE` seeds the generic profile and
the water-quality initials from them (via `update_init()`) before any explicit
arguments are applied -- so explicit arguments still win. `update_init()` takes a
three-month window centred on the start month, fits temperature and salinity
against depth, and uses the window median for each water-quality pool.

```{r set-ic-from-obs}
aeme <- set_initial_conditions(aeme, from_obs = TRUE)
get_initial_conditions(aeme, model = "glm_aed")$profile
```

Two things happen automatically regardless of what you set here. For backwards
compatibility the generic `depth`/`profile` are also written to
`input(aeme)$init_depth` / `input(aeme)$init_profile`. And when the lake has an
observed water level at `time$start`, `build_aeme()` overrides the initial water
level with that observation.

# Initial conditions as a source of uncertainty

It helps to think of the initial guess as one axis of an ensemble, alongside the
forcing and the parameters. If you perturbed only the initial state -- ran the
same model, same forcing, same parameters, but from a range of plausible
starting profiles -- the runs would disagree at first and then converge as
spin-up dissipates the differences.

"How long a spin-up do I need?" is then a concrete question:

> How long until the spread that initial-condition uncertainty injects into the
> reported variables has collapsed below the accuracy I care about?

`assess_spin_up()` answers it by building exactly that ensemble.

# `assess_spin_up()` end to end

`assess_spin_up()` builds and runs one simulation per *(spin-up length,
ensemble member)* combination, each in its own subdirectory of `path`, with
written output restricted to `vars`. It then extracts the modelled state at the
first timestep of the analysis period (the first timestep after spin-up) and
summarises how the members agree.

```{r assess-temp, message = FALSE}
res_temp <- assess_spin_up(
  aeme,
  model     = "glm_aed",
  spin_up   = c(0, 30, 90, 120, 150, 180),
  perturb   = list(temperature = c(-4, 0, 4)),
  vars      = "HYD_temp",
  metric    = "spread",
  tolerance = 0.5,
  build_args = list(ext_elev = 5),
  path      = file.path(tmpdir, "spinup_temp")
)
res_temp
```

`perturb` is a named list that defines the ensemble:

- `"temperature"` / `"salt"` -- **additive** offsets applied to that column of
  the baseline initial profile (here: baseline &minus;4 &deg;C, baseline, and
  baseline &plus;4 &deg;C);
- any `model_controls$var_aeme` name (e.g. `CHM_oxy`) -- a **multiplicative**
  factor on that water-quality initial value.

The vectors are recycled to a common length, which becomes the ensemble size;
member *i* takes element *i* of every vector. Here that is 4 spin-up lengths
&times; 3 members = 12 builds and runs. The baseline is
`get_initial_conditions(aeme, model = "glm_aed")`, falling back to
`input(aeme)$init_profile` / `model_controls$initial_wc`, so run
`set_initial_conditions()` (or `update_init()`) first.

# Reading the result

The return value is a list of class `aeme_spin_up`:

- **`$data`** -- long data.frame (`spin_up`, `member`, `var`, `depth`, `value`)
  of the state at the start of the analysis period.
- **`$summary`** -- one row per `spin_up` &times; `var`, each column a mean over
  depth of a per-depth quantity:
  - `spread` -- standard deviation across members, in the variable's units;
  - `spread_cv` -- that SD divided by the ensemble mean (dimensionless);
  - `drift` -- distance of this spin-up's ensemble mean from the
    *longest* spin-up's ensemble mean, in the variable's units;
  - `drift_cv` -- that distance divided by the longest spin-up's mean.
- **`$recommended`** -- the shortest spin-up whose `metric` meets `tolerance`,
  per variable, plus `overall` (the largest of those). `NULL` when `tolerance`
  is `NULL`.
- **`$failures`** -- any *(spin-up, member)* runs that errored.

```{r summary-temp}
res_temp$summary
res_temp$recommended
```

`plot_spin_up()` draws the chosen metric against spin-up length, with a dashed
line at the tolerance:

```{r plot-temp}
plot_spin_up(res_temp)
```

## Spread versus drift

`spread` and `drift` answer different questions:

- **`spread`** -- have the members converged *to each other*? A small spread
  means the choice of initial profile no longer matters.
- **`drift`** -- has the ensemble converged to the *long-spin-up* answer? A small
  drift means the spin-up is long enough that extending it further would not move
  the result.

You want both small. A shallow temperature perturbation can collapse in spread
within 30 days while the ensemble mean still carries a deep cold bias -- visible
as drift -- out to 180 days. Plot them side by side:

```{r plot-temp-both, fig.width = 9, fig.height = 4}
wrap_plots(
  plot_spin_up(res_temp, metric = "spread"),
  plot_spin_up(res_temp, metric = "drift"),
  nrow = 1
)
```

# A biogeochemical variable needs longer

Repeat the assessment for dissolved oxygen, with biogeochemistry switched on.
Oxygen is perturbed multiplicatively (half, baseline, double), and because
concentrations are strictly positive and on a completely different scale from
temperature we compare against `spread_cv` -- the coefficient of variation --
with a 5&nbsp;% tolerance.

```{r assess-oxy, message = FALSE}
mc <- get_model_controls(use_bgc = TRUE)


res_bgc <- assess_spin_up(
  aeme,
  model          = "glm_aed",
  spin_up        = c(0, 30, 90, 120, 150, 180, 365),
  perturb        = list(CHM_oxy = c(0.5, 1, 2),
                        PHS_frp = c(0.5, 1, 2),
                        NIT_nit = c(0.5, 1, 2),
                        NIT_amm = c(0.5, 1, 2)
                        ),
  vars           = c("CHM_oxy", "PHS_frp", "NIT_nit", "NIT_amm"),
  metric         = "spread_cv",
  tolerance      = 0.05,
  model_controls = mc,
  build_args     = list(use_bgc = TRUE, ext_elev = 5),
  path           = file.path(tmpdir, "spinup_oxy")
)
res_bgc$summary
res_bgc$recommended
```

```{r plot-oxy}
plot_spin_up(res_bgc)
```

Compare `res_bgc$recommended` with `res_temp$recommended`. For this lake the
oxygen pool takes longer to lose the memory of its starting value than the
temperature profile does, because it is set by the balance of primary
production, respiration and sediment demand rather than by mixing alone. The
spin-up that is adequate for a physics-only study is not necessarily adequate
once biogeochemistry is reported.

A note on the `*_cv` metrics: they are dimensionless and therefore comparable
across variables of different magnitude, which is what makes them the right
choice for water-quality concentrations. They are undefined where the ensemble
mean crosses zero, and those depths are dropped from the depth-average, so keep
`spread` / `drift` (in physical units) for variables that can be near zero.

# Choosing the assessment parameters

| Argument | Guidance |
|---|---|
| `vars` | The variables your study actually reports. The slowest-relaxing one -- usually a deep or biogeochemical variable -- sets the spin-up requirement, not surface temperature. |
| `perturb` | Size the offsets/factors to your *real* initial-condition uncertainty: roughly &pm;3--4 &deg;C for a profile from a single cast or climatology; a factor of 0.5--2 for a nutrient or oxygen pool. |
| `spin_up` | Start coarse (`c(0, 30, 90, 180, 365)`), then add points around the knee of the curve if you need a sharper estimate. |
| `metric` + `tolerance` | Use `spread` / `drift` with a tolerance in physical units (e.g. `0.5` &deg;C) when you have a target accuracy; use `spread_cv` / `drift_cv` with a fractional tolerance (e.g. `0.05`) to compare across variables. |
| `model` | One model per call. Different engines mix at different rates, so assess each model you intend to run. |

Every *(spin-up, member)* cell is a full `build_aeme()` + `run_aeme()`, so cost
is `length(spin_up) * n_members` runs and one subdirectory of `path` per cell.
Restricting `vars` keeps the written output small; drop `spin_up` points or
ensemble members if a run is too slow.

# Applying the result

Set the spin-up you chose on the object with `set_time()`. `spin_up` is in
**days** -- a single number for all models, or a named list per model -- and is
stored at `time(aeme)$spin_up`:

```{r apply, eval = FALSE}
tm <- time(aeme)
aeme <- set_time(aeme,
                 start   = format(tm$start, "%Y-%m-%d"),
                 stop    = format(tm$stop, "%Y-%m-%d"),
                 spin_up = res_temp$recommended[["overall"]])
```

From then on the spin-up window is simulated but stripped from everything you
read back: `get_var()`, `plot_ts()`, `plot_output()` and the rest default to
`remove_spin_up = TRUE`. Just make sure the forcing still covers
`start - spin_up` through `stop`, or `build_aeme()` will stop at `check_time()`.

# Scope

`assess_spin_up()` perturbs only the initial state. It is not a full uncertainty
or sensitivity analysis -- for forcing and parameter sensitivity see
`vignette("testing-parameters")` and the `aemetools` package
(`run_aeme_param()`, `sa_aeme()`, `calib_aeme()`).
