---
title: "Using LERNZmp with AEME"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
lernzmp_exam <- system.file("extdata/lernzmp", package = "AEME")
dir.create("lernzmp")
file.copy(list.files(lernzmp_exam, full.names = TRUE), "lernzmp")


```

```{r setup}
library(AEME)
```

# Introduction

The [Lake Ecosystem Research New Zealand Model Platform
(LERNZmp)](https://limnotrack.shinyapps.io/LERNZmp/) is a web platform that
provides a user-friendly interface to the lake ecosystem model output for New
Zealand lakes. The platform is designed to provide a simple way to explore the
model output across multiple lakes and also to compare the model results with
observed data.

The platform is located here:
[LERNZmp](https://limnotrack.shinyapps.io/LERNZmp/). Select and load a lake
model output on the "Overview" tab. The model output can be downloaded in the
"Download Models" tab. This downloads a ".zip" folder containing the model
output for the selected lake(s) as ".rds" files and a lake metadata file
"LERNZmp_lake_metadata.csv".

# Using LERNZmp output with AEME

Once you have downloaded the LERNZmp model output and unzipped the folder, you
should a similar file structure to the following:

```{r}
#| label: lernzmp-file-structure
list.files("lernzmp")
```

## LERNZmp metadata

The metadata file contains information about the all the lakes in the LERNZmp 
platform. This includes the lake ID, name surface area (ha), region, geomorphic 
type, depth, depth measurement (measured or predicted), data (no data/minimal 
data/limited and irregular/periodic but sparse/seasonal but detailed) and 
lernzmp file name.

```{r}
#| label: lernzmp-metadata
metadata <- read.csv("lernzmp/LERNZmp_lake_metadata.csv")
head(metadata)
```

The lakes included in this example have ID's LID11133 and LID40102. We will 
filter the metadata to examine these two lakes.

```{r}
#| label: lernzmp-metadata-filter

metadata <- metadata |> 
  dplyr::filter(aeme_file %in% c("LID11133", "LID40102"))
metadata
```

These are lakes Rotorua (LID11133) and Rotoma (LID40102), respectively.

# Build AEME models

We will now build AEME models for these two lakes using the LERNZmp model 
output. We will first load the AEME object from the ".rds" files.

The `Aeme` object contains the lake metadata, model output, and model controls.
More details can be found `vignette("intro-aeme")`. It is an S4 object of the
class `Aeme`.

```{r}
#| label: lernzmp-load-aeme
aeme <- readRDS("lernzmp/LID11133.rds")
class(aeme)
```

It can be printed to the console to see the contents of the object.

```{r}
#| label: lernzmp-print-aeme
aeme
```

This allows for quick inspection of all the different slots within the `Aeme`
object. The lake section has the lake metadata, the time section has the start, 
stop and spin-up dates, the configuration section has the model configuration 
which allows for the building of the AEME models locally.

```{r}
#| label: lernzmp-build-aeme
#| echo: false

model <- c("glm_aed", "gotm_wet") # models to build
path <- "aeme" # directory in which the model configuration will be built

aeme <- build_aeme(aeme = aeme, model = model, path = path,
                    use_aeme = TRUE, use_bgc = TRUE)
```

```{r}
#| label: lernzmp-aeme-list-files

list.files("aeme", recursive = TRUE)
```

# Run AEME models

We will now run the AEME models for the two lakes. This will run the models
with the configurations built in the `path` directory. The `parallel` argument
is set to `TRUE` to run the models in parallel which can speed up the process.

```{r}
#| label: lernzmp-run-aeme
aeme <- run_aeme(aeme = aeme, model = model, path = path, parallel = TRUE)
aeme
```

In the "Output" section of the `Aeme` object, the "Number of ensembles" is set to
1 indicating that there is now output for each model in the `Aeme` object.


```{r}
#| label: lernzmp-plot-temperature
plot_output(aeme = aeme, model = model, var_sim = "HYD_temp")
```


```{r}
#| label: lernzmp-plot-oxygen
plot_output(aeme = aeme, model = model, var_sim = "CHM_oxy")
```


