---
title: "Reservoir Simulation with Multiple Outlets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reservoir Simulation with Multiple Outlets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

## Introduction

Reservoirs differ from natural lakes in several ways that are important for
hydrodynamic modelling:

1. **Regulated water levels** – reservoir levels are managed through controlled
   releases rather than a natural outlet or seepage.
2. **Multiple outlets at different depths** – a single reservoir can have:
   - A **surface spillway** that only activates when the reservoir is near full
     capacity.
   - A **penstock or bottom outlet** for daily operational releases.
   - A **mid-level selective withdrawal outlet** for water quality management
     (e.g. to avoid releasing cold, anoxic hypolimnetic water).
3. **Thermal stratification management** – withdrawals from different depth layers
   affect the downstream temperature and water quality.

AEME supports all three major 1-D hydrodynamic models
(DYRESM-CAEDYM, GLM-AED, GOTM-WET) with multiple outlets at arbitrary
elevations. This vignette walks through setting up a reservoir ensemble model
with two outlets at different levels: a regulated **penstock** at depth and a
**surface spillway**.

## Setup

```{r setup, message=FALSE}
library(AEME)
library(ggplot2)
library(dplyr)
```

## Reservoir data

We will use a fictional reservoir, Reservoir, located in the North
Island of New Zealand. The reservoir has a full supply level of 150 m above sea
level and a maximum depth of 25 m.

```{r reservoir-info}
lat   <- -37.80
lon   <- 176.00
elev  <- 150   # full supply level (m a.s.l.)
depth <- 25    # maximum depth (m)
area  <- 1.2e6 # surface area at full supply level (m²)

reservoir <- list(
  name      = "Reservoir",
  id        = "res001",
  latitude  = lat,
  longitude = lon,
  elevation = elev,
  depth     = depth,
  area      = area
)
```

## Simulation period

```{r time-setup}
time <- list(
  start = "2020-08-01 00:00:00",
  stop  = "2021-06-30 00:00:00"
)
```

## Meteorological data

Meteorological data bundled with the AEME package (originally from Lake Wainamu,
North Island) is reused here to drive the reservoir simulation.

```{r meteo}
meteo_file <- system.file("extdata/lake/data/meteo.csv", package = "AEME")
met <- read.csv(meteo_file) |>
  mutate(Date = as.Date(Date))

head(met[, 1:5])
```

## Hypsograph

The hypsograph describes the relationship between elevation and surface area
(and thus volume). Constructed reservoirs typically have a more prismatic
(box-like) shape compared to natural lakes. This is reflected by a higher
`volume_development` parameter (values > 1.5 give a convex, cylindrical shape;
a value of 3 corresponds to a perfect cylinder).

```{r hypsograph}
hypsograph <- generate_hypsograph(
  max_depth          = depth,
  surface_area       = area,
  volume_development = 2.5,   # more prismatic than a natural lake
  elev               = elev,
  ext_elev           = 2      # extend 2 m above full supply level
)

head(hypsograph)
```

```{r plot-hypsograph}
ggplot(hypsograph, aes(x = area / 1e6, y = elev)) +
  geom_line(linewidth = 1, colour = "#0065a9") +
  geom_point(size = 1.5, colour = "#0065a9") +
  labs(
    x     = expression("Area (km"^2 * ")"),
    y     = "Elevation (m a.s.l.)",
    title = "Reservoir – hypsograph"
  ) +
  theme_bw()
```

## Inflow data

We create a synthetic seasonal inflow time series typical of a New Zealand
catchment, with higher flows in winter (June–August) and lower flows in summer.
The inflow data frame must contain at minimum `Date` and `HYD_flow` (m³ day⁻¹).
Temperature (`HYD_temp`, °C) and salinity (`CHM_salt`, PSU) columns are
recommended.

```{r inflow-data}
set.seed(42)
sim_dates <- seq(as.Date("2020-01-01"), as.Date("2021-12-31"), by = "day")
doy       <- as.integer(format(sim_dates, "%j"))

# Seasonal signal: peak winter flow ~day 200 (NZ Southern Hemisphere winter)
inflow_flow <- pmax(5000,
  80000 + 60000 * cos(2 * pi * (doy - 200) / 365) +
    rnorm(length(sim_dates), mean = 0, sd = 8000))

# River temperature: coolest in winter, warmest in summer
inflow_temp <- pmax(4, 15 - 8 * cos(2 * pi * (doy - 15) / 365))

inflow_data <- data.frame(
  Date     = sim_dates,
  HYD_flow = inflow_flow,
  HYD_temp = inflow_temp,
  CHM_salt = 0
)

head(inflow_data[, 1:3])
```

## Multiple outlet data

The key feature demonstrated in this vignette is the configuration of **two
outlets at different elevations**:

| Outlet | Elevation | Description |
|--------|-----------|-------------|
| **Penstock** | 130 m | Regulated bottom release; 20 m below full supply level |
| **Spillway** | -1 (surface) | Uncontrolled overflow at the water surface |

```{r outflow-data}
# Penstock: regulated base flow with small seasonal variation
penstock_flow <- pmax(0,
  60000 + 15000 * sin(2 * pi * (doy - 100) / 365) +
    rnorm(length(sim_dates), mean = 0, sd = 4000))

# Spillway: only active during the high-inflow winter period
spillway_flow <- pmax(0,
  25000 * cos(2 * pi * (doy - 200) / 365) +
    rnorm(length(sim_dates), mean = 0, sd = 5000))

penstock_data <- data.frame(
  Date     = sim_dates,
  HYD_flow = penstock_flow
)

spillway_data <- data.frame(
  Date     = sim_dates,
  HYD_flow = spillway_flow
)
```

Visualise the inflow and both outflows over the simulation period:

```{r plot-flows, fig.height = 4}
flows <- bind_rows(
  mutate(inflow_data[, c("Date", "HYD_flow")], stream = "Inflow"),
  mutate(penstock_data, stream = "Penstock (130 m)"),
  mutate(spillway_data, stream = "Spillway (surface)")
) |>
  filter(Date >= as.Date("2020-08-01"),
         Date <= as.Date("2021-06-30"))

ggplot(flows, aes(x = Date, y = HYD_flow / 1e3, colour = stream)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("Inflow"             = "#0065a9",
               "Penstock (130 m)"   = "#d73027",
               "Spillway (surface)" = "#4dac26"),
    name = NULL
  ) +
  labs(
    x     = "Date",
    y     = expression("Flow (×10"^3 * " m"^3 * " day"^{-1} * ")"),
    title = "Simulated inflow and outflow discharges"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The penstock provides a near-constant regulated release throughout the year,
while the spillway flows only during the high-flow winter months.

### Outlet positions on the hypsograph

It is important that each subsurface outlet elevation falls within the elevation
range of the hypsograph. Below we illustrate where the two outlets sit relative
to the reservoir hypsograph.

```{r plot-outlet-positions, fig.height = 5}
outlet_lines <- data.frame(
  elev  = c(130, max(hypsograph$elev)),
  label = c("Penstock (130 m)", "Spillway (surface)")
)

ggplot(hypsograph, aes(x = area / 1e6, y = elev)) +
  geom_ribbon(aes(xmin = 0, xmax = area / 1e6), fill = "#cce5f7", alpha = 0.6) +
  geom_line(linewidth = 1, colour = "#0065a9") +
  geom_hline(data = outlet_lines,
             aes(yintercept = elev, colour = label),
             linetype = "dashed", linewidth = 0.9) +
  scale_colour_manual(
    values = c("Penstock (130 m)"   = "#d73027",
               "Spillway (surface)" = "#4dac26"),
    name = "Outlet"
  ) +
  labs(
    x     = expression("Area (km"^2 * ")"),
    y     = "Elevation (m a.s.l.)",
    title = "Outlet positions on reservoir hypsograph"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Build the AEME object

We assemble the AEME object with `aeme_constructor()`. The `input` list must
include at minimum the hypsograph, meteorological data, and light extinction
coefficient (`Kw`).

```{r build-input}
input <- list(
  init_depth = depth,
  hypsograph = hypsograph,
  meteo      = met,
  use_lw     = TRUE,
  Kw         = 0.5   # light extinction coefficient (m⁻¹)
)
```

```{r create-aeme}
aeme <- aeme_constructor(
  lake  = reservoir,
  time  = time,
  input = input
)
aeme
```

## Add inflows

Inflows are added as a named list of data frames. The name (here `"river"`) is
used as the stream identifier in each model.

```{r add-inflow}
aeme <- add_inflows(aeme, data = list(river = inflow_data))
```

## Add multiple outflows

The `add_outflows()` function accepts:

- `data` – a named list of outflow data frames (each with at least `Date` and
  `HYD_flow` columns).
- `elevation` – a named list matching the names in `data`.  
  Use `elevation = -1` for a surface outlet; supply the elevation in metres
  above sea level (within the hypsograph range) for a subsurface outlet.

```{r add-outflows}
aeme <- add_outflows(
  aeme,
  data = list(
    penstock = penstock_data,
    spillway = spillway_data
  ),
  elevation = list(
    penstock = 130,   # 20 m below the full supply level
    spillway = -1     # surface outlet (overflow)
  )
)

# Inspect the outflows slot
outf <- outflows(aeme)
cat("Outflow names :", paste(names(outf$data),      collapse = ", "), "\n")
cat("Elevations    :",
    paste(names(outf$elevation),
          unlist(outf$elevation), sep = " = ", collapse = "; "), "\n")
```

Printing the `aeme` object now shows both outlets registered in the outflows
slot:

```{r print-aeme}
aeme
```

## Build model configurations

`build_aeme()` translates the AEME object into the configuration files required
by each model. Each model handles multiple outlets in its own way:

- **GLM-AED**: creates individual CSV boundary condition files
  (`bcs/outflow_<name>.csv`) and sets `num_outlet`, `outl_elvs`, and
  `outlet_type` in `glm3.nml`.
- **GOTM-WET**: creates individual data files (`inputs/outf_<name>.dat`) and
  updates `gotm.yaml`.
- **DYRESM-CAEDYM**: writes all outlets into a single `.wdr` file with a column
  per outlet and the number of outlets in the header.

```{r model-controls}
model_controls <- get_model_controls()
model          <- c("glm_aed")
path           <- file.path(tempdir(), "reservoir")
```

```{r build-aeme, eval=TRUE}
aeme <- build_aeme(
  aeme           = aeme,
  model          = model,
  model_controls = model_controls,
  path           = path,
  ext_elev       = 2,
  use_bgc        = FALSE,
  wb_method = 1
)
```

## Run the models

```{r run-models, eval=TRUE}
aeme <- run_aeme(aeme = aeme, model = model, path = path)
```

## View the output

```{r view-output, eval=TRUE}
plot_output(aeme = aeme)
```

## Summary

This vignette has shown how to configure AEME for a reservoir with multiple
outlets at different elevations. The key steps are:

1. **Define the reservoir geometry** using a more prismatic hypsograph
   (`volume_development` > 1.5) appropriate for a constructed reservoir.
2. **Create time series** for each inflow and each outlet independently.
3. **Register multiple outlets** with `add_outflows()`, providing a named
   `data` list and a matching named `elevation` list. Use `elevation = -1` for a
   surface outlet; provide the actual elevation (m a.s.l.) for subsurface
   outlets, ensuring values fall within the hypsograph elevation range.
4. **Build and run** the model ensemble with `build_aeme()` and `run_aeme()`.

Currently only one of the three hydrodynamic models, GLM-AED, supported by AEME 
(DYRESM-CAEDYM, GLM-AED, and GOTM-WET) can handle multiple outlets and produce 
separate outlet files during the build step, allowing each outlet's contribution
to be tracked independently.
