---
title: "Lake Rotoehu Water Balance and Evaporation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

This article demonstrates how to work with Lake Rotoehu (ID: 40188) to:

1. Set up an AEME model using data from the aemetools package
2. Explore different water balance approaches
3. Calibrate lake surface temperature
4. Extract and analyze evaporation data

Lake Rotoehu is a small lake in the Rotorua region of New Zealand's North Island. 
It has a surface area of approximately 8 km² and a maximum depth of around 13 m.

# Setup

First, load the required packages:

```{r setup, message=FALSE}
library(AEME)
library(aemetools)
library(bathytools)
library(ggplot2)
library(dplyr)
library(tmap)

tmap_mode("view")
```

# Load Lake Rotoehu data

The `aemetools` package provides access to pre-configured AEME objects for New 
Zealand lakes. We can load Lake Rotoehu using its ID (40188):

```{r load-lake, eval=TRUE}
# Install aemetools if needed
# remotes::install_github("limnotrack/aemetools")

# Load Lake Rotoehu
aeme <- aemetools::get_aeme(id = 40188)
aeme
```

# View hypsograph

For this demonstration, we'll generate a hypsograph based on Lake Rotoehu's 
known characteristics:

```{r hypsograph}
hyps <- get_hypsograph(aeme)

orig_hyps <- hyps |> 
  dplyr::filter(depth <= 0)

ggplot(orig_hyps, aes(x = area, y = depth)) +
  geom_line(colour = "#0065a9", linewidth = 0.8) +
  labs(
    x = "Area (km²)",
    y = "Depth (m)",
    title = "Original Hypsograph for Lake Rotoehu"
  ) +
  theme_bw()

```


Uh oh!

This does not look right. We will update the hypsograph using contours available on the Limnotrack API.

```{r get-contours}
shoreline <- aemetools::get_lake_shape(id = 40188)
# contours <- aemetools::get_depth_contours(id = 40188)
contours <- aemetools::lt_fetch("lake_contours",
                                filter = aemetools::lt_filter(lernzmp_id == "LID40188")) |> 
  dplyr::mutate(depth = depth_m)


```

Let's have a look what these contours look like:

```{r plot-contours}
tm_shape(contours) +
  tm_lines(col = "depth", scale = tm_scale("-brewer.blues")) +
  tm_shape(shoreline) +
  tm_borders(col = "black") +
  tm_title("Lake Rotoehu Depth Contours")

```

Use `bathytools` package to rasterise the contours and generate an updated
hypsograph:

```{r rasterise-contours, warning=FALSE}
bathy_raster <- rasterise_bathy(shoreline = shoreline,
                                contours = contours,
                                res = 8, crs = 2193)
```

Calculate the hypsograph from this raster:

```{r hypsograph-from-raster}
lake_elev <- max(contours$elevation_m)
upd_hyps <- bathytools::bathy_to_hypso(bathy_raster) |> 
  dplyr::mutate(elev = lake_elev + depth) 

ggplot(orig_hyps, aes(x = area, y = depth)) +
  geom_line(colour = "#0065a9", linewidth = 0.8) +
  geom_line(data = upd_hyps, aes(x = area, y = depth), colour = "#d73027", 
            linewidth = 0.8) +
  labs(
    x = "Area (km²)",
    y = "Depth (m)",
    title = "Updated Hypsograph for Lake Rotoehu"
  ) +
  theme_bw()
```
```{r}
aeme <- add_hypsograph(aeme, hypsograph = upd_hyps)
```




# Water balance approaches

AEME supports three different water balance methods:

## Method 1: No inflows or outflows

This method assumes a closed lake with no inflows or outflows. Water balance 
is determined only by precipitation and evaporation.

```{r wb-method-1}
model <- c("glm_aed")
path <- file.path(tempdir(), "rotoehu_wb1")

# Build with water balance method 1
aeme_wb1 <- aeme |>
  set_time(start = "2010-06-01", spin_up = 365) |> 
  build_aeme(
    aeme = ,
    model = model,
    path = path,
    use_bgc = FALSE,
    wb_method = 1
  ) |> 
  run_aeme()
```

```{r}
plot_output_base(aeme_wb1)
```


```{r}
plot_output(aeme_wb1, var_sim = "surft")
```

```{r}
plot_output(aeme_wb1, var_sim = "evap_rate")
```
```{r}
ann_summ_evap <- aeme_wb1 |> 
  get_var(var_sim = "evap_rate") |> 
  dplyr::mutate(year = lubridate::year(Date)) |>
  dplyr::group_by(year) |>
  dplyr::summarise(ann_evap_mm = sum(value, na.rm = TRUE) * 1000)
ann_summ_evap

```


## Method 2: Outflows calculated

This method calculates outflows based on the water balance, assuming precipitation 
and evaporation are known.

```{r wb-method-2}
path2 <- file.path(tempdir(), "rotoehu_wb2")

aeme_wb2 <- build_aeme(
  aeme = aeme,
  model = model,
  path = path2,
  use_bgc = FALSE,
  wb_method = 2
)
```

## Method 3: Inflows and outflows calculated

This method estimates both inflows and outflows from the water balance.

```{r wb-method-3}
path3 <- file.path(tempdir(), "rotoehu_wb3")

aeme_wb3 <- build_aeme(
  aeme = aeme,
  model = model,
  path = path3,
  use_bgc = FALSE,
  wb_method = 3
)
```

# Run models

Now we'll run the models with different water balance methods:

```{r run-models}
# Run method 1
aeme_wb1 <- run_aeme(aeme = aeme_wb1)

# Run method 2
aeme_wb2 <- run_aeme(aeme = aeme_wb2)

# Run method 3
aeme_wb3 <- run_aeme(aeme = aeme_wb3)
```

# Calibrating lake surface temperature

Lake surface temperature is a critical variable for calculating evaporation. 
AEME uses a regression approach to estimate surface temperature from air 
temperature when observations are limited.

The relationship follows the Stefan & Preud'homme (2007) approach:

$$T_s = a + b \times T_{air,5day}$$

where $T_s$ is the lake surface temperature, $T_{air,5day}$ is the 5-day 
moving average of air temperature, and $a$ and $b$ are calibration coefficients.

```{r surface-temp-calibration}
# Extract water balance data
wb1 <- water_balance(aeme_wb1)

# Plot surface temperature vs air temperature
wb_data <- wb1$data$wbal

ggplot(wb_data, aes(x = Date, y = Ts)) +
  geom_line(colour = "#0065a9", linewidth = 0.8) +
  labs(
    x = "Date",
    y = "Surface Temperature (°C)",
    title = "Estimated Lake Surface Temperature",
    subtitle = "Based on air temperature regression"
  ) +
  theme_bw()
```

You can also visualize the relationship between air and surface temperature:

```{r temp-relationship}
# Create a plot showing the calibration relationship
met_with_ts <- met |>
  left_join(wb_data |> select(Date, Ts), by = "Date") |>
  mutate(T5avg = zoo::rollmean(MET_tmpair, 5, na.pad = TRUE, align = "right"))

ggplot(met_with_ts, aes(x = T5avg, y = Ts)) +
  geom_point(alpha = 0.3, colour = "#0065a9") +
  geom_smooth(method = "lm", colour = "#d73027", se = TRUE) +
  labs(
    x = "5-day Average Air Temperature (°C)",
    y = "Surface Temperature (°C)",
    title = "Lake Surface Temperature Calibration"
  ) +
  theme_bw()
```

# Extracting evaporation

One of the key outputs from the water balance is evaporation. AEME calculates 
evaporation using bulk aerodynamic methods that account for wind speed, air 
temperature, humidity, and lake surface temperature.

## Extract evaporation data

```{r extract-evaporation}
# Get evaporation from water balance
evap_data <- wb_data |>
  select(Date, evap_flux, evap_m3, area, Ts) |>
  mutate(
    evap_mm = abs(evap_flux) * 1000,  # Convert m to mm
    evap_m3_day = abs(evap_m3)        # Daily evaporation volume
  )

head(evap_data)
```

## Visualize evaporation

```{r plot-evaporation}
# Plot evaporation flux over time
ggplot(evap_data, aes(x = Date, y = evap_mm)) +
  geom_line(colour = "#0065a9", linewidth = 0.8) +
  labs(
    x = "Date",
    y = "Evaporation (mm/day)",
    title = "Daily Evaporation from Lake Rotoehu"
  ) +
  theme_bw()
```

## Evaporation vs temperature

Evaporation rates are strongly influenced by surface temperature:

```{r evap-vs-temp}
ggplot(evap_data, aes(x = Ts, y = evap_mm)) +
  geom_point(alpha = 0.3, colour = "#0065a9") +
  geom_smooth(method = "loess", colour = "#d73027") +
  labs(
    x = "Surface Temperature (°C)",
    y = "Evaporation (mm/day)",
    title = "Evaporation vs Surface Temperature"
  ) +
  theme_bw()
```

## Annual evaporation totals

Calculate total annual evaporation:

```{r annual-evaporation}
# Calculate total annual evaporation
annual_evap <- evap_data |>
  summarise(
    total_evap_mm = sum(evap_mm, na.rm = TRUE),
    total_evap_m3 = sum(evap_m3_day, na.rm = TRUE),
    mean_daily_evap_mm = mean(evap_mm, na.rm = TRUE)
  )

# Display results
cat("Annual Evaporation Summary:\n")
cat(sprintf("  Total evaporation: %.1f mm/year\n", annual_evap$total_evap_mm))
cat(sprintf("  Total volume: %.0f m³/year\n", annual_evap$total_evap_m3))
cat(sprintf("  Mean daily evaporation: %.2f mm/day\n", 
            annual_evap$mean_daily_evap_mm))
```

## Compare water balance components

Finally, let's visualize all water balance components together:

```{r water-balance-components}
# Prepare data for plotting
wb_components <- wb_data |>
  select(Date, rain, evap_m3, HYD_flow, HYD_outflow) |>
  mutate(
    evap_m3 = abs(evap_m3)
  ) |>
  tidyr::pivot_longer(
    cols = c(rain, evap_m3, HYD_flow, HYD_outflow),
    names_to = "Component",
    values_to = "Volume"
  ) |>
  mutate(
    Component = factor(Component, 
                       levels = c("rain", "HYD_flow", "evap_m3", "HYD_outflow"),
                       labels = c("Precipitation", "Inflow", "Evaporation", 
                                  "Outflow"))
  )

ggplot(wb_components, aes(x = Date, y = Volume / 1e3, colour = Component)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(
    values = c("Precipitation" = "#4dac26",
               "Inflow" = "#0065a9",
               "Evaporation" = "#d73027",
               "Outflow" = "#f4a582")
  ) +
  labs(
    x = "Date",
    y = expression("Volume (×10"^3 * " m"^3 * " day"^{-1} * ")"),
    title = "Water Balance Components - Lake Rotoehu",
    colour = "Component"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

# Summary

This article demonstrated how to:

1. **Load Lake Rotoehu data** using the lake ID (40188) with the aemetools package
2. **Build an AEME model** with appropriate inputs and hypsograph
3. **Explore three water balance methods**:
- Method 1: No inflows/outflows (closed lake)
- Method 2: Calculated outflows
- Method 3: Calculated inflows and outflows
4. **Calibrate lake surface temperature** using the Stefan & Preud'homme 
regression approach
5. **Extract evaporation data** from the water balance calculations
6. **Visualize evaporation patterns** and relationships with temperature

The water balance approach in AEME provides a flexible framework for 
understanding lake hydrology and estimating critical variables like evaporation, 
even when direct measurements are not available.

# References

- Stefan, H. G., & Preud'homme, E. B. (2007). Stream temperature estimation 
from air temperature. *JAWRA Journal of the American Water Resources 
Association*, 29(1), 27-45.
