---
title: "Simstrat-AED2: The 1-D k-epsilon lake model coupled with AED2"
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  message   = FALSE,
  warning   = FALSE
)
asciicast::init_knitr_engine(
  startup = quote({
    library(AEME)
    set.seed(1)
  }),
  echo = TRUE,
  echo_input = FALSE
)
options(AEME.inform = FALSE)
library(AEME)
install_simstrat_aed2()
```

<a href="#parameter-library" class="skip-btn">&#8681; Skip to Parameter Library</a>

<style>
.skip-btn {
  display: inline-block;
  margin: 8px 0 16px 0;
  padding: 8px 16px;
  background-color: #2c7bb6;
  color: white !important;
  border-radius: 6px;
  text-decoration: none;
  font-size: 14px;
  font-weight: bold;
  transition: background-color 0.2s;
}
.skip-btn:hover {
  background-color: #1a5a8a;
  text-decoration: none;
}
</style>

## Introduction

### Simstrat

**Simstrat** is a one-dimensional (1-D), k-epsilon lake model for the
physical simulation of stratification and mixing in water reservoirs,
including basin morphology, atmospheric interaction, and inflow/outflow.
A reservoir is represented as a horizontally-averaged, vertically-resolved
water column whose evolution is driven by atmospheric forcing. Simstrat was
originally described by Goudsmit *et al.* (2002) and has since been applied
to lakes and reservoirs with a wide range of physical properties. Its key
architectural features include:

- **k-epsilon turbulence closure** — turbulent kinetic energy (*k*) and its
  dissipation rate (epsilon) are explicitly modelled, alongside water
  velocities, temperature, salinity, and buoyancy.
- **Internal seiche parameterisation** — a fraction of wind energy is
  converted into seiche (internal wave) energy, which is redistributed
  through the water column and contributes to deep mixing — a process
  1-D models otherwise struggle to represent.
- **Flexible inflows** — inflows can be placed at a fixed depth ("manual"
  mode) or allowed to plunge and stratify according to their density
  ("density-driven" mode), entraining ambient water as they sink until
  reaching neutral buoyancy.
- **Ice/snow model** — a 3-layer ice model (black ice, white ice, snow)
  based on MyLake, with depth-varying light penetration.
- **Coupling with AED2** — from version 3.0 onwards, Simstrat always
  includes a coupling to the AED2 biogeochemical library (the coupling can
  be switched off for purely physical simulations).

Source code: :github: <https://github.com/Eawag-AppliedSystemAnalysis/Simstrat>

#### References

Goudsmit, G-H., Burchard, H., Peeters, F., and Wuest, A. (2002). Application
of k-epsilon turbulence models to enclosed basins: The role of internal
seiches. *Journal of Geophysical Research: Oceans*, 107(C12), 23-1--23-13.

Schmid, M. and Koster, O. (2016). Excess warming of a Central European lake
driven by solar brightening. *Water Resources Research*, 52, 8103--8116.
(Heat-flux parameterisation used by Simstrat since v1.6.)

Gaudard, A., Schwefel, R., Vinna, L.R., Schmid, M., Wuest, A., and Bouffard,
D. (2017). Optimizing the parameterization of deep mixing and internal
seiches in one-dimensional hydrodynamic models: a case study with Simstrat
v1.3. *Geoscientific Model Development*, 10, 3411--3423. (An applied
sensitivity/calibration study of Simstrat's seiche and mixing parameters.)

Schwefel, R., Gaudard, A., Wuest, A., and Bouffard, D. (2016). Effects of
climate change on deepwater oxygen and winter mixing in a deep lake (Lake
Geneva): Comparing observational findings and modeling. *Water Resources
Research*, 52, 8811--8826. (An application of Simstrat to a large,
deep, dimictic lake.)

---

### The AED2 biogeochemical library

Simstrat couples to the same **AED2** (Aquatic Ecodynamics v2) library used
by the GLM-AED workflow in AEME (see the
[GLM-AED article](glm-aed.html)) — confirmed directly: the phytoplankton
and zooplankton group-parameter files AEME bundles for the two models
(`aed2_phyto_pars.nml`, `aed2_zoop_pars.nml`) are byte-identical. Each
biogeochemical process is implemented as a self-contained module that can be
switched on or off independently.

Further information is available from the
[AED science website](https://aquatic.science.uwa.edu.au/research/models/AED/index.html),
which the Simstrat User Manual itself refers readers to for AED2 details.

#### AED2 modules available in Simstrat-AED2

| Module | AED2 name | Description |
|--------|-----------|-------------|
| Oxygen | `aed2_oxygen` | Dissolved-oxygen dynamics including reaeration and sediment oxygen demand (SOD). |
| Carbon | `aed2_carbon` | Dissolved inorganic carbon (DIC), pH, and methane (CH~4~) cycling, including air-water gas exchange. |
| Silica | `aed2_silica` | Reactive silica cycling — important for diatom growth. |
| Nitrogen | `aed2_nitrogen` | Ammonium (NH~4~^+^), nitrite/nitrate (NO~2~^-^/NO~3~^-^), and N~2~O; nitrification, denitrification. |
| Phosphorus | `aed2_phosphorus` | Dissolved reactive phosphorus (FRP), including PO~4~ adsorption. |
| Organic matter | `aed2_organic_matter` | Particulate and dissolved organic carbon/nitrogen/phosphorus pools; hydrolysis and mineralisation. |
| Phytoplankton | `aed2_phytoplankton` | Multi-group phytoplankton dynamics: growth, respiration, nutrient uptake, light limitation, settling, mortality. |
| Zooplankton | `aed2_zooplankton` | Zooplankton grazing and dynamics. |
| Totals | `aed2_totals` | Diagnostic aggregates: TN, TP, TOC. |

: AED2 biogeochemical modules available in Simstrat-AED2.

Unlike GLM-AED (where `set_glm_aed_models()` lets you toggle modules
explicitly), Simstrat-AED2's active module list is derived **automatically**
from the biogeochemical variables requested via `model_controls`/
`set_vars_sim()` — AEME writes the corresponding entries into `aed2.nml`'s
`&aed2_models` block for you (including force-including any module that an
active module's target-variable links depend on, e.g. phytoplankton
depends on carbon, nitrogen, phosphorus, silica, and oxygen).

---

### Simstrat-AED2 Parameter Library {#parameter-library}

The `simstrat_aed2_parameter_library` dataset provides a comprehensive list
of parameters used in the Simstrat-AED2 configuration, including their
default (template) values, units, and a brief description of each
parameter's role. The physical parameters are documented in Table 1 of the
Simstrat User Manual; the AED2 biogeochemical parameters largely share their
descriptions with `glm_aed_parameter_library` (see the
[GLM-AED article](glm-aed.html)), since both models couple to the same AED2
library.

```{r simstrat-aed2-lib, echo=FALSE}
# Prep the data
df <- simstrat_aed2_parameter_library |>
  dplyr::mutate(
    # Make source URLs into clickable hyperlinks
    source = ifelse(
      !is.na(source) & grepl("^http", source),
      paste0('<a href="', source, '" target="_blank">&#x1F517;</a>'),
      source
    ),
    # Make symbol italic if not NA
    symbol = ifelse(!is.na(symbol), paste0("<i>", symbol, "</i>"), NA)
  )

DT::datatable(
  df,
  escape = FALSE,  # allows HTML rendering in cells
  rownames = FALSE,
  colnames = c(
    "Module", "Group", "Parameter", "Label",
    "Symbol", "Description", "Units", "Default", "Source"
  ),
  filter = "top",  # per-column filters
  extensions = c("Buttons", "FixedHeader"),
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel", "pdf"),
    pageLength = 10,
    lengthMenu = c(10, 20, 50, 100),
    fixedHeader = TRUE,
    scrollX = TRUE,
    autoWidth = TRUE,
    columnDefs = list(
      list(width = "200px", targets = 5),   # Description column wider
      list(width = "80px",  targets = c(0, 1, 6, 7)),  # narrow cols
      list(className = "dt-center", targets = c(4, 6, 7, 8))  # center symbol/units/default/source
    )
  )
) |>
  DT::formatStyle(
    "module",
    fontWeight = "bold"
  ) |>
  DT::formatStyle(
    columns = 1:9,
    fontSize = "13px"
  ) |>
  DT::formatStyle(
    "default",
    color = "#2c7bb6",
    fontWeight = "bold"
  )

```

### AED2 Phytoplankton Group Parameters

Simstrat-AED2 and GLM-AED share the same bundled AED2 phytoplankton
parameter database (`aed2_phyto_pars.nml`), so the group parameter table is
identical to the one shown in the [GLM-AED article](glm-aed.html):

```{r aed2-phyto-lib, echo=FALSE}

phyto_file <- system.file(
  "extdata/glm_aed/aed_phyto_pars.csv",
  package = "AEME"
)
aed_phyto_params <- readr::read_csv(phyto_file, show_col_types = FALSE)

param_desc <- aed_phyto_pars |>
  dplyr::select(parameter_name, description) |>
  dplyr::distinct(parameter_name, .keep_all = TRUE)

df_wide <- aed_phyto_params |>
  dplyr::left_join(param_desc, by = c("p_name" = "parameter_name"))

# Get group column names (everything except p_name and description)
group_cols <- setdiff(names(df_wide), c("p_name", "description"))
n_groups   <- length(group_cols)

DT::datatable(
  df_wide,
  escape    = FALSE,
  rownames  = FALSE,
  filter    = "top",
  extensions = c("Buttons", "FixedHeader", "FixedColumns"),
  options   = list(
    dom        = "Bfrtip",
    buttons    = c("colvis", "csv", "excel"),
    pageLength = 25,
    lengthMenu = c(10, 25, 50),
    fixedHeader = TRUE,
    scrollX    = TRUE,
    scrollY    = "600px",
    fixedColumns = list(leftN = 1, rightN = 1),
    autoWidth  = FALSE,
    columnDefs = list(
      list(width = "120px", targets = 0),
      list(width = "80px",  targets = seq_len(n_groups)),
      list(width = "300px", targets = n_groups + 1),
      list(className = "dt-center", targets = seq_len(n_groups))
    )
  )
) |>
  DT::formatStyle(
    "p_name",
    fontWeight = "bold",
    fontFamily = "monospace"
  ) |>
  DT::formatStyle(
    "description",
    color    = "grey40",
    fontSize = "12px"
  ) |>
  DT::formatRound(
    columns = group_cols,
    digits  = 3
  ) |>
  DT::formatStyle(
    columns   = 1:(n_groups + 2),
    fontSize  = "13px"
  )

```

---

## Getting started

```{r setup}
library(AEME)
library(ggplot2)
library(dplyr)
```

We will use the example lake dataset bundled with the AEME package throughout
this vignette.

```{r load-aeme}
aeme_file <- system.file("extdata/aeme.rds", package = "AEME")
aeme     <- readRDS(aeme_file)
aeme
```

```{asciicast show-aeme}
#| echo: FALSE
aeme_file <- system.file("extdata/aeme.rds", package = "AEME")
aeme     <- readRDS(aeme_file)
aeme
```

---

## Building a Simstrat-AED2 simulation

### Model controls

As with the other models, `get_model_controls()` returns a data frame that
governs which biogeochemical variables are simulated and how they are
initialised. Pass `use_bgc = TRUE` to enable AED2.

```{r model-controls}
model_controls <- get_model_controls(use_bgc = TRUE)
head(model_controls)
```

```{r set-vars}
vars_sim <- c(
  "HYD_strat",   # stratification flag
  "HYD_temp",    # water temperature
  "HYD_thmcln",  # thermocline depth
  "CHM_oxy",     # dissolved oxygen
  "CHM_oxycln",  # oxycline depth
  "NIT_amm",     # ammonium
  "NIT_nit",     # nitrate
  "NIT_tn",      # total nitrogen
  "PHS_frp",     # filterable reactive phosphorus
  "PHS_tp",      # total phosphorus
  "PHY_tchla"    # total chlorophyll-a
)

model_controls <- set_vars_sim(
  model_controls = model_controls,
  vars_sim       = vars_sim
)
```

### Build the model

We will build the model in a directory called `aeme`,

```{r config-path}
path <- "aeme"
```

`build_aeme()` translates the AEME object into all configuration files
required by Simstrat-AED2: `simstrat.par`, `Bathymetry.dat`, `Grid.dat`,
`MeteoForcing.dat`, `InitialConditions.dat`, the inflow/outflow/temperature/
salinity time-depth grid files, and (with `use_bgc = TRUE`) `aed2.nml` plus
the `AED2_initcond/`/`AED2_inflow/` folders.

```{r build-aeme}
model <- "simstrat_aed2"

aeme <- build_aeme(
  aeme           = aeme,
  model          = model,
  path           = path,
  model_controls = model_controls,
  ext_elev       = 3,
  use_bgc        = TRUE
)
```

The configuration files are stored in the `configuration` slot of the `aeme`
object. For Simstrat-AED2 the slot contains the parsed `simstrat.par` (under
`hydrodynamic`) and `aed2.nml` (under `bgc`):

```{r inspect-config}
cfg <- configuration(aeme)
names(cfg[["simstrat_aed2"]])
cfg[["simstrat_aed2"]]$hydrodynamic$ModelConfig[c("InflowMode", "IceModel",
                                                  "SnowModel", "TurbulenceModel")]
```

---

## Simstrat-AED2 specific features {#simstrat-features}

### Automatic water balance fitting

Before writing the Simstrat configuration, `build_aeme()` estimates a lake
water balance for each model, fitting a weir-style outflow correction so
that simulated lake level tracks observations. This uses each model's own
evaporation formulation — Simstrat's is a Livingstone & Imboden-style wind
function (free convection driven by the air-water temperature difference,
combined in quadrature with a wind-driven forced-convection term, using a
Gill (1992) saturation vapour pressure), taken directly from Simstrat's own
`strat_forcing.f90` rather than the simpler bulk-aerodynamic formula shared
by GLM-AED and DYRESM-CAEDYM. This matters because the fitted outflow
correction is added directly to the model's own outflow file, so using the
wrong evaporative demand would bias the fitted water balance.

```{r water-balance-messages, eval=FALSE}
# (build_aeme() prints, among other things:)
#> -- Calculating water balance --
#> Estimating lake water levels for simstrat_aed2
#>   i Optimizing parameters for water balance
#>   v Optimization Complete: C = 0.3571, h_inv = 23.4849, Final RMSE = 0.136
```

### Inflow modes

Simstrat supports two inflow placement strategies, set via
`ModelConfig.InflowMode` in `simstrat.par`:

- **Manual (`InflowMode = 1`)** — inflows enter at a fixed depth given in
  the inflow file, unaffected by water density.
- **Density-driven (`InflowMode = 2`)** — inflows plunge and stratify
  according to their density relative to the water column, entraining
  water as they sink until reaching neutral buoyancy. This is the default
  used by AEME's Simstrat-AED2 template, since a 1-D model would otherwise
  spread an entire inflow across whole horizontal layers at an arbitrary
  fixed depth, which can noticeably bias the vertical distribution of heat
  and dissolved constituents.

### Ice and snow

The bundled template also enables Simstrat's 3-layer ice/snow model
(`IceModel = 1`, `SnowModel = 1`), which tracks black ice, white (snow) ice,
and snow thickness, with a distinct light-penetration formulation for each.
This is most relevant for seasonally-ice-covered lakes; the `freez_temp`,
`snow_temp`, `wat_albedo`, and `p_sw_ice` parameters in the parameter
library above configure it.

### The `simstrat.par` template is copied once, not re-applied

`build_simstrat()` only copies AEME's bundled `simstrat.par` template into a
lake's `simstrat_aed2/` directory the *first* time that lake is built —
every subsequent `build_aeme()` call reads back and rewrites that same
per-lake file (updating only the specific fields `build_aeme()` itself
manages, like file paths, `Simulation` dates, and `lat`/`p_air`). This
keeps any manual edits you make to a lake's own `simstrat.par` from being
silently discarded on rebuild, but it also means a change to the *package
template* (e.g. `AED2Config.OutputDiagnosticVars`, or any `ModelParameters`
default) only takes effect for lakes built *after* the change — an
already-built lake directory keeps whatever it already had. To pick up a
template change on an existing lake, either delete that lake's
`simstrat_aed2/` directory and rebuild, or edit the per-lake `simstrat.par`
directly (as shown for `OutputDiagnosticVars` below).

---

## Running Simstrat-AED2

```{r run-aeme}
aeme <- run_aeme(aeme = aeme, model = model, path = path)
```

---

## Visualising model output

### Temperature and stratification

`plot_output()` produces a filled contour plot for depth-varying variables
(e.g. temperature) and a time-series line plot for scalar variables:

```{r plot-temp, fig.cap="Simulated water temperature (degC) at each model depth over time."}
plot_output(aeme = aeme, model = model, var_sim = "HYD_temp")
```

### Water quality variables

Any variable that was listed in `vars_sim` and is present in the output can
be plotted the same way:

```{r plot-oxy, fig.cap="Simulated dissolved oxygen (mmol O2 m-3) profiles."}
plot_output(aeme = aeme, model = model, var_sim = "CHM_oxy")
```

```{r plot-nit, fig.cap="Simulated nitrate (mmol N m-3) profiles."}
plot_output(aeme = aeme, model = model, var_sim = "NIT_nit")
```

`PHY_tchla` (total chlorophyll-*a*) also plots without any extra setup:

```{r plot-tchla, fig.cap="Simulated total chlorophyll-a (ug/L) profiles."}
plot_output(aeme = aeme, model = model, var_sim = "PHY_tchla")
```

`PHY_tchla` is not itself an AED2 *state* variable — it is a *diagnostic*
variable, computed internally by `aed2_phytoplankton` from the active
phytoplankton groups' biomass. It only reaches `output.nc` because AEME's
bundled `simstrat.par` template sets `AED2Config.OutputDiagnosticVars` to
`true`. See the [AED2 diagnostic variables](#aed2-diagnostic-variables)
section below for what that switch controls, why it matters, and how to
interpret the wider set of diagnostics it unlocks.

---

## Assessing model performance

When observations are stored in the `aeme` object, `assess_model()` computes
a suite of skill metrics (RMSE, NSE, bias, Pearson *r*, etc.) for each
simulated variable:

```{r assess-model, eval=TRUE}
skill <- assess_model(aeme = aeme, model = model)
skill
```

---

## AED2 diagnostic variables {#aed2-diagnostic-variables}

### State variables vs. diagnostic variables

Every AED2 module tracks two kinds of quantities:

- **State variables** (`OXY_oxy`, `NIT_amm`, `NIT_nit`, `PHS_frp`,
  `CAR_dic`, `OGM_doc`, `PHY_<group>`, ...) are the variables AED2 actually
  integrates forward in time — mass-balance pools with their own initial
  condition, inflow loading, and (where applicable) sediment exchange.
  Simstrat writes these to `output.nc` unconditionally whenever the owning
  module is active — that's why `CHM_oxy`/`NIT_nit` above needed no extra
  configuration.
- **Diagnostic variables** (`PHY_TCHLA`, `PHY_GPP`, `OXY_atm_oxy_flux`,
  `NIT_nitrif`, ...) are quantities *computed from* the state variables at
  each timestep — reaction rates, fluxes across an interface, or derived
  aggregates — but not themselves part of the mass balance. They exist
  purely for interpretation and diagnosis of *why* the state variables are
  behaving the way they are.

Because there can be dozens of these per module, Simstrat gates all AED2
diagnostics behind a single switch, `AED2Config.OutputDiagnosticVars` in
`simstrat.par`. AEME's bundled template now sets this to `true` (it
defaulted to `false` in earlier package versions), so diagnostics are
available out of the box — as demonstrated by `PHY_tchla` above. You can
confirm the setting for a built lake with:

```{r check-diag-switch, eval=TRUE}
sim_dir <- file.path(get_lake_dir(aeme, path), model)
par <- jsonlite::fromJSON(file.path(sim_dir, "simstrat.par"), simplifyVector = FALSE)
par$AED2Config$OutputDiagnosticVars
```

If you build on top of a lake directory that was created with an older
version of the template (`OutputDiagnosticVars = false` already written to
its own `simstrat.par`), the switch will *not* update automatically — see
[Simstrat-AED2 specific features](#simstrat-features) above for why `build_aeme()` only copies
the template into directories that don't already have a `simstrat.par`. Flip
it manually and re-run in that case, exactly as shown for `sim_dir`/`par`
above but setting `par$AED2Config$OutputDiagnosticVars <- TRUE` before
writing it back out.

### What's available, and how to interpret it

Diagnostics fall into a few interpretable categories. Units and
descriptions below are taken directly from the AED2 source
(`aed2_oxygen.F90`, `aed2_carbon.F90`, `aed2_nitrogen.F90`,
`aed2_phosphorus.F90`, `aed2_silica.F90`, `aed2_organic_matter.F90`,
`aed2_phytoplankton.F90`):

**Interface exchange fluxes** — mass crossing the atmosphere-water or
sediment-water interface. Verified from source (`aed2_oxygen.F90`'s
`aed2_calculate_surface_oxygen`/`aed2_calculate_benthic_oxygen`): **positive
means flux *into* the water column** (a source to the pelagic), negative
means *out of* the water column (a sink), consistently for both interface
types.

| Variable | Units | Description |
|----------|-------|-------------|
| `OXY_atm_oxy_flux` | mmol/m2/d | O~2~ exchange across the atm/water interface |
| `OXY_sed_oxy` | mmol/m2/d | O~2~ exchange across the sediment/water interface (sediment oxygen demand, when negative) |
| `CAR_atm_co2_flux` | mmol/m2/d | CO~2~ exchange across the atm/water interface |
| `CAR_atm_ch4_flux` | mmol/m2/d | CH~4~ exchange across the atm/water interface |
| `CAR_sed_dic` | mmol/m2/d | DIC sediment flux |
| `CAR_sed_ch4` | mmol/m2/d | CH~4~ sediment flux |
| `SIL_sed_rsi` | mmol/m2/d | Reactive silica sediment flux |
| `NIT_sed_amm` / `NIT_sed_nit` | mmol/m2/d | Ammonium / nitrate sediment flux |
| `NIT_atm_din_flux` | mmol/m2/d | Dissolved inorganic nitrogen atmospheric deposition flux |
| `PHS_sed_frp` | mmol/m2/d | Phosphate (FRP) sediment flux |
| `PHS_atm_dip_flux` | mmol/m2/d | Particulate inorganic P dry deposition flux |
| `OGM_Psed_poc`/`Psed_pon`/`Psed_pop` | mmol/m2/s | POC/PON/POP settling flux to the sediment |
| `PHY_Psed_phy` | mmol/m2/s | Phytoplankton settling flux to the sediment |

**Internal reaction rates** — volumetric transformation rates within the
water column (mmol/m3/d); always non-negative (a rate of zero simply means
the process isn't occurring, e.g. no oxygen for nitrification):

| Variable | Description |
|----------|-------------|
| `NIT_nitrif` | Nitrification rate (NH~4~ -> NO~3~) |
| `NIT_denit` | De-nitrification rate (NO~3~ -> N~2~) |
| `NIT_anammox` | Anammox rate (NH~4~ + NO~2~ -> N~2~) |
| `NIT_dnra` | Dissimilatory nitrate reduction to ammonium (DNRA) rate |
| `CAR_ch4ox` | Methane oxidation rate |

**Water-column state diagnostics** — derived quantities describing the
current state, not fluxes:

| Variable | Units | Description |
|----------|-------|-------------|
| `OXY_sat` | % | Oxygen saturation |
| `CAR_pCO2` | atm | Partial pressure of CO~2~ |
| `CAR_CO2` | mmol/m3 | Dissolved (aqueous) CO~2~ concentration |
| `OGM_CDOM` | /m | Chromophoric dissolved organic matter (a light-absorption proxy) |

**Phytoplankton production and nutrient uptake** — from `aed2_phytoplankton`,
computed whenever that module is active:

| Variable | Units | Description |
|----------|-------|-------------|
| `PHY_TCHLA` | ug/L | Total chlorophyll-*a*, summed across active groups |
| `PHY_GPP` | mmol/m3/d | Gross primary production |
| `PHY_NCP` | mmol/m3/d | Net community production (GPP less respiration/losses) |
| `PHY_NUP_no3` / `PHY_NUP_nh4` | mmol/m3/d | Nitrogen uptake, as nitrate / ammonium |
| `PHY_PUP` | mmol/m3/d | Phosphorus uptake |
| `PHY_CUP` | mmol/m3/d | Carbon uptake |
| `PHY_IN` / `PHY_IP` | mmol/m3 | Total phytoplankton internal nitrogen / phosphorus |
| `PHY_<group>_IN` / `_IP` / `_NtoP` | mmol/m3, mmol/m3, - | Per-group internal N, internal P, and N:P ratio (one set per active phytoplankton group, e.g. `PHY_diatom_IN`) |

A few additional per-group diagnostics (`_fI`, `_fNit`, `_fPho`, `_fSil`,
`_fT`, `_fSal` limitation factors 0-1, `PPR`/`NPR` production ratios, `PAR`)
and an organic-matter photolysis rate are gated behind a further
`extra_diag = true` switch in `aed2.nml` (`false` in the bundled template,
since they roughly double the number of output variables for detail most
users won't need day-to-day).

### Example: production and gas exchange

`PHY_tchla` above plots through `plot_output()` because it already has a
`var_aeme` entry in AEME's `key_naming` catalog, mapped to Simstrat-AED2's
`PHY_TCHLA`. Most of the diagnostics tabulated above (`PHY_GPP`,
`OXY_atm_oxy_flux`, `OXY_sat`, ...) do **not** have a `var_aeme` entry yet —
AEME hasn't adopted every AED2 diagnostic into its standardised variable
catalog. `plot_output()`/`assess_model()` only recognise `var_aeme` names,
and passing an unregistered one through can silently resolve to the *wrong*
variable via `guess_aeme_vars()`'s fuzzy keyword matching rather than
failing loudly (`"PHY_GPP"`, for instance, fuzzy-matches to `"PHY_green"` —
a phytoplankton group's biomass, not gross primary production — since
neither name is registered). Until these diagnostics get their own
`var_aeme` entries, read them straight from `output.nc` instead:

```{r read-diagnostics-nc}
outfile <- get_model_outfile(aeme = aeme)
nc <- ncdf4::nc_open(outfile$simstrat_aed2)

time_sec <- ncdf4::ncvar_get(nc, "time")
origin <- gsub("seconds since ", "", ncdf4::ncatt_get(nc, "time", "units")$value)
dates <- as.Date(as.POSIXct(time_sec, origin = origin, tz = "UTC"))

# PHY_GPP is depth-resolved (z, time); take the near-surface layer (z[1] = 0 m,
# since Output.OutputDepthReference = "surface")
gpp_surf <- ncdf4::ncvar_get(nc, "PHY_GPP")[1, ]

# OXY_atm_oxy_flux is a surface-only "sheet" diagnostic (time only, no depth)
atm_oxy_flux <- ncdf4::ncvar_get(nc, "OXY_atm_oxy_flux")

ncdf4::nc_close(nc)

diag_df <- data.frame(Date = dates, GPP = gpp_surf, atm_oxy_flux = atm_oxy_flux)
head(diag_df)
```

```{r plot-diagnostics-nc, fig.cap="Near-surface gross primary production and atmospheric oxygen exchange flux."}
ggplot(diag_df, aes(x = Date)) +
  geom_line(aes(y = GPP, colour = "Near-surface GPP (mmol C m-3 d-1)")) +
  geom_line(aes(y = atm_oxy_flux, colour = "Atm. O2 flux (mmol O2 m-2 d-1)")) +
  labs(y = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

A sustained positive `OXY_atm_oxy_flux` indicates the lake surface is
undersaturated and drawing oxygen in from the atmosphere (e.g. following a
period of high respiration or low photosynthesis); a sustained negative
flux indicates supersaturation venting oxygen to the atmosphere (typically
during a strong phytoplankton bloom). Compare its timing against `PHY_GPP`
and `OXY_sat` to build a physical picture of the oxygen budget, rather than
reading `CHM_oxy` in isolation.

---

## Calibration

### The `simstrat_aed2_parameters` dataset

`simstrat_aed2_parameters` is a ready-to-use calibration parameter set:
physical parameters from `simstrat.par` and a subset of AED2
initial-concentration parameters from `aed2.nml`, each with the template
value and a +/-50% range suitable as a starting point for a sensitivity
analysis or calibration exercise.

```{r simstrat-aed2-parameters, echo=FALSE}
DT::datatable(
  simstrat_aed2_parameters,
  rownames = FALSE,
  filter = "top",
  extensions = c("Buttons", "FixedHeader"),
  options = list(
    dom = "Bfrtip",
    buttons = c("csv", "excel"),
    pageLength = 10,
    lengthMenu = c(10, 20, 50),
    fixedHeader = TRUE,
    scrollX = TRUE
  )
) |>
  DT::formatStyle(columns = seq_len(ncol(simstrat_aed2_parameters)),
                  fontSize = "13px")
```

### Retrieving parameters by module

`get_aeme_parameters()` provides a convenient way to query the AEME
parameter library for Simstrat-AED2 parameters:

```{r get-params, eval=TRUE}
# Physical (hydrodynamic) parameters
phys_params <- get_aeme_parameters(model = "simstrat_aed2",
                                   module = "hydrodynamic")
phys_params[, c("name", "value", "min", "max")]
```

```{r get-params-bgc, eval=TRUE}
# AED2 biogeochemical initial-concentration parameters
bgc_params <- get_aeme_parameters(model = "simstrat_aed2",
                                  module = "bgc")
bgc_params[, c("name", "value", "min", "max")]
```

These parameter sets — together with `simstrat_aed2_parameter_library`'s
richer descriptions above — are the natural inputs to an automated
calibration workflow via the
[aemetools](https://github.com/limnotrack/aemetools) package, which wraps
`build_aeme()`/`run_aeme()`/`assess_model()` in an optimiser.

### Simstrat's native PEST-based calibration

Independently of AEME/aemetools, Simstrat ships its own parameter-estimation
workflow built on [PEST](http://www.pesthomepage.org/), described in
section 8 of the Simstrat User Manual. In short: a PEST control file
specifies parameter values/ranges and observation data; a template file
mirrors `simstrat.par` but with parameter placeholders instead of values;
and a Python script (`PEST.py`, bundled in the Simstrat repository)
generates the PEST configuration and launches the optimisation
(`PEST.runPEST(configFile)`), which can run in parallel across multiple
CPUs. This is a useful option if you want to calibrate a standalone
Simstrat-AED2 run outside of the AEME/R ecosystem, using the same
`simstrat.par`/`aed2.nml` files AEME writes.

---

## Working with the Simstrat configuration directly

The raw `simstrat.par` (JSON) file and `aed2.nml` file can be read,
modified, and written directly:

```{r nml-helpers, eval=FALSE}
# Retrieve the parsed configuration
cfg <- read_model_config(model = "simstrat_aed2",
                         lake_dir = get_lake_dir(aeme))

# Access the physical (Simstrat) configuration
cfg$hydrodynamic$ModelParameters$a_seiche   # seiche energy parameter
cfg$hydrodynamic$ModelConfig$InflowMode     # inflow placement mode

# Access the AED2 biogeochemistry configuration
aed2_nml <- read_nml(file.path(get_lake_dir(aeme), "simstrat_aed2", "aed2.nml"))
get_nml_value(aed2_nml, "models")           # active AED2 modules
get_nml_value(aed2_nml, "Rnitrif")          # nitrification rate
```

`check_simstrat_par()` validates a `simstrat.par` file — checking that all
required JSON sections and referenced input files are present, and that
`AED2ConfigFile`/`PathAED2initial`/`PathAED2inflow` exist when `CoupleAED2`
is enabled:

```{r check-par, eval=TRUE}
sim_dir <- file.path(get_lake_dir(aeme, path), "simstrat_aed2")
check_simstrat_par(file.path(sim_dir, "simstrat.par"))
```

---

## Installing the Simstrat-AED2 binary

Simstrat-AED2 binaries are distributed as GitHub release assets rather than
bundled inside the package. `install_simstrat_aed2()` downloads, checksum
verifies, and caches a pre-compiled binary for the current platform:

```{r install-simstrat, eval=FALSE}
list_simstrat_aed2_versions()
install_simstrat_aed2(version = "latest")
```

---

## Summary

This vignette demonstrated the key Simstrat-AED2-specific features
available in AEME:

1. **Model description** — Simstrat provides k-epsilon 1-D hydrodynamics
   with an internal-seiche mixing parameterisation and ice/snow model; AED2
   supplies the same modular biogeochemistry used by GLM-AED.

2. **Automatic module selection** — Simstrat-AED2's active AED2 modules are
   derived directly from `model_controls`/`set_vars_sim()`, rather than a
   separate model-toggling function.

3. **Model-specific water balance** — `build_aeme()` fits the lake water
   balance using Simstrat's own evaporation formula, distinct from the
   bulk-aerodynamic formula shared by GLM-AED and DYRESM-CAEDYM.

4. **Inflow modes** — manual (fixed-depth) vs. density-driven (plunging)
   inflow placement, configured via `ModelConfig.InflowMode`.

5. **Parameter access** — `simstrat_aed2_parameter_library` and
   `get_aeme_parameters()` make it easy to explore and modify individual
   model parameters, and `simstrat_aed2_parameters` provides ready-to-use
   calibration ranges (see the
   [aemetools](https://github.com/limnotrack/aemetools) package for
   automated calibration, or Simstrat's native PEST workflow for a
   standalone calibration outside AEME).
