---
title: "Sub-daily output and daily-mean storage"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
options(AEME.inform = FALSE)

# The worked example needs model binaries. Detect what is available so the
# article still renders on a machine with no models installed.
.ready <- function(f) tryCatch(nzchar(suppressMessages(f(quiet = TRUE))),
                               error = function(e) FALSE)
glm_ready      <- .ready(AEME::install_glm_aed)
simstrat_ready <- .ready(AEME::install_simstrat_aed2)
run_example    <- glm_ready && simstrat_ready
```

# Introduction

Every AEME run integrates the physics at a fixed internal step
(`time(aeme)$time_step`, 3600 s by default for GLM and GOTM, 300 s for
Simstrat) and *writes* output at a separate, usually coarser, cadence set by
`time(aeme)$output_time_step`. The historical default is **daily**
(`output_time_step = 86400`): one record per simulated day.

This article is about running with **sub-daily output** — hourly, 3-hourly,
6-hourly — and about the companion feature added alongside it: **daily-mean
storage**, where a sub-daily run also produces a small file holding the mean
of each day.

It is the sibling of `vignette("hourly-vs-daily-met")`, which is about the
*forcing*. The one rule that ties them together:

> **AEME never temporally disaggregates forcing.** Sub-daily *output* is only
> meaningful when the meteorological forcing is itself sub-daily. With daily
> meteo, a sub-daily run still resolves a weak diurnal cycle from the model's
> internal solar geometry, but most of the sub-daily structure you would want
> is simply not there to be written.

# Setting the output cadence

`set_output_time_step()` takes a frequency — a number of seconds, a named
value (`"hourly"`, `"daily"`, `"weekly"`), or an `"<n> <unit>"` string — and
each `build_*()` translates it into that model's native setting:

| model | native output control | for `output_time_step` `s` |
|---|---|---|
| GLM-AED | `&output` `nsave` | `round(s / dt)` integration steps per record |
| GOTM-WET | `output.yaml` `time_step` / `time_unit` | `s` seconds between records |
| Simstrat | `Output.Times` | `s / "Timestep s"` steps between records (a thinning interval) |

```{r, eval = FALSE}
aeme <- set_output_time_step(aeme, "3 hours")   # or 10800, or "6 hours", ...
time(aeme)$output_time_step
#> [1] 10800
```

`output_time_step` must be `>= time_step`, and on read-back the results are
matched positionally against an axis stepped by `output_time_step`, so the
value stored on the object must match the cadence the model was actually run
at. Change it, and you must rebuild and re-run.

# Daily-mean storage

`set_output_time_step(aeme, freq, daily_mean = TRUE)` (equivalently
`time(aeme)$output_daily_mean <- TRUE`) makes **every** model additionally
produce a daily-mean stream, so a run can be done at a sub-daily cadence — for
accuracy, or to study the diurnal cycle — while the series compared against
daily observations is a true daily *mean* rather than an instantaneous
snapshot at whatever time of day the output step happened to land on.

```{r, eval = FALSE}
aeme <- set_output_time_step(aeme, "hourly", daily_mean = TRUE)
```

How each model provides it:

* **GOTM-WET** writes daily means natively. Its shipped `output.yaml` already
  carries an `output_daily` block with `time_method: mean`; with the option on,
  that block is restricted to the targeted variables (the `model_controls`
  `simulate` set plus the internals the reader needs) instead of writing every
  variable.
* **GLM-AED** and **Simstrat** have no native time-averaging, so `run_aeme()`
  post-processes their sub-daily `output.nc` into a companion
  `output_daily.nc` (`write_output_daily_nc()`): the mean over each calendar
  day, one record per day, same file structure. The raw sub-daily file is kept
  — you get both.

The companion carries **only the variables the reader consumes**. For GLM,
whose own netCDF cannot be sub-selected, this is also where the stored output
is pruned: its large 4-D fields (`light`, `umean`, the wave spectra, …) are
dropped, leaving ~20 variables instead of ~60.

On read-back (`run_aeme()`, `get_var()`, …) the daily means are picked up
automatically: when `output_daily_mean` is `TRUE` the `output_daily.nc`
companion is read if present, otherwise the raw `output.nc` is averaged by
calendar day on read. Downstream code sees a plain daily `Date` axis, exactly
as for a native daily run.

```{r worked, eval = run_example, message = FALSE, warning = FALSE}
library(AEME)

aeme <- readRDS(system.file("extdata/aeme.rds", package = "AEME"))
aeme <- set_time(aeme, start = "2020-09-01", stop = "2021-09-01",
                 time_step = 3600)
aeme <- set_output_time_step(aeme, "hourly", daily_mean = TRUE)

path <- file.path(tempdir(), "subdaily-demo")
dir.create(path, showWarnings = FALSE)
aeme <- build_aeme(path = path, aeme = aeme,
                   model = c("glm_aed", "simstrat_aed2"),
                   model_controls = get_model_controls(),
                   ext_elev = 3, use_bgc = FALSE)
aeme <- run_aeme(aeme, model = c("glm_aed", "simstrat_aed2"), path = path)

ld <- AEME:::get_lake_dir(aeme, path)
for (m in c("glm_aed", "simstrat_aed2")) {
  od <- file.path(ld, m, "output")
  raw   <- file.size(file.path(od, "output.nc")) / 1e6
  daily <- file.size(file.path(od, "output_daily.nc")) / 1e6
  cat(sprintf("%-14s output.nc %6.1f MB   output_daily.nc %5.2f MB   (%.0fx)\n",
              m, raw, daily, raw / daily))
}

# get_var() returns the daily means, keyed by Date
tmp <- get_var(aeme, "HYD_temp", model = "glm_aed")
cat("HYD_temp Date class:", class(tmp$Date), " — days:",
    length(unique(tmp$Date)), "\n")
```

# What sub-daily output costs

The numbers below are from `data-raw/benchmark_output_cadence.R` (one lake,
one simulated year, `time_step` 3600 s, **daily** meteo, median of 2 reps on a
laptop). Absolute times are machine-specific; the *ratios* are the point.

**GLM-AED**

| cadence | model run | raw `output.nc` | daily companion | daily-averaging pass |
|---|--:|--:|--:|--:|
| 1 h  | 0.94 s | 279 MB | 1.04 MB | 1.3 s |
| 3 h  | 0.52 s |  93 MB | 1.03 MB | 0.6 s |
| 6 h  | 0.42 s |  47 MB | 1.02 MB | 0.4 s |
| 24 h | 0.34 s |  12 MB | —      | —     |

**Simstrat-AED2**

| cadence | model run | raw `output.nc` | daily companion | daily-averaging pass |
|---|--:|--:|--:|--:|
| 1 h  | 4.6 s | 32 MB  | 0.13 MB | 0.14 s |
| 3 h  | 2.1 s | 11 MB  | 0.13 MB | 0.08 s |
| 6 h  | 1.5 s | 5.3 MB | 0.12 MB | 0.06 s |
| 24 h | 1.1 s | 1.4 MB | —      | —      |

Reading these:

* **The integration is unchanged.** AEME does not touch GLM's `dt` or
  Simstrat's `"Timestep s"`. Every extra second above is output I/O, not
  physics — `nsave` / `Output.Times` only change how often a record is
  written.
* **Run cost scales roughly linearly with output frequency.** 1 h versus 24 h
  is about 2.8× the run time for GLM and 4.4× for Simstrat. Simstrat is the
  more sensitive because it writes one plain-text `.dat` row per variable per
  step; GLM writes netCDF. In absolute terms for a year it is +0.6 s (GLM) /
  +3.5 s (Simstrat) — small next to a BGC run, large next to a bare physics
  run.
* **Disk scales linearly too.** GLM's raw hourly output is large (the netCDF
  carries every 3-D field at every step); Simstrat's is modest.
* **The daily-mean companion is cadence-independent and tiny** (~1 MB/yr for
  GLM, ~0.13 MB/yr for Simstrat) because it is daily *and* restricted to the
  reader's variables. Keeping the companion and deleting the raw file is the
  real storage win — it is smaller than even the native 24 h file.
* **The daily-averaging pass is cheap** once GLM's companion is restricted:
  well under 2 s for a year.

`data-raw/benchmark_daily_mean_output.R` decomposes the feature cost further
(in-run I/O delta, the averaging pass, the read-back delta, and — for Simstrat
— the existing `.dat` → `output.nc` step for context).

# How accurate is the daily mean?

Same benchmark. `t1m_dmean_rmse_vs_1h` is the RMSE of the daily-mean 1 m water
temperature at each cadence against the 1 h run's daily mean;
`t_surf_daily_range` is the mean daily (max − min) of the surface temperature
cell — the diurnal signal the cadence resolves.

| model | cadence | daily-mean RMSE vs 1 h | surface diurnal range |
|---|---|--:|--:|
| GLM-AED | 3 h | 0.029 °C | 1.34 °C |
| GLM-AED | 6 h | 0.040 °C | 1.11 °C |
| GLM-AED | 1 h | — (ref) | 1.41 °C |
| Simstrat | 3 h | 0.006 °C | 0.18 °C |
| Simstrat | 6 h | 0.014 °C | 0.16 °C |
| Simstrat | 1 h | — (ref) | 0.20 °C |
| Simstrat | 24 h *(instantaneous)* | 0.062 °C | 0 |

* **3–6 h output is enough for an unbiased daily mean.** Sampling the mean
  from 3-hourly instead of hourly records adds only ~0.03 °C (GLM) / ~0.006 °C
  (Simstrat). Taking the *daily instantaneous* value from a 24 h run instead
  is worse than a 6 h mean (Simstrat: 0.062 vs 0.014 °C) — a midnight snapshot
  is a biased estimator of the daily mean, which is the whole reason the
  `daily_mean` option exists.
* **1 h is worth it when the diurnal cycle itself is the subject.** 6 h output
  resolves ~78 % of GLM's surface diurnal range, 3 h ~95 %.
* These ranges are small because the fixture forcing is daily — the diurnal
  signal is only the model's internal solar cycle. **With sub-daily meteo the
  ranges (and the case for 1 h output) grow substantially.**

## Rule of thumb

| you want … | run at | store |
|---|---|---|
| daily statistics, unbiased | 3 h (6 h is fine) | `daily_mean = TRUE` |
| the diurnal cycle | 1 h | sub-daily, `daily_mean` optional |
| just faster archiving of an existing sub-daily run | — | `daily_mean = TRUE`, then keep only `output_daily.nc` |

# Gotchas

* **Sub-daily output needs sub-daily meteo.** `check_time()` stops you from
  pairing daily meteo with a sub-daily `output_time_step`. Inflows and
  outflows may stay daily.
* **`daily_mean = TRUE` with a daily `frequency` warns** — there is nothing
  sub-daily to average.
* **Rebuild after changing the cadence.** `output_time_step` /
  `output_daily_mean` are stored on the object and drive both the model
  setting and the read-back axis; a stale value against a re-run model is
  caught with a warning and dropped records, not a silent wrong answer.
* **Both files are kept.** The daily companion sits next to the raw
  `output.nc` / `*_out.dat`. To realise the storage saving, delete the raw
  file yourself once the companion is written.
* **GLM's companion is a subset.** It holds the reader's variables, not every
  field GLM wrote. If you need a dropped 3-D field (`umean`, wave spectra, …)
  at daily resolution, read and average the raw `output.nc` yourself.
* **Water balance is always daily**, sub-daily run or not — deliberate.

# See also

* `vignette("hourly-vs-daily-met")` — sub-daily *forcing*, and how each model
  fills the gaps between timestamps.
* `?set_output_time_step` — the `frequency` and `daily_mean` arguments.
* `?set_time` — `time_step` vs `output_time_step`.
* `data-raw/benchmark_output_cadence.R`,
  `data-raw/benchmark_daily_mean_output.R` — the scripts behind the tables
  above.
