---
title: "Testing Parameters Without a Full Rebuild"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

<!--
DRAFT -- sketched from the vignette-coverage audit, gap 02 (per-model config
editing, 0/22 functions demonstrated anywhere). Every code chunk below has
been run end-to-end (not just knit-checked): get_glm_param/set_glm_param
round-trip confirmed, run_model_test()'s param_overrides confirmed to reach
the model (Kw ends up 0.5 in the nml after the call), and the returned list's
$HYD_temp confirmed as a [depth x time] matrix. Referenced from
glm-aed.Rmd's "Working with the GLM configuration directly" section --
`vignette("testing-parameters")` there resolves to this file's name.
TODO before publishing:
  - the aemetools package (Suggests) offers the same pattern at a higher
    level for full sensitivity analysis / calibration (run_aeme_param(),
    sa_aeme(), calib_aeme()) -- add a "when to reach for aemetools instead"
    closing note once that boundary is settled
  - tgt_vars is demonstrated but undersold below: confirm whether it's
    meant to *restrict* the reader's output and, if so, why HYD_temp isn't
    alone in names(out) in the current build (every LKE_*/diagnostic field
    from read_glm_output()'s defaults comes back regardless) -- reword the
    "Reading back only what you need" section once that's confirmed against
    read_glm_output.R rather than inferred from one run
-->

# Introduction

`build_aeme()` -> `run_aeme()` is the right pipeline for a real ensemble run,
but re-running the whole `Aeme` object just to check what one parameter does
is slow, and doesn't matter for a five-minute "does this move the answer in
the direction I expect" check. Two lower-level tools skip the `Aeme` object
entirely and work straight on a model's configuration directory:

- `get_glm_param()`/`set_glm_param()` (and the GOTM-WET/DYRESM-CAEDYM/
  Simstrat-AED2 equivalents) -- read or write one parameter in the raw
  config file.
- `run_model_test()` -- apply a batch of changes (parameters, initial
  conditions, inflows, outflows) and run, all from a path alone.

Both need an existing build to work on -- they edit files `build_aeme()`
already wrote, they don't create one.

```{r setup, message = FALSE, warning = FALSE}
library(AEME)

# Plotting libraries for the scenario comparison at the end
library(ggplot2)
library(patchwork)

```

```{r build, message = FALSE, results = "hide"}
tmpdir <- tempdir()
aeme_dir <- system.file("extdata/lake/", package = "AEME")
file.copy(aeme_dir, tmpdir, recursive = TRUE)
path <- file.path(tmpdir, "lake")

aeme <- yaml_to_aeme(path = path, "aeme.yaml")
model_controls <- get_model_controls(use_bgc = FALSE)
model <- "glm_aed"

aeme <- build_aeme(path = path, aeme = aeme, model = model,
                   model_controls = model_controls, ext_elev = 5,
                   use_bgc = FALSE)
path_glm <- file.path(get_lake_dir(aeme, path), "glm_aed")
```

# Reading and setting one parameter

`path_glm` now points at a real, built GLM-AED configuration directory.
`get_glm_param()` reads a value straight out of the nml, wherever it's
nested:

```{r get-param}
get_glm_param(path_glm, "Kw")
```

`set_glm_param()` writes one back -- multiple `name = value` pairs in one
call, no need to know which `&block` each belongs to:

```{r set-param}
set_glm_param(path_glm, Kw = 0.8, coef_mix_hyp = 0.3)
get_glm_param(path_glm, "Kw")
```

This is a permanent edit to the file on disk (there's no `Aeme` object
tracking it), so it's ideal for a quick before/after comparison but not for
anything you want to reproduce later -- keep the parameter table AEME's
calibration tools use (see `vignette("aeme-inputs")`) as the record for
anything that matters.

The AED biogeochemistry file takes the same call, passed explicitly via
`glm_file`:

```{r set-param-aed, eval = FALSE}
set_glm_param(path_glm, glm_file = file.path(path_glm, "aed", "aed.nml"),
             theta_sed_oxy = 1.08)
```

GOTM-WET, DYRESM-CAEDYM and Simstrat-AED2 configuration directories work
the same way, with `get_gotm_param()`/`set_gotm_param()`,
`get_dy_cd_param()`/`set_dy_cd_param()`, and
`get_simstrat_param()`/`set_simstrat_param()`.

# Running a what-if scenario

`run_model_test()` goes one step further: apply a batch of overrides and run
the model, in one call, without touching the `Aeme` object at all.

```{r run-test}
out <- run_model_test("glm_aed", path_glm,
                      param_overrides = list(Kw = 0.5),
                      tgt_vars = "HYD_temp")
```

The parameter change is written to disk exactly as `set_glm_param()` would:

```{r run-test-check}
get_glm_param(path_glm, "Kw")
```

`param_overrides` is forwarded to `set_glm_param()`, so any parameter name
that works there works here. `init`, `inflow_args` and `outflow_args`
reach the model's `set_*_init()`/`set_*_inflows()`/`set_*_outflows()`
wrappers the same way -- e.g. to test a different initial temperature
profile without re-running `build_aeme()`:

```{r run-test-init, eval = FALSE}
run_model_test("glm_aed", path_glm,
              init = list(temp = seq(20, 10, length.out = 10)),
              tgt_vars = "HYD_temp")
```

## Reading back the result

`out` is the model's own reader output (`read_glm_output()`, for
`model = "glm_aed"`) -- `tgt_vars` names what you're interested in, but the
reader's own diagnostic fields (lake level, heat fluxes, ice, meteorology
echoed back, ...) still come along with it:

```{r run-test-result}
dim(out$HYD_temp)
out$HYD_temp[1:3, 1:3]
```

`out$HYD_temp` is a `[depth x time]` matrix, matching what `get_var()`
returns from a full `Aeme` object -- so the same downstream code (plotting,
summary statistics) works whichever path produced it.

## Comparing scenarios

The pattern for comparing a handful of candidate values: loop, collecting
one `tgt_vars` matrix per scenario.

```{r compare-scenarios}
kw_values <- c(0.3, 0.5, 0.8, 1.2)
results <- lapply(kw_values, function(kw) {
  run_model_test("glm_aed", path_glm, param_overrides = list(Kw = kw),
                 tgt_vars = "HYD_temp")
})
names(results) <- paste0("Kw_", kw_values)
```

```{r fig-temp-compare}
plist <- lapply(seq_along(results), \(i) {
  plot_model_output(results[[i]], var_sim = "HYD_temp", var_lims = c(10, 29)) +
    ggtitle(paste0("Kw = ", kw_values[i]))
})

wrap_plots(plist, ncol = 2, guides = "collect") & 
  theme(legend.position = "bottom")
```


For anything past a handful of candidates -- a real sensitivity sweep or a
calibration -- reach for aemetools instead
(`aemetools::run_aeme_param()`/`sa_aeme()`/`calib_aeme()`), which builds on
the same idea but works from the full parameter table and the `Aeme`
object, with proper sampling designs and PEST++ integration.

## When something goes wrong

By default (`safe = TRUE`) a failed edit or model run is caught and reported
with a message rather than stopping -- useful when looping over many
candidate scenarios and one bad combination shouldn't abort the rest:

```{r run-test-safe, eval = FALSE}
run_model_test("glm_aed", path_glm, param_overrides = list(lake_depth = -1))
# message: run_model_test error: ... ; returns NULL rather than erroring
```

Set `safe = FALSE` while developing a scenario, so a mistake surfaces
immediately instead of silently returning `NULL`.

# Next steps

`vignette("visualising-output")` covers turning `out$HYD_temp` (or the
`Aeme` object itself) into the plots this kind of comparison is usually
for.
