---
title: "Visualising AEME Output"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

<!--
DRAFT -- sketched from the vignette-coverage audit, gap 01 (plot_* family,
0/20 functions demonstrated in any vignette; every other vignette uses the
S4 plot(aeme, ...) dispatcher or hand-rolled ggplot instead). Every eval=TRUE
chunk has been run end-to-end against this vignette's own build() (not just
knit-checked) -- plot_ts, plot_var, plot_obs, plot_resid, plot_nit,
plot_met_tile, plot_hyps, plot_flows and plot_model_output all confirmed
working with these exact arguments. TODO before publishing:
  - swap the toy example for one of the richer example objects
    (aeme.rds / LID45819_wainamu) so the biogeochemical plots have more than
    one timestep of variation to show, then flip the plot_phs/plot_car/
    plot_phytos/plot_zoops/plot_fluxes chunks from eval=FALSE to TRUE and
    confirm each one
  - review plot quality/captions once real screenshots exist -- correctness
    was checked, visual polish wasn't
  - decide whether plot_glm_output()/plot_output_base() belong here or in
    glm-aed.Rmd next to the raw-config-editing section (they operate on a
    raw output list, not an Aeme object, unlike everything else on this
    page)
-->

# Introduction

`plot(aeme, "output")` and friends -- the S4 dispatcher used throughout the
other vignettes -- cover the common case. Underneath it, AEME ships a wider
set of purpose-built plotting functions: multi-variable timeseries, group
plots for individual biogeochemical modules, residuals against observations,
and diagnostics on the inputs (hypsograph, meteorology, flows) rather than
the output. This vignette works through them using one small built model.

```{r setup, message = FALSE, warning = FALSE}
library(AEME)
library(ggplot2)
options(AEME.inform = FALSE)
```

```{r build, message = FALSE, results = "hide"}
aeme_file <- system.file("extdata/aeme.rds", package = "AEME")
aeme <- readRDS(aeme_file)

model_controls <- get_model_controls(use_bgc = TRUE)
model <- "glm_aed"
path <- tempdir()

aeme <- build_aeme(path = path, aeme = aeme, model = model,
                   model_controls = model_controls, ext_elev = 5,
                   use_bgc = TRUE)
aeme <- run_aeme(aeme = aeme, verbose = FALSE)
```

# Timeseries

`plot_ts()` is the general-purpose multi-variable plot -- pass one or more
`var_sim`, and it facets automatically. Observations overlay by default
(`add_obs = TRUE`) where the `Aeme` object has them.

```{r plot-ts}
plot_ts(aeme, model = model, var_sim = c("HYD_temp", "CHM_oxy"))
```

Restrict to a depth range (e.g. surface only) with `depth_range`:

```{r plot-ts-depth}
plot_ts(aeme, model = model, var_sim = "HYD_temp", depth_range = c(0, 2))
```

`plot_var()` is the lower-level building block `plot_ts()` is written on top
of -- reach for it directly when you already have a `data.frame` from
`get_var()` and want the same plot style without re-deriving it from the
`Aeme` object:

```{r plot-var}
df <- get_var(aeme, model = model, var_sim = "HYD_temp", return_df = TRUE)
plot_var(df = df, aeme = aeme, model = model, var_sim = "HYD_temp")
```

# Observations and residuals

`plot_obs()` shows the observation record alone -- useful for checking
coverage before deciding what to calibrate against (see also
`summarise_obs()` in `vignette("setup-new-lake")`):

```{r plot-obs}
plot_obs(aeme, var_sim = "HYD_temp")
```

`plot_resid()` plots modelled-minus-observed directly, rather than the two
series overlaid -- the faster way to see whether a model is biased high/low
versus just imprecise:

```{r plot-resid}
plot_resid(aeme, model = model, var_sim = "HYD_temp")
```

# Biogeochemical group plots

Each major AED module has its own plot, all sharing the `plot_ts()`
argument shape (`add_obs`, `depth_range`, `remove_spin_up`, `ens_n`):

```{r plot-nit}
plot_nit(aeme)
```

```{r plot-phs}
plot_phs(aeme)      # phosphorus
```

```{r plot-car}
plot_car(aeme)      # carbon
```

```{r plot-phytos}
plot_phytos(aeme)   # phytoplankton groups
```
Similarly, `plot_zoops()` shows the zooplankton groups but requires the 
zooplankton module to be active in the model configuration. This example's 
minimal config does not carry zooplankton/full phyto groups.

```{r plot-zoops eval = FALSE}
plot_zoops(aeme)    # zooplankton groups
```

# Inputs, not output

Three functions plot what went *into* the model rather than what came out --
useful when a run looks wrong and the question is whether the forcing data
is the problem.

Meteorology, as a tile plot (one row per variable, coloured by value,
faceted by hydrological year):

```{r plot-met-tile}
plot_met_tile(aeme, var_inp = "MET_tmpair")
```

The hypsograph the model was built with:

```{r plot-hyps}
plot_hyps(aeme)
```

Inflows and/or outflows:

```{r plot-flows}
plot_flows(aeme, flow = "inflow")
```

Cumulative heat fluxes (e.g. checking an energy budget):

```{r plot-fluxes}
plot_fluxes(aeme, cumulative = TRUE)
```

# Working with a raw output list

`plot_model_output()` is the lowest-level of the plotting functions: pass it
an `Aeme` object directly, as below, or -- its other accepted input -- the
specifically classed list `read_glm_output()`/`read_gotm_output()`/
`read_simstrat_output()`/`read_dy_output()`/`read_model_outputs()` return
(`output(aeme)` itself is a plain list and isn't accepted; use one of those
readers if you need output outside the `Aeme` object). It's what every
other `plot_*` function on this page is ultimately built on:

```{r plot-model-output}
outfile <- get_model_outfile(aeme, model = "glm_aed")$glm_aed
outfile
```

`read_glm_output()` reads the model output, projects onto a standard grid
and also does some unit conversions, especially for the AED variables that are 
stored in mmol/m3 in the raw output but are more commonly expressed in mg/L.

```{r plot-model-output}
out <- read_glm_output(file = outfile)
plot_model_output(out, var_sim = "CHM_oxy")
```

However, you can also load all raw output from the GLM netcdf output file and 
plot any variable in its raw form e.g. model units and Lagrangian grid, rather 
than the standardised grid and converted units. This is useful for plotting 
variables that are not included in the `Aeme` object by default.
You need to use the raw variable name for plotting instead of the standardised
AEME variable name.

```{r}
raw <- read_glm_output(file = outfile, raw = TRUE, load_all = TRUE)
raw
```

```{r}
plot_model_output(raw, var_sim = "OXY_oxy")
```



The AED sediment fluxes, which are not included in the `Aeme` object by default.
For example, to plot the sediment oxygen flux:

```{r}
plot_model_output(raw, var_sim = "SDF_Fsed_oxy_Z")
```

# Next steps

`vignette("testing-parameters")` covers changing a parameter and re-running
without a full rebuild -- these plotting functions are the natural way to
compare the result against what you started with.
